
# Death

```{r}
#| label: prep
#| output: false

# source general functions
source(paste0(here::here(), "/1_code/0000_general/general_utils.R"))

# read general variables for common use
params <- yaml::read_yaml(paste0(here::here(), "/Plots/_variables.yml"))

# source utility functions
run_scen(params$scen, "dea")

```

## Mortality

### age plots

#### focus: difference in mortality by sex

```{r}
#| label: fig-mor_yasr_age_sex
#| fig-cap: "mortality by age, year, sex, region"

# years (past only)
year_past <- (date_start + 1):date_end
year_5 <- sort(unique(year_past[year_past %% 5 == 0]))

sszplot(filter(mor_yasr, year %in% year_5),
  aes_x = "age", aes_y = "mor_yasr", aes_col = "sex",
  labs_y = "mortality rate (in % per year)",
  scale_y = "log",
  grid = c("region", "year"),
  name = plot_name("0200_mortality_by-year-age-sex-region_focus-age-sex"),
  width = 12, height = 7
)
```

#### focus: difference in mortality by region

```{r}
#| label: fig-mor_yasr_age_region
#| fig-cap: "mortality by age, year, region, sex"

sszplot(filter(mor_yasr, year %in% year_5),
  aes_x = "age", aes_y = "mor_yasr", aes_col = "region",
  labs_y = "mortality rate (in % per year)",
  scale_y = "log",
  grid = c("sex", "year"),
  name = plot_name("0201_mortality_by-year-age-sex-region_focus-age-region"),
  width = 12, height = 7
)
```

### year plots

#### focus: difference in mortality by sex

```{r}
#| label: fig-mor_yasr_year_rsex
#| fig-cap: "mortality by year, age, sex, region"

# age
age_select <- c(0, 20, 40, 50, 60, 70, 80)

sszplot(filter(mor_yasr, age %in% age_select),
  aes_x = "year", aes_y = "mor_yasr", aes_col = "sex",
  labs_y = "mortality rate (in % per year)",
  scale_y = "log",
  grid = c("region", "age"),
  name = plot_name("0202_mortality_by-year-age-sex-region_focus-year-sex"),
  width = 14, height = 7,
  angle = 90
)
```

#### focus: difference in mortality by region

```{r}
#| label: fig-mor_yasr_year_region
#| fig-cap: "mortality by year, age, region, sex"

sszplot(filter(mor_yasr, age %in% age_select),
  aes_x = "year", aes_y = "mor_yasr", aes_col = "region",
  labs_y = "mortality rate (in % per year)",
  scale_y = "log",
  grid = c("sex", "age"),
  name = plot_name("0203_mortality_by-year-age-sex-region_focus-year-region"),
  width = 14, height = 7,
  angle = 90
)
```

### method comparison (based on rate vs. death/birth/population)

method differences (rate vs. deaths/births/population) details: see life expectancy function definition

```{r}
#| label: fig-le_ysr
#| fig-cap: "life expectancy at birth by year and method"

# plot data
le_methods <- filter(le_ysr, region == "Zurich") %>%
  right_join(le, by = c("year", "sex")) %>%
  rename(basic = life, rate = le_ysr) %>%
  pivot_longer(c(rate, basic), names_to = "method", values_to = "life")

sszplot(le_methods,
  aes_x = "year", aes_y = "life", aes_col = "method",
  labs_y = "life expectancy at birth",
  grid = c(".", "sex"),
  name = plot_name("0204_life-expectancy-at-birth_by-year-method"),
  width = 9, height = 5
)
```

### life expectancy

#### focus: differences by sex

```{r}
#| label: fig-le_ysr_sex
#| fig-cap: "life expecancy at birth by year, sex and region"

# years (all: past and future)
year_all <- (date_start + 1):scen_end
year_all_5 <- sort(unique(year_all[year_all %% 5 == 0]))

sszplot(le_ysr,
  aes_x = "year", aes_y = "le_ysr", aes_col = "sex",
  i_x = year_all_5,
  labs_y = "life expectancy at birth",
  grid = c(".", "region"),
  name = plot_name("0205_life-expectancy-at-birth_by-year-sex-region_focus-sex"),
  width = 10, height = 6
)
```

#### focus: differences by region

```{r}
#| label: fig-le_ysr_region
#| fig-cap: "life expecancy at birth by year, region and sex"

sszplot(le_ysr,
  aes_x = "year", aes_y = "le_ysr", aes_col = "region",
  i_x = year_all_5,
  labs_y = "life expectancy at birth",
  grid = c(".", "sex"),
  name = plot_name("0206_life-expectancy-at-birth_by-year-sex-region_focus-region"),
  width = 10, height = 6
)
```

### mortality over base years (including tail correction)

```{r}
#| label: fig-mor_asr
#| fig-cap: "mortality by age, sex, region; tails corrected"

# plot
sszplot(mor_asr,
  aes_x = "age", aes_y = "mor_asr", aes_col = "region",
  labs_y = "mortality rate (in % per year)",
  scale_y = "log",
  grid = c("sex", "."),
  name = plot_name("0207_mortality_by-age-sex-region_tails-corrected"),
  width = 8, height = 7
)
```

### smoothing with LOESS

```{r}
#| label: fig-plot_dat_fit
#| fig-cap: "mortality bby age, sex and region; fitted"

# fit: levels
fit_lev <- c("initial", "smoothed")

# plot data
plot_dat_fit <- mor_fit %>%
  pivot_longer(c(mor_asr, mor_fit), names_to = "category", values_to = "mor") %>%
  mutate(cat = factor(if_else(category == "mor_asr",
    fit_lev[1], fit_lev[2]
  ), levels = fit_lev))

# plot
sszplot(plot_dat_fit,
  aes_x = "age", aes_y = "mor", aes_col = "region", aes_ltyp = "cat",
  labs_y = "mortality rate (in % per year)",
  scale_y = "log",
  grid = c("sex", "."),
  name = plot_name("0208_mortality_by-age-sex-region_fitted"),
  width = 8, height = 7
)
```

### ratio Zurich / Switzerland

```{r}
#| label: fig-ratio_as
#| fig-cap: "ratio Zurich/Switzerland by age and sex"

# plot
sszplot(ratio_as,
  aes_x = "age", aes_y = "ratio", aes_col = "sex",
  i_x = c(dea_lower, dea_upper), i_y = 1,
  labs_y = "mortality ratio (Zurich / Switzerland)",
  scale_y = c(0, NA),
  name = plot_name("0209_ratio-Zurich-Switzerland_by-age-sex"),
  width = 10, height = 5
)
```

### Zurich: future mortality rate

```{r}
#| label: fig-mor_zh_yas_past_future
#| fig-cap: mortality prediction by year, age and sex

# plot
sszplot(mor_zh_yas_past_future,
  aes_x = "age", aes_y = "mor_yas", aes_col = "year",
  labs_y = "mortality rate (in % per year)",
  scale_y = "log",
  grid = c("sex", "."),
  name = plot_name("0210_mortality-prediction_by-year-age-sex"),
  width = 12, height = 8
)
```

### Zurich: life expectancy (including the model data)

#### focus: differences by sex

```{r}
#| label: le_ysr_model_sex
#| fig-cap: "life expectancy at birth by year, sex, region; with model"

# why? visual check, if life expectancy based on predicted rates are ok

# ZH: life expectancy
le_ys_ZH <- life_exp(
  data = mor_zh_yas_past_future, mor = "mor_yas",
  age = "age", group_cols = c("year", "sex"),
  age_max = dea_age_max_le, qx_NA = dea_qx_NA_le,
  age_at = dea_age_at, radix = dea_radix
) %>%
  mutate(region = factor(text_r[1], uni_r)) %>%
  rename(le_ysr = life_exp) %>%
  select(year, sex, region, le_ysr)

# plot data
le_ysr_model <- filter(le_ysr, region != "Zurich") %>%
  bind_rows(le_ys_ZH) %>%
  arrange(year, sex, region)


sszplot(le_ysr_model,
  aes_x = "year", aes_y = "le_ysr", aes_col = "sex",
  labs_y = "life expectancy at birth",
  grid = c(".", "region"),
  name = plot_name("0211_life-expectancy-at-birth_by-year-sex-region_focus-sex_model"),
  width = 10, height = 6
)

```

#### focus: differences by region

```{r}
#| label: le_ysr_model_region
#| fig-cap: "life expectancy at birth by year, region, sex; with model"

sszplot(le_ysr_model,
  aes_x = "year", aes_y = "le_ysr", aes_col = "region",
  labs_y = "life expectancy at birth",
  grid = c(".", "sex"),
  name = plot_name("0212_life-expectancy-at-birth_by-year-sex-region_focus-region_model"),
  width = 10, height = 6
)
```
