
# Immigration

```{r}
#| label: prep
#| output: false

# source general functions
source(paste0(here::here(), "/1_code/0000_general/general_utils.R"))

# read general variables for common use
params <- yaml::read_yaml(paste0(here::here(), "/Plots/_variables.yml"))

# source utility functions
run_scen(i_scen, "ims")

# assign the return values of the migration functions so they can be used inside the quarto environment
rate <- ims_rate_dy
prop <- ims_prop_so_dy
pred <- ims_prop_a_dyso

```

## immigration\* rate (per district and year)

```{r}
#| label: fig-0300
#| fig-cap: immigration* rate by district and year
#| fig-height: 10

  # years of the past (for plot)
  year_past <- (date_start + 1):date_end
  year_past_5 <- sort(unique(year_past[year_past %% 5 == 0]))

  # plot
  sszplot(rate$mis_rate_dy,
    aes_x = "year", aes_y = "mis_rate_dy",
    i_x = year_past_5,
    wrap = "district", ncol = 4,
    labs_y = "immigration* rate (in % per year)",
    name = plot_name("0300_immigration-star-rate_by-district-year"),
    width = 12, height = 14
  )

```

## prediction: future immigration\* rate (per year and district)

### overview

```{r}
#| label: fig-0301
#| fig-cap: immigration* rate by district and year; prediction
#| fig-height: 10

  # levels
  time_lev <- c("past", "future")

  # plot data
  plot_dat_mis_pred <- select(rate$mis_past_pred, district, year, rate_all) %>%
    mutate(time = factor(if_else(year <= rate$mis_base_end,
      time_lev[1], time_lev[2]
    ), levels = time_lev))

  sszplot(plot_dat_mis_pred,
    aes_x = "year", aes_y = "rate_all", aes_ltyp = "time",
    i_x = c(rate$mis_base_begin, rate$mis_base_end),
    wrap = "district", ncol = 4,
    labs_y = "immigration* rate (in % per year)",
    name = plot_name("0301_immigration-star-rate_by-district-year_prediction"),
    width = 12, height = 14
  )
```

### more detailed: regression and limits

```{r}
#| label: fig-0302
#| fig-cap: immigration* rate by district and year; prediction with details
#| fig-height: 10

  sszplot(rate$mis_past_pred,
    aes_x = "year", aes_y = "mis_rate_dy",
    geom = "point",
    i_x = c(rate$mis_base_begin, rate$mis_base_end),
    wrap = "district", ncol = 4,
    labs_y = "immigration* rate (in % per year)",
    name = plot_name("0302_immigration-star-rate_by-district-year_prediction-details"),
    width = 12, height = 14,
    quotes = c(
      quote(geom_line(aes(x = year, y = pred), linetype = 2)),
      quote(geom_line(aes(x = year, y = pred_mean), linetype = 3)),
      quote(geom_line(aes(x = year, y = pred_roll)))
    )
  )
```

## migration: distribution of sex and origin

```{r}
#| label: fig-0303
#| fig-cap: immigration proportion of sex by origin, district and year
#| fig-height: 10

  # years of the past (for plot)
  year_past <- (date_start + 1):date_end
  year_past_5 <- sort(unique(year_past[year_past %% 5 == 0]))

  # plot
  sszplot(prop$mis_dyso,
    aes_x = "year", aes_y = "mis_prop_dyso", aes_col = "sex", aes_ltyp = "origin",
    i_x = year_past_5,
    wrap = "district", ncol = 4,
    labs_y = "proportion in % (per district and year, in immigration*)",
    name = plot_name("0303_immigration_proportion-sex-origin_by-district-year"),
    width = 12, height = 14
  )
  
```

## distribution of sex and origin: prediction

```{r}
#| label: fig-0304
#| fig-cap: immigration proportion of sex by origin, district and year; prediction
#| fig-height: 10

  sszplot(prop$mis_so_past_pred,
    aes_x = "year", aes_y = "prop_all", aes_col = "sex", aes_ltyp = "origin",
    i_x = c(prop$mis_so_base_begin, prop$mis_so_base_end),
    wrap = "district", ncol = 4,
    labs_y = "proportion in % (per district and year, in immigration*)",
    name = plot_name("0304_immigration_proportion-sex-origin_by-district-year_prediction"),
    width = 12, height = 14
  )
```

## age proportion

### immigration\*

```{r}
#| label: fig-0310
#| fig-cap: immigration* per district, year, sex, origin
#| fig-height: 10

  # years of the past (for plot)
  year_past <- date_start:date_end
  year_past_5 <- sort(unique(year_past[year_past %% 5 == 0]))

  # plot
  sszplot(pred$mis_dyso,
    aes_x = "year", aes_y = "mis_dyso", aes_col = "sex", aes_ltyp = "origin",
    i_x = year_past_5,
    wrap = "district", ncol = 4,
    labs_y = "immigration* per year",
    name = plot_name("0310_immigration_star_per-district-year-sex-origin"),
    width = 12, height = 14
  )
```

```{r}
#| label: fig-0311
#| fig-cap: immigration* per district, year, sex, origin; free scales

  sszplot(pred$mis_dyso,
    aes_x = "year", aes_y = "mis_dyso", aes_col = "sex", aes_ltyp = "origin",
    i_x = year_past_5,
    wrap = "district", ncol = 4, gridscale = "free",
    labs_y = "immigration* per year",
    name = plot_name("0311_immigration_star_per-district-year-sex-origin_free-scales"),
    width = 12, height = 14
  )
```

### age proportion: per district, year, sex, origin

#### focus age

```{r}
#| label: fig-0312
#| fig-cap: immigration* age proportion per district, year, sex, origin; focus on age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  # years (subjectively selected)
  # WHY with rev? To have the last year in the plot
  years_plot <- rev(seq(date_end, date_start, by = -8))

  sszplot(filter(pred$mis_dyaso, year %in% years_plot),
    aes_x = "age", aes_y = "mis_prop_a", aes_col = "year",
    grid = c("sex", "origin"),
    labs_y = "proportion in %", labs_col = "year",
    name = plot_name("0312_immigration_star_age-proportion_per-district-year-sex-origin_focus-age"),
    width = 11, height = 8,
    multi = uni_d
  )

```

#### focus years

```{r}
#| label: fig-0313
#| fig-cap: immigration* age proportion per district, year, sex, origin; focus on years
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  # age (subjectively selected)
  age_plot <- seq(0, 60, by = 20)

  sszplot(filter(pred$mis_dyaso, age %in% age_plot),
    aes_x = "year", aes_y = "mis_prop_a", aes_col = "age",
    grid = c("sex", "origin"),
    labs_y = "proportion in %", labs_col = "age",
    name = plot_name("0313_immigration_star_age-proportion_per-district-year-sex-origin_focus-years"),
    width = 11, height = 8,
    multi = uni_d
  )
```

## smoothing immigration\* with LOESS over years (by district, age, sex)

#### focus age

```{r}
#| label: fig-0314
#| fig-cap: immigration* smoothed over year; focus on age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  # plot preparation
  fit_lev <- c("initial", "smoothed")  
  mis_smooth_plot <- pred$mis_smooth %>% 
    pivot_longer(c(mis_dyaso, mis_smooth), names_to = "category", values_to = "mis") %>% 
    mutate(cat = factor(if_else(category == "mis_dyaso",
      fit_lev[1], fit_lev[2]
    ), levels = fit_lev)) %>%
    select(district, year, age, sex, origin, cat, mis)

  # plot: focus age distribution
  sszplot(filter(mis_smooth_plot, year %in% year_past_5),
    aes_x = "age", aes_y = "mis", aes_col = "cat",
    grid = c("as.factor(year)", "origin*sex"),
    labs_y = "immigration* per year",
    name = plot_name("0314_immigration_star_smoothed-over-year_focus-age"),
    width = 11, height = 8,
    multi = uni_d
  )
```

#### focus years

```{r}
#| label: fig-0315
#| fig-cap: immigration* smoothed over year; focus on years
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  # age (subjectively selected)
  age_plot_smooth <- seq(0, 60, by = 20)

  sszplot(filter(mis_smooth_plot, age %in% age_plot_smooth),
    aes_x = "year", aes_y = "mis", aes_col = "cat",
    grid = c("as.factor(age)", "origin*sex"),
    labs_y = "immigration* per year",
    name = plot_name("0315_immigration_star_smoothed-over-year_focus-years"),
    width = 11, height = 8,
    multi = uni_d
  )

```

## age proportion (after smoothing immigration\* over years)

#### focus age

```{r}
#| label: fig-0316
#| fig-cap: immigration* age proportion after smoothing; focus on age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  sszplot(filter(pred$mis_age_prop_smooth, year %in% year_past_5),
    aes_x = "age", aes_y = "prop_a_smooth", aes_col = "year",
    grid = c("sex", "origin"),
    labs_y = "proportion in %",
    name = plot_name("0316_immigration_star_age-proportion_after-smoothing_focus-age"),
    width = 11, height = 8,
    multi = uni_d
  )
```

#### focus years

```{r}
#| label: fig-0317
#| fig-cap: immigration* age proportion after smoothing; focus on years
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  # age (subjectively selected)
  age_plot_smooth_prop <- seq(0, 60, by = 20)

  sszplot(filter(pred$mis_age_prop_smooth, age %in% age_plot_smooth_prop),
    aes_x = "year", aes_y = "prop_a_smooth", aes_col = "age",
    grid = c("sex", "origin"),
    labs_y = "proportion in %", labs_col = "age",
    name = plot_name("0317_immigration_star_age-proportion_after-smoothing_focus-years"),
    width = 11, height = 8,
    multi = uni_d
  )
```

### smoothing proportion by age with LOESS

#### focus age

```{r}
#| label: fig-0318
#| fig-cap: immigration* proportion smoothed; focus on age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7
  
  # plot preparation
  mis_fit_plot <- pred$prop_fit %>% 
    pivot_longer(c(prop_a_smooth, prop_fit), names_to = "category", values_to = "prop") %>% 
    mutate(cat = factor(if_else(category == "prop_a_smooth",
      fit_lev[1], fit_lev[2]
    ), levels = fit_lev)) %>%
    select(district, year, age, sex, origin, cat, prop)    
    

  sszplot(filter(mis_fit_plot, year %in% year_past_5),
    aes_x = "age", aes_y = "prop", aes_col = "cat",
    grid = c("as.factor(year)", "origin*sex"),
    labs_y = "proportion in %",
    name = plot_name("0318_immigration_star_proportion_smoothed_focus-age"),
    width = 11, height = 8,
    multi = uni_d
  )
```

## regression

```{r}
#| label: fig-0319
#| fig-cap: immigration* age proportion by district, year, sex, origin; past and future
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  sszplot(pred$mis_a_past_pred,
    aes_x = "age", aes_y = "prop_a", aes_col = "year",
    grid = c("origin", "sex"),
    labs_y = "proportion in %",
    name = plot_name("0319_immigration_star_age-proportion_by-district-year-sex-origin_past-future"),
    width = 11, height = 8,
    multi = uni_d
  )
```

### look precisely at the age plots over years

#### focus age

```{r}
#| label: fig-0320
#| fig-cap: immigration* age proportion by district, year, sex, origin; past and future, focus on age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  # plot data
  plot_a_past_pred <- mutate(pred$mis_a_past_pred,
    time = factor(
      if_else(year <= pred$mis_age_base_end, time_lev[1], time_lev[2]),
      levels = time_lev
    )
  )

  # years
  year_plot <- seq(date_start + 1, scen_end, by = 8)

  sszplot(filter(plot_a_past_pred, year %in% year_plot),
    aes_x = "age", aes_y = "prop_a", aes_col = "year",
    grid = c("sex", "origin"),
    labs_y = "proportion in %", labs_col = "year",
    fix_size = 1,
    name = plot_name("0320_immigration_star_age-proportion_by-district-year-sex-origin_past-future_focus-age"),
    width = 11, height = 8,
    multi = uni_d
  )
```

#### focus years

```{r}
#| label: fig-0321
#| fig-cap: immigration* age proportion by district, year, sex, origin; past and future, focus on years
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 7

  # age (subjectively selected)
  age_plot_pred <- seq(0, 60, by = 20)

  sszplot(filter(plot_a_past_pred, age %in% age_plot_pred),
    aes_x = "year", aes_y = "prop_a", aes_col = "age",
    grid = c("sex", "origin"),
    labs_y = "proportion in %", labs_col = "age",
    name = plot_name("0321_immigration_star_age-proportion_by-district-year-sex-origin_past-future_focus-years"),
    width = 11, height = 8,
    multi = uni_d
  )
```
