
# Relocation on immigration

```{r}
#| label: prep
#| output: false

# source general functions
source(paste0(here::here(), "/1_code/0000_general/general_utils.R"))

# read general variables for common use
params <- yaml::read_yaml(paste0(here::here(), "/Plots/_variables.yml"))

# source utility functions
run_scen(i_scen, "rei")

# assign the return values of the migration functions so they can be used inside the quarto environment
rate <- ims_rel_prop

```

## Relocation proportion

### differences by sex (sum over years)

WHY? Check if minor differences between sex? Because omitted in prediction

```{r}
#| label: fig-0500
#| fig-cap: proportion of relocation by district, age, sex and origin
#| fig-keep: all
# #| fig-subcap: !expr paste("Origin", levels(uni_o))
#| fig-width: 12
#| fig-height: 14

  # proportion by daso
  rel_prop_daso <- group_by(rate$mis_rel, district, age, sex, origin) %>%
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) %>%
    mutate(rel_prop_daso = if_else(mis == 0, NA_real_, round(rel / mis * 100, round_prop)))

  # plot: focus sex
  sszplot(rel_prop_daso,
    aes_x = "age", aes_y = "rel_prop_daso", aes_col = "sex",
    wrap = "district", ncol = 4,
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0500_proportion-daso_focus-sex"),
    width = 12, height = 14,
    multi = uni_o
  )
```

### differences by origin (sum over years and sex)

WHY? first: to emphasize the need of origin in the model second: the plots can be used to get an idea the value of an age threshold

```{r}
#| label: fig-0501
#| fig-cap: proportion of relocation by district, age and origin
#| fig-width: 11
#| fig-height: 10

  # proportion by dao
  rel_prop_dao <- group_by(rate$mis_rel, district, age, origin) %>%
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) %>%
    mutate(rel_prop_dao = if_else(mis == 0, NA_real_, round(rel / mis * 100, round_prop)))

  # plot
  sszplot(rel_prop_dao,
    aes_x = "age", aes_y = "rel_prop_dao", aes_col = "origin",
    wrap = "district", ncol = 4,
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0501_proportion-dao_focus-origin"),
    width = 11, height = 10
  )
```

### which base years?

#### year and origin

WHY? simplest plot to determine the base years

```{r}
#| label: fig-0502
#| fig-cap: proportion by year and origin
#| fig-width: 15
#| fig-height: 12

  # proportion by yo
  rel_prop_yo <- group_by(rate$mis_rel, year, origin) %>%
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) %>%
    mutate(rel_prop_yo = if_else(mis == 0, NA_real_, round(rel / mis * 100, round_prop)))

  # plot
  sszplot(rel_prop_yo,
    aes_x = "year", aes_y = "rel_prop_yo", aes_col = "origin",
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0502_proportion-yo"),
    width = 8, height = 6,
    quotes = quote(expand_limits(y = 0))
  )
```

#### year, age, origin

```{r}
#| label: fig-0503
#| fig-cap: propportion by year, age and origin
#| fig-width: 9
#| fig-height: 15

  # proportion by yao
  rel_prop_yao <- group_by(rate$mis_rel, year, age, origin) %>%
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) %>%
    rename(age_num = age) %>%
    mutate(
      rel_prop_yao = if_else(mis == 0, NA_real_, round(rel / mis * 100, round_prop)),
      age = as.factor(age_num)
    )

  # age (subjectively selected)
  age_plot <- seq(0, 80, by = 10)

  # plot
  sszplot(rel_prop_yao %>% filter(age_num %in% age_plot),
    aes_x = "year", aes_y = "rel_prop_yao", aes_col = "origin",
    grid = c("age", "."),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0503_proportion-yao"),
    width = 9, height = 15
  )
```

### aggregate over sex

This plot is the starting point (before smoothing, age corrections)

#### focus on age

```{r}
#| label: fig-0504
#| fig-cap: processes by district, age, year and origin
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 11
#| fig-height: 14

  # plot preparation
  process_lev <- c("relocation", "immigration*")

  process_plot <- rate$rem_dyao %>% gather(`rel`, `mis`, key = category, value = count) %>%
    mutate(cat = factor(if_else(category == "rel",
      process_lev[1], process_lev[2]
    ), levels = process_lev)) %>%
    select(district, year, age, origin, cat, count)


  # plot: focus age

  # years (subjectively, but last year in the plot)
  year_plot <- seq(date_end, date_start, by = -8)

  sszplot(process_plot %>% filter(year %in% year_plot),
    aes_x = "age", aes_y = "count", aes_col = "cat",
    grid = c("year", "origin"),
    labs_y = "quantity per year",
    name = plot_name("0504_processes_dyao_focus-age"),
    width = 11, height = 14,
    multi = uni_d
  )
```

#### focus on years

```{r}
#| label: fig-0505
#| fig-cap: processes by district, year, age and origin
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 12
#| fig-height: 8

  # plot: focus years

  # age (subjectively selected)
  age_plot <- seq(0, 80, by = 20)

  # with age as factor
  process_plot_age <- rename(process_plot, age_num = age) %>%
    mutate(age = as.factor(age_num))

  sszplot(process_plot_age %>% filter(age_num %in% age_plot),
    aes_x = "year", aes_y = "count", aes_col = "cat",
    grid = c("age", "origin"),
    labs_y = "quantity per year",
    name = plot_name("0505_processes_dyao_focus-years"),
    width = 12, height = 8,
    multi = uni_d
  )
```

### dyao: smoothing

#### focus age

```{r}
#| label: fig-0506
#| fig-cap: processes by district, age, year and origin
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))

  # plot preparation
  smooth_lev <- c("initial", "smoothed")

  temp_initial <- gather(rate$dyao_smooth, `rel`, `mis`, key = category, value = count) %>%
    mutate(
      cat = factor(if_else(category == "rel",
        process_lev[1], process_lev[2]
      ), levels = process_lev),
      smooth = factor(smooth_lev[1], levels = smooth_lev)
    ) %>%
    select(district, year, age, origin, cat, smooth, count)

  temp_smooth <- gather(rate$dyao_smooth, `rel_a`, `mis_a`, key = category, value = count) %>%
    mutate(
      cat = factor(if_else(category == "rel_a",
        process_lev[1], process_lev[2]
      ), levels = process_lev),
      smooth = factor(smooth_lev[2], levels = smooth_lev)
    ) %>%
    select(district, year, age, origin, cat, smooth, count)

  smooth_plot <- bind_rows(temp_initial, temp_smooth)

  # base years (subjectively, but last year in the plot)
  year_plot <- seq(rate$rem_base_end, rate$rem_base_begin, by = -7)

  # plot
  sszplot(smooth_plot %>% filter(year %in% year_plot),
    aes_x = "age", aes_y = "count", aes_col = "cat", aes_ltyp = "smooth", aes_alpha = "smooth",
    grid = c("origin", "year"),
    labs_y = "quantity per year",
    name = plot_name("0506_processes_dyao_smooth_focus-age"),
    quotes = quote(scale_alpha_manual(values = c(0.3, 1))),
    width = 15, height = 8,
    multi = uni_d
  )
```

#### focus years

```{r}
#| label: fig-0507
#| fig-cap: processes by district, year, age and origin
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))

  # age (subjectively selected)
  age_plot <- seq(0, 60, by = 20)

  # plot
  sszplot(smooth_plot %>% filter(age %in% age_plot),
    aes_x = "year", aes_y = "count", aes_col = "cat", aes_ltyp = "smooth", aes_alpha = "smooth",
    grid = c("origin", "age"),
    labs_y = "quantity per year",
    name = plot_name("0507_processes_dyao_smooth_focus-years"),
    quotes = quote(scale_alpha_manual(values = c(0.3, 1))),
    width = 15, height = 8,
    multi = uni_d
  )
```

### dyao: proportion after smoothing

#### focus age

```{r}
#| label: fig-0508
#| fig-cap: proportion smooothed by district, age, year and origin
#| fig-width: 11
#| fig-height: 10

  # base years (subjectively, but last year in the plot)
  year_plot <- seq(rate$rem_base_end, rate$rem_base_begin, by = -3)

  # plot
  sszplot(rate$prop_dyao_smooth,
    aes_x = "age", aes_y = "rel_prop_dyao", aes_col = "origin",
    wrap = "district", ncol = 4,
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0508_proportion_dyao_smooth_focus-age"),
    width = 11, height = 10,
    multi = year_plot, multif = "filter(year == x)"
  )
```

#### plot: focus year

```{r}
#| label: fig-0509
#| fig-cap: proportion smooothed by district, year, age and origin
#| fig-width: 11
#| fig-height: 10

  # age (subjectively selected)
  age_plot <- seq(0, 70, by = 10)

  # plot
  sszplot(rate$prop_dyao_smooth,
    aes_x = "year", aes_y = "rel_prop_dyao", aes_col = "origin",
    wrap = "district", ncol = 4,
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0509_proportion_dyao_smooth_focus-years"),
    title = "paste0('age: ', as.character(x))",
    width = 11, height = 10,
    multi = age_plot, multif = "filter(age == x)"
  )
```

### smoothing

#### processes

```{r}
#| label: fig-0510
#| fig-cap: processes by district, age and origin; smoothed
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 10
#| fig-height: 8

  # plot preparation
  temp_initial <- gather(rate$dao_smooth, `rel`, `mis`, key = category, value = count) %>%
    mutate(
      cat = factor(if_else(category == "rel",
        process_lev[1], process_lev[2]
      ), levels = process_lev),
      smooth = factor(smooth_lev[1], levels = smooth_lev)
    ) %>%
    select(district, age, origin, cat, smooth, count)

  temp_smooth <- gather(rate$dao_smooth, `rel_a`, `mis_a`, key = category, value = count) %>%
    mutate(
      cat = factor(if_else(category == "rel_a",
        process_lev[1], process_lev[2]
      ), levels = process_lev),
      smooth = factor(smooth_lev[2], levels = smooth_lev)
    ) %>%
    select(district, age, origin, cat, smooth, count)

  dao_smooth_plot <- bind_rows(temp_initial, temp_smooth)

  # plot
  sszplot(dao_smooth_plot,
    aes_x = "age", aes_y = "count", aes_col = "cat", aes_ltyp = "smooth", aes_alpha = "smooth",
    grid = c("origin", "."),
    labs_y = "quantity per year",
    name = plot_name("0510_processes_dao_smooth"),
    quotes = quote(scale_alpha_manual(values = c(0.3, 1))),
    width = 10, height = 8,
    multi = uni_d
  )
```

#### proportions

```{r}
#| label: fig-0511
#| fig-cap: proportion by district, age and origin; smoothed
#| fig-width: 11
#| fig-height: 10

  sszplot(rate$prop_dao_smooth,
    aes_x = "age", aes_y = "rel_prop_dao", aes_col = "origin",
    wrap = "district", ncol = 4,
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0511_proportion_dao_smooth"),
    width = 11, height = 10
  )
```

### proportion from different aggregation levels ----------------------------

#### entire curves (different aggregations levels)

```{r}
#| label: fig-0512
#| fig-cap: proportion from different aggregation levels
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 12
#| fig-height: 8

  entire_curves <- select(rate$prop_agg, district, year, age, origin, rel_prop_dyao, rel_prop_dao) %>%
    rename(dyao = rel_prop_dyao, dao = rel_prop_dao) %>%
    gather(`dyao`, `dao`, key = category, value = count) %>%
    mutate(cat = factor(category, levels = rate$sources))

  # base years (subjectively, but last year in the plot)
  year_plot <- seq(rate$rem_base_end, rate$rem_base_begin, by = -3)

  sszplot(entire_curves %>% filter(year %in% year_plot),
    aes_x = "age", aes_y = "count", aes_col = "cat",
    grid = c("year", "origin"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0512_proportion_different-aggregation-levels"),
    width = 12, height = 8,
    multi = uni_d
  )
```

#### focus age

```{r}
#| label: fig-0513
#| fig-cap: proportion from different aggregation levels, focus on age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 12
#| fig-height: 8

  sszplot(rate$prop_agg %>% filter(year %in% year_plot),
    aes_x = "age", aes_y = "rel_prop", aes_col = "rel_prop_source",
    quotes_top = quote(geom_line(aes(x = age, y = rel_prop), color = "black")),
    geom = "point",
    grid = c("year", "origin"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0513_one-proportion-from-different-aggregation-levels_focus-age"),
    width = 12, height = 8,
    multi = uni_d
  )
```

#### focus years

```{r}
#| label: fig-0514
#| fig-cap: proportion from different aggregation levels, focus on years
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 12
#| fig-height: 8

  # age (subjectively selected)
  age_plot <- seq(0, 60, by = 20)

  sszplot(rate$prop_agg %>% filter(age %in% age_plot),
    aes_x = "year", aes_y = "rel_prop", aes_col = "rel_prop_source",
    quotes_top = quote(geom_line(aes(x = year, y = rel_prop), color = "black")),
    geom = "point",
    grid = c("age", "origin"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0514_one-proportion-from-different-aggregation-levels_focus-years"),
    width = 12, height = 8,
    multi = uni_d
  )
```

### smoothing (after different levels of aggregation were brought to

#### focus age

```{r}
#| label: fig-0515
#| fig-cap: one proportion after smoothing, focus on age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))

  # plot preparation
  plot_agg_smooth <- gather(rate$prop_agg_smooth, `rel_prop`, `prop_smooth`, key = category, value = count) %>%
    mutate(cat = factor(if_else(category == "rel_prop",
      smooth_lev[1], smooth_lev[2]
    ), levels = smooth_lev)) %>%
    select(district, year, age, origin, cat, count)

  # plot
  sszplot(plot_agg_smooth %>% filter(year %in% year_plot),
    aes_x = "age", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
    grid = c("year", "origin"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0515_one-proportion_smoothing_focus-age"),
    quotes = quote(scale_alpha_manual(values = c(0.3, 1))),
    width = 15, height = 8,
    multi = uni_d
  )
```

#### focus years

```{r}
#| label: fig-0516
#| fig-cap: one proportion after smoothing, focus on year
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))

  # plot
  sszplot(plot_agg_smooth %>% filter(age %in% age_plot),
    aes_x = "year", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
    grid = c("origin", "age"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0516_one-proportion_smoothing_focus-years"),
    quotes = quote(scale_alpha_manual(values = c(0.3, 1))),
    width = 15, height = 8,
    multi = uni_d
  )
```

## Prediction

### (constrained regression)

```{r}
#| label: fig-0517
#| fig-cap: prediction by district, year, age and origin; smoothed
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))

  # plot preparation
  plot_pred_smooth <- select(rate$pred_smooth, district, year, age, origin, pred_all_age, pred_smooth) %>%
    gather(pred_smooth, `pred_all_age`, `pred_smooth`, key = category, value = count) %>%
    mutate(cat = factor(if_else(category == "pred_all_age",
      smooth_lev[1], smooth_lev[2]
    ), levels = smooth_lev)) %>%
    select(district, year, age, origin, cat, count)

  # plot: focus age

  # base years (subjectively, but last year in the plot)
  year_plot <- seq(scen_begin, scen_end, by = 7)

  # plot
  sszplot(plot_pred_smooth %>% filter(year %in% year_plot),
    aes_x = "age", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
    grid = c("origin", "year"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0517_prediction_dyao_smooth"),
    quotes = quote(scale_alpha_manual(values = c(0.3, 1))),
    width = 15, height = 8,
    multi = uni_d
  )
```

## final predictions

### past and future

```{r}
#| label: fig-0518
#| fig-cap: predicted and past proportions by district, year, age, origin
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))

  # past rate (before smoothing)
  rem_dyao_not_smoothed <- mutate(rate$rem_dyao,
    rel_prop_dyao = pmax(0, pmin(100, if_else(mis == 0, NA_real_, round(rel / mis * 100, round_prop))))
  ) %>%
    select(district, year, age, origin, rel_prop_dyao)

  # past and prediction
  rem_past_pred <- as_tibble(expand_grid(
    district = uni_d,
    year = date_start:scen_end,
    age = age_min:age_max,
    origin = uni_o
  )) %>%
    left_join(rem_dyao_not_smoothed, by = c("district", "year", "age", "origin")) %>%
    left_join(select(rate$pred_smooth, district, year, age, origin, pred_smooth),
      by = c("district", "year", "age", "origin")
    ) %>%
    mutate(prop = if_else(year < scen_begin, rel_prop_dyao, pred_smooth))

  # colors

  # plot: focus age distribution
  sszplot(rem_past_pred,
    aes_x = "age", aes_y = "prop", aes_col = "year",
    grid = c(".", "origin"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0518_proportion_dyao_past-future"),
    width = 14, height = 7,
    multi = uni_d
  )
```

### focus age distribution

WHY this plot: it is recommendable to look precisely at the age plots over years therefore, a plot that focuses on certain years

```{r}
#| label: fig-0519
#| fig-cap: predicted and past proportions by district, year, age, origin; focus age
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))

 # plot levels
  time_lev <- c("past", "future")

  # plot data
  plot_a_past_pred <- mutate(rem_past_pred,
    time = factor(if_else(year < scen_begin,
      time_lev[1], time_lev[2]
    ), levels = time_lev)
  )

  # years
  year_temp <- date_start:scen_end
  year_plot <- seq(min(year_temp), max(year_temp), by = 8)

  sszplot(plot_a_past_pred %>% filter(year %in% year_plot),
    aes_x = "age", aes_y = "prop", aes_col = "year", aes_alpha = "time",
    grid = c(".", "origin"),
    labs_y = paste0("proportion in % (relocation on immigration*)"),
    name = plot_name("0519_proportion_dyao_past-future_focus-age"),
    quotes = quote(scale_alpha_manual(values = c(0.4, 1))),
    width = 14, height = 7,
    multi = uni_d
  )
```

### focus year distribution

```{r}
#| label: fig-0520
#| fig-cap: predicted and past proportions by district, year, age, origin; focus year
#| fig-keep: all
# #| fig-subcap: !expr paste("District of", levels(uni_d))
#| fig-width: 14
#| fig-height: 6

  sszplot(plot_a_past_pred %>% filter(age %in% age_plot),
    aes_x = "year", aes_y = "prop", aes_col = "age",
    i_x = c(rate$rem_base_begin, rate$rem_base_end),
    grid = c(".", "origin"),
    labs_y = paste0("proportion in % (relocation on immigration*)"), labs_col = "age",
    name = plot_name("0520_proportion_dyao_past-future_focus-years"),
    quotes = quote(scale_alpha_manual(values = c(0.4, 1))),
    width = 14, height = 6,
    multi = uni_d
  )
```
