
# Allocation

```{r}
#| label: prep
#| output: false

# source general functions
source(paste0(here::here(), "/1_code/0000_general/general_utils.R"))

# read general variables for common use
params <- yaml::read_yaml(paste0(here::here(), "/Plots/_variables.yml"))

# source utility functions
run_scen(params$scen, "aca")

```

## allocation yearly

### entire city by year
WHY? e.g. to support the base year decision

```{r}
#| label: fig-1000
#| fig-cap: allocation by year

aca_y <- group_by(aca_dat, year) %>%
  summarize(
    apartments = sum_NA(apartments),
    people = sum_NA(people)
  ) %>%
  ungroup() %>%
  mutate(aca_y = round(people / apartments, round_aca))

sszplot(aca_y,
  aes_x = "year", aes_y = "aca_y",
  labs_y = "allocation (people per apartment)",
  scale_y = c(0, NA),
  name = plot_name("1000_allocation_y"),
  width = 8, height = 5
)
```

### entire city by year and owner

```{r}
#| label: fig-1001
#| fig-cap: allocation by year and ownership

sszplot(aca_yw,
  aes_x = "year", aes_y = "aca_yw", aes_col = "owner",
  labs_y = "allocation (people per apartment)",
  scale_y = c(0, NA),
  name = plot_name("1001_allocation_yw"),
  width = 8, height = 5
)
```

### allocation by district, year, owner

```{r}
#| label: fig-1002
#| fig-cap: allocation by district, year and ownership
#| fig-width: 12
#| fig-height: 14

sszplot(aca_dyw,
  aes_x = "year", aes_y = "aca_dyw", aes_col = "owner",
  labs_y = "allocation (people per apartment)",
  wrap = "district", ncol = 4,
  scale_y = c(0, NA),
  name = plot_name("1002_allocation_dyw"),
  width = 12, height = 14
)
```

### apartments by district, year, owner

```{r}
#| label: fig-1003
#| fig-cap: apartments by district, year and ownership
#| fig-width: 12
#| fig-height: 14

sszplot(aca_dyw,
  aes_x = "year", aes_y = "apartments", aes_col = "owner",
  labs_y = "apartments",
  wrap = "district", ncol = 4,
  scale_y = c(0, NA),
  name = plot_name("1003_apartments_dyw"),
  width = 12, height = 14
)
```

### people by district, year, owner

```{r}
#| label: fig-1004
#| fig-cap: people by district, year and ownership
#| fig-width: 12
#| fig-height: 14

sszplot(aca_dyw,
  aes_x = "year", aes_y = "people", aes_col = "owner",
  labs_y = "people",
  wrap = "district", ncol = 4,
  scale_y = c(0, NA),
  name = plot_name("1004_people_dyw"),
  width = 12, height = 14
)
```

## predictions

### entire city, by owner

```{r}
#| label: fig-1005
#| fig-cap: allocation prediction by year and ownership
#| fig-width: 10
#| fig-height: 7

sszplot(aca_yw_past_pred,
  aes_x = "year", aes_y = "aca_yw_all", aes_col = "owner",
  labs_y = "allocation (people per apartment)",
  i_x = c(aca_base_begin, aca_base_end),
  scale_y = c(0, NA),
  name = plot_name("1005_allocation_prediction_yw"),
  width = 10, height = 7
)
```

### entire city, by district and owner

```{r}
#| label: fig-1006
#| fig-cap: allocation prediction by district, year and ownership
#| fig-width: 12
#| fig-height: 14

sszplot(aca_dyw_past_pred,
  aes_x = "year", aes_y = "aca_dyw_all", aes_col = "owner",
  labs_y = "allocation (people per apartment)",
  wrap = "district", ncol = 4,
  i_x = c(aca_base_begin, aca_base_end),
  scale_y = c(0, NA),
  name = plot_name("1006_allocation_prediction_dyw"),
  width = 12, height = 14
)
```

### Escher Wyss, by owner
WHY a different approach for the Escher Wyss district, private housing?
same approach as in the living space model (see @sec-escherwyss_ls)

```{r}
#| label: fig-1007
#| fig-cap: allocation prediction for Escher Wyss district, by year and ownership
#| fig-width: 10
#| fig-height: 7

sszplot(aca_yw_past_pred_52,
  aes_x = "year", aes_y = "aca_dyw_all", aes_col = "owner",
  labs_y = "allocation (people per apartment)",
  i_x = c(aca_base_begin_52p, aca_base_end),
  scale_y = c(0, NA),
  name = plot_name("1007_allocation_prediction_52yw"),
  width = 10, height = 7
)
```

## combine datasets

###  past and prediction

```{r}
#| label: fig-1008
#| fig-cap: allocation prediction by district, year and owner; different datasets
#| fig-width: 12
#| fig-height: 14

sszplot(aca_comb,
  aes_x = "year", aes_y = "aca", aes_col = "owner",
  labs_y = "allocation (people per apartment)",
  wrap = "district", ncol = 4,
  i_x = c(spa_base_begin, spa_base_end),
  scale_y = c(0, NA),
  name = plot_name("1008_allocation_prediction_dyw_different-datasets"),
  width = 12, height = 14
)
```

### plot both (i.e. per district, and entire city)

```{r}
#| label: fig-1009
#| fig-cap: allocation prediction by district, year and owner plus entire city
#| fig-width: 10
#| fig-height: 5
#| fig-keep: all
#| fig-subcap: !expr paste("Origin", levels(uni_d))

names_before <- c("aca_dyw", "aca_yw", "aca")
names_plot <- c("by district", "entire city", "prediction")

aca_both <- select(aca_comb, district, year, owner, aca_dyw, aca_yw, aca) %>%
  pivot_longer(cols = names_before, names_to = "category", values_to = "aca") %>%
  mutate(cat = factor(case_when(
    category == names_before[1] ~ names_plot[1],
    category == names_before[2] ~ names_plot[2],
    TRUE ~ names_plot[3]
  ), levels = names_plot))

sszplot(aca_both,
  aes_x = "year", aes_y = "aca", aes_col = "cat",
  grid = c(".", "owner"),
  labs_y = "allocation (people per apartment)",
  i_x = c(spa_base_begin, spa_base_end),
  scale_y = c(0, NA),
  name = plot_name("1009_allocation_dyw-yw"),
  width = 8, height = 4,
  multi = uni_d
)
```
