
# Naturalization

```{r}
#| label: prep

# source general functions
source(paste0(here::here(), "/1_code/0000_general/general_utils.R"))

# read general variables for common use
params <- yaml::read_yaml(paste0(here::here(), "/Plots/_variables.yml"))

# source utility functions
run_scen(params$scen, "nat")

```

## naturalization vs. denaturalization

```{r}
#| label: fig-proc
#| fig-cap: naturalization on citizenship change

# process levels
processes <- c("nat", "denat")

# processes
proc <- filter(nat_denat, origin_prev != origin) %>%
  mutate(proc = factor(if_else(origin_prev == uni_o[2],
    processes[1], processes[2]
  ), levels = processes)) %>%
  group_by(year, proc) %>%
  summarize(
    count = sum(nat),
    .groups = "drop"
  ) %>%
  spread(key = proc, value = count) %>%
  replace_na(list(nat = 0, denat = 0)) %>%
  mutate(
    total = nat + denat,
    prop_nat = if_else(total == 0, NA_real_, round(nat / total * 100, round_prop))
  )

# plot
sszplot(proc,
  aes_x = "year", aes_y = "prop_nat",
  labs_y = "proportion in % (naturalization on citizenship change)",
  name = plot_name("0700_naturalization-on-citizenship-change"),
  width = 8, height = 6
)
```

## naturalization rate: which variables?

### age and sex

```{r}
#| label: fig-nat_as
#| fig-cap: rate age and sex

# rate by as
nat_as <- group_by(nat_pop, age, sex) %>%
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) %>%
  mutate(nat_as = if_else(pop == 0, NA_real_, round(nat / pop * 100, round_rate)))

sszplot(nat_as,
  aes_x = "age", aes_y = "nat_as", aes_col = "sex",
  labs_y = "naturalization rate (in % per year)",
  scale_y = c(0, NA),
  name = plot_name("0701_rate_as"),
  width = 8, height = 6
)
```

### year and sex

```{r}
#| label: fig-nat_ys
#| fig-cap: rate year and sex

# rate by ys
nat_ys <- group_by(nat_pop, year, sex) %>%
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) %>%
  mutate(nat_ys = if_else(pop == 0, NA_real_, round(nat / pop * 100, round_rate)))

sszplot(nat_ys,
  aes_x = "year", aes_y = "nat_ys", aes_col = "sex",
  labs_y = "naturalization rate (in % per year)",
  scale_y = c(0, NA),
  name = plot_name("0702_rate_ys"),
  width = 8, height = 6
)
```

### year, age and sex

```{r}
#| label: fig-nat_yas_age
#| fig-cap: rate year, age, sex; focus on age
#| fig-width: 10
#| fig-height: 10

# rate by yas
nat_yas <- group_by(nat_pop, year, age, sex) %>%
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) %>%
  mutate(nat_yas = if_else(pop == 0, NA_real_, round(nat / pop * 100, round_rate)))


# years (subjectively, but last year in the plot)
year_plot <- seq(date_end, (date_start + 1), by = -4)

sszplot(filter(nat_yas, year %in% year_plot),
  aes_x = "age", aes_y = "nat_yas", aes_col = "sex",
  labs_y = "naturalization rate (in % per year)",
  wrap = "year", ncol = 3,
  name = plot_name("0703_rate_yas_focus-age"),
  width = 10, height = 10
)
```

```{r}
#| label: fig-nat_yas_year
#| fig-cap: rate year, age, sex; focus on year
#| fig-width: 8
#| fig-height: 10

# age (subjectively selected)
age_plot <- seq(0, 70, by = 10)

sszplot(filter(nat_yas, age %in% age_plot),
  aes_x = "year", aes_y = "nat_yas", aes_col = "sex",
  labs_y = "naturalization rate (in % per year)",
  wrap = "age", ncol = 2,
  name = plot_name("0704_rate_yas_focus-years"),
  width = 8, height = 10
)
```

### district, age and sex

```{r}
#| label: fig-nat_das
#| fig-cap: rate district, age and sex
#| fig-width: 12
#| fig-height: 14

# rate by das
nat_das <- group_by(nat_pop, district, age, sex) %>%
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) %>%
  mutate(nat_das = if_else(pop == 0, NA_real_, round(nat / pop * 100, round_rate)))

sszplot(nat_das,
  aes_x = "age", aes_y = "nat_das", aes_col = "sex",
  labs_y = "naturalization rate (in % per year)",
  scale_y = c(0, 25),
  wrap = "district", ncol = 4,
  name = plot_name("0705_rate_das"),
  width = 12, height = 14
)


```

### district, year and sex

```{r}
#| label: fig-nat_dys
#| fig-cap: rate district, year and sex
#| fig-width: 12
#| fig-height: 14

# rate by dys
nat_dys <- group_by(nat_pop, district, year, sex) %>%
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) %>%
  mutate(nat_dys = if_else(pop == 0, NA_real_, round(nat / pop * 100, round_rate)))

sszplot(nat_dys,
  aes_x = "year", aes_y = "nat_dys", aes_col = "sex",
  labs_y = "naturalization rate (in % per year)",
  wrap = "district", ncol = 4,
  name = plot_name("0706_rate_dys"),
  width = 12, height = 14
)
```

## das: smoothing (nat and pop)

```{r}
#| label: fig-smooth_plot
#| fig-cap: processes by district, age, sex; smoothed
#| fig-keep: all
#| #| fig-subcap: !expr paste("Origin", levels(uni_d))

# plot preparation
process_lev <- c("naturalization", "population")
smooth_lev <- c("initial", "smoothed")

temp_initial <- gather(nat_pop_das_smooth, `nat`, `pop`, key = category, value = count) %>%
  mutate(
    cat = factor(if_else(category == "nat",
      process_lev[1], process_lev[2]
    ), levels = process_lev),
    smooth = factor(smooth_lev[1], levels = smooth_lev)
  ) %>%
  select(district, age, sex, cat, smooth, count)

temp_smooth <- gather(nat_pop_das_smooth, `nat_a`, `pop_a`, key = category, value = count) %>%
  mutate(
    cat = factor(if_else(category == "nat_a",
      process_lev[1], process_lev[2]
    ), levels = process_lev),
    smooth = factor(smooth_lev[2], levels = smooth_lev)
  ) %>%
  select(district, age, sex, cat, smooth, count)

smooth_plot <- bind_rows(temp_initial, temp_smooth)

# plot
sszplot(smooth_plot,
  aes_x = "age", aes_y = "count", aes_col = "cat", aes_ltyp = "smooth", aes_alpha = "smooth",
  labs_y = "quantity per year (naturalization), quantity (population)",
  grid = c("cat", "sex"), gridscale = "free_y",
  name = plot_name("0707_processes_das_smooth"),
  width = 15, height = 8,
  multi = uni_d,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

## das: naturalization rate

```{r}
#| label: fig-rate_plot
#| fig-cap: rate by district, age and sex
#| fig-width: 12
#| fig-height: 5
#| fig-keep: all
#| fig-subcap: !expr paste("Origin", levels(uni_d))

# plot preparation
rate_plot <- gather(nat_rate_das, `rate_thres`, `rate_smooth`, key = category, value = count) %>%
  mutate(cat = factor(if_else(category == "rate_thres",
    smooth_lev[1], smooth_lev[2]
  ), levels = smooth_lev)) %>%
  select(district, age, sex, cat, count)

# plot
sszplot(rate_plot,
  aes_x = "age", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
  labs_y = "naturalization rate (in % per year)",
  grid = c(".", "sex"),
  name = plot_name("0708_rate_das"),
  width = 12, height = 5,
  multi = uni_d,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

## ya: smooth processes (preparation for trend by ya)

### by years (subjectively, but last year in the plot)

```{r}
#| label: fig-smooth_plot_ya_age
#| fig-cap: processes by year and age; smoothed, focus on age

# plot preparation
temp_initial_ya <- gather(nat_ya, `nat`, `pop`, key = category, value = count) %>%
  mutate(
    cat = factor(if_else(category == "nat",
      process_lev[1], process_lev[2]
    ), levels = process_lev),
    smooth = factor(smooth_lev[1], levels = smooth_lev)
  ) %>%
  select(year, age, cat, smooth, count)

temp_smooth_ya <- gather(nat_ya, `nat_a`, `pop_a`, key = category, value = count) %>%
  mutate(
    cat = factor(if_else(category == "nat_a",
      process_lev[1], process_lev[2]
    ), levels = process_lev),
    smooth = factor(smooth_lev[2], levels = smooth_lev)
  ) %>%
  select(year, age, cat, smooth, count)

smooth_plot_ya <- bind_rows(temp_initial_ya, temp_smooth_ya)

year_plot <- seq(nat_base_end, nat_base_begin, by = -4)

sszplot(filter(smooth_plot_ya, year %in% year_plot),
  aes_x = "age", aes_y = "count", aes_col = "cat", aes_ltyp = "smooth", aes_alpha = "smooth",
  labs_y = "quantity per year (naturalization), quantity (population)",
  grid = c("cat", "year"), gridscale = "free_y",
  name = plot_name("0709_processes_ya_smooth_focus-age"),
  width = 15, height = 8,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

### by age (subjectively selected)

```{r}
#| label: fig-smooth_plot_ya_year
#| fig-cap: processes by year and age; smoothed, focus on years

age_plot <- seq(0, 100, by = 20)

sszplot(filter(smooth_plot_ya, age %in% age_plot),
  aes_x = "year", aes_y = "count", aes_col = "cat", aes_ltyp = "smooth", aes_alpha = "smooth",
  labs_y = "quantity per year (naturalization), quantity (population)",
  grid = c("cat", "age"), gridscale = "free_y",
  name = plot_name("0710_processes_ya_smooth_focus-years"),
  width = 15, height = 8,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

## ya: rate

### by years (subjectively, but last year in the plot)

```{r}
#| label: fig-rate_ya_plot_year
#| fig-cap: rate by year and age, focus on age

# plot preparation
rate_ya_plot <- gather(nat_rate_ya, `rate_thres`, `rate_smooth`, key = category, value = count) %>%
  mutate(cat = factor(if_else(category == "rate_thres",
    smooth_lev[1], smooth_lev[2]
  ), levels = smooth_lev)) %>%
  select(year, age, cat, count)

# years (subjectively, but last year in the plot)
year_plot <- seq(nat_base_end, nat_base_begin, by = -2)

sszplot(filter(rate_ya_plot, year %in% year_plot),
  aes_x = "age", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
  labs_y = "naturalization rate (in % per year)",
  wrap = "year", ncol = 3,
  name = plot_name("0711_rate_ya_focus-age"),
  width = 15, height = 8,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

### by age (subjectively selected)

```{r}
#| label: fig-rate_ya_plot_age
#| fig-cap: rate by year and age, focus on years

sszplot(filter(rate_ya_plot, age %in% age_plot),
  aes_x = "year", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
  labs_y = "naturalization rate (in % per year)",
  grid = c(".", "age"),
  name = plot_name("0712_rate_ya_focus-years"),
  width = 15, height = 6,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

## constrained regression

### focus age

```{r}
#| label: fig-rate_ya_past_pred_age
#| fig-cap: rate by year and age; past and future, focus on age
#| fig-width: 10
#| fig-height: 7

# years (subjectively, but last year in the plot)
year_plot <- seq(nat_base_begin, scen_end, by = 10)

sszplot(rate_ya_past_pred,
  aes_x = "age", aes_y = "rate_ya", aes_col = "year",
  labs_y = "naturalization rate (in % per year)",
  name = plot_name("0713_rate_ya_past-future_focus-age"),
  width = 10, height = 7
)
```

### focus years

```{r}
#| label: fig-rate_ya_past_pred_year
#| fig-cap: rate by year and age; past and future, focus on years
#| fig-width: 10
#| fig-height: 7

# age (subjectively selected)
age_plot <- seq(0, 100, by = 10)

sszplot(filter(rate_ya_past_pred, age %in% age_plot),
  aes_x = "year", aes_y = "rate_ya",
  labs_y = "naturalization rate (in % per year)",
  wrap = "age", ncol = 5,
  name = plot_name("0714_rate_ya_past-future_focus-years"),
  width = 10, height = 7
)
```

## smooth processes (preparation for the trend factor calculation)

```{r}
#| label: fig-smooth_plot_a
#| fig-cap: processes by age, smoothed
#| fig-width: 10
#| fig-height: 8

# plot preparation
temp_initial_a <- gather(nat_a, `nat`, `pop`, key = category, value = count) %>%
  mutate(
    cat = factor(if_else(category == "nat",
      process_lev[1], process_lev[2]
    ), levels = process_lev),
    smooth = factor(smooth_lev[1], levels = smooth_lev)
  ) %>%
  select(age, cat, smooth, count)

temp_smooth_a <- gather(nat_a, `nat_a`, `pop_a`, key = category, value = count) %>%
  mutate(
    cat = factor(if_else(category == "nat_a",
      process_lev[1], process_lev[2]
    ), levels = process_lev),
    smooth = factor(smooth_lev[2], levels = smooth_lev)
  ) %>%
  select(age, cat, smooth, count)

smooth_plot_a <- bind_rows(temp_initial_a, temp_smooth_a)


# plots
sszplot(smooth_plot_a,
  aes_x = "age", aes_y = "count", aes_col = "cat", aes_ltyp = "smooth", aes_alpha = "smooth",
  labs_y = "quantity per year (naturalization), quantity (population)",
  grid = c("cat", "."), gridscale = "free_y",
  name = plot_name("0715_processes_a_smooth"),
  width = 10, height = 8,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

## rate

```{r}
#| label: fig-rate_a_plot
#| fig-cap: rate by age
#| fig-width: 12
#| fig-height: 7

# plot preparation
rate_a_plot <- gather(nat_rate_a, `rate_thres`, `rate_smooth`, key = category, value = count) %>%
  mutate(cat = factor(if_else(category == "rate_thres",
    smooth_lev[1], smooth_lev[2]
  ), levels = smooth_lev)) %>%
  select(age, cat, count)

# plots
sszplot(rate_a_plot,
  aes_x = "age", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
  labs_y = "naturalization rate (in % per year)",
  name = plot_name("0716_rate_a"),
  width = 12, height = 7,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

## trend factor ('rate ya' to 'rate a')

```{r}
#| label: fig-tf_ya_age
#| fig-cap: trend factor by year and age, focus on age
#| fig-width: 10
#| fig-height: 8

sszplot(tf_ya,
  aes_x = "age", aes_y = "tf_ya", aes_col = "year",
  i_y = 1,
  labs_y = "trend factor",
  scale_y = c(0, 2),
  name = plot_name("0717_trend-factor_focus-age"),
  width = 10, height = 8
)
```

```{r}
#| label: fig-tf_ya_year
#| fig-cap: trend factor by year and age, focus on years
#| fig-width: 10
#| fig-height: 8

# age (subjectively selected)
age_plot <- seq(0, 100, by = 10)

sszplot(filter(tf_ya, age %in% age_plot),
  aes_x = "year", aes_y = "tf_ya",
  i_y = 1,
  labs_y = "trend factor",
  wrap = "age", ncol = 4,
  name = plot_name("0718_trend-factor_focus-years"),
  width = 10, height = 8
)
```

## future rate (dyas)

```{r}
#| label: fig-plot_pred_smooth
#| fig-cap: prediction by district, year, age and origin; smoothed
#| fig-keep: all
#| fig-subcap: !expr paste("Origin", levels(uni_d))

# plot preparation
plot_pred_smooth <- select(rate_dyas_future, district, year, age, sex, rate_dyas, rate_smoothed) %>%
  gather(`rate_dyas`, `rate_smoothed`, key = category, value = count) %>%
  mutate(cat = factor(if_else(category == "rate_dyas",
    smooth_lev[1], smooth_lev[2]
  ), levels = smooth_lev)) %>%
  select(district, year, age, sex, cat, count)

# base years (subjectively)
year_plot <- seq(scen_begin, scen_end, by = 7)

# plot
sszplot(filter(plot_pred_smooth, year %in% year_plot),
  aes_x = "age", aes_y = "count", aes_ltyp = "cat", aes_alpha = "cat",
  labs_y = "naturalization rate (in % per year)",
  grid = c("sex", "year"),
  name = plot_name("0719_prediction_dyao_smooth"),
  width = 15, height = 8,
  multi = uni_d,
  quotes = quote(scale_alpha_manual(values = c(0.3, 1)))
)
```

## past and rate (dyas)

### overview

```{r}
#| label: fig-rate_dyas_past_future
#| fig-cap: rate by district, year, age and origin; past and future
#| fig-keep: all
#| fig-subcap: !expr paste("Origin", levels(uni_d))

# past
rate_dyas_past <- group_by(nat_pop, district, year, age, sex) %>%
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) %>%
  mutate(rate_past = pmin(100, if_else(pop == 0, NA_real_, round(nat / pop * 100, round_rate))))

# past and future
rate_dyas_past_future <- as_tibble(expand_grid(
  district = uni_d,
  year = (date_start + 1):scen_end,
  age = age_min:age_max,
  sex = uni_s
)) %>%
  left_join(select(rate_dyas_past, district, year, age, sex, rate_past),
    by = c("district", "year", "age", "sex")
  ) %>%
  left_join(select(rate_dyas_future, district, year, age, sex, rate_smoothed),
    by = c("district", "year", "age", "sex")
  ) %>%
  replace_na(list(rate_past = 0, rate_smoothed = 0)) %>%
  mutate(rate_dyas = if_else(year < scen_begin, rate_past, rate_smoothed))

# plot: focus age distribution
sszplot(rate_dyas_past_future,
  aes_x = "age", aes_y = "rate_dyas", aes_col = "year",
  labs_y = "naturalization rate (in % per year)",
  grid = c(".", "sex"),
  name = plot_name("0720_rate_dyao_past-future"),
  width = 14, height = 7,
  multi = uni_d
)
```

### focus age

plot: focus age distribution WHY this plot: it is recommendable to look precisely at the age plots over years therefore, a plot that focuses on certain years

```{r}
#| label: fig-plot_a_past_pred_age
#| fig-cap: rate by district, year, age and origin; past and future, focus on age
#| fig-keep: all
#| fig-subcap: !expr paste("Origin", levels(uni_d))

# plot levels
time_lev <- c("past", "future")

# plot data
plot_a_past_pred <- mutate(rate_dyas_past_future,
  time = factor(if_else(year < scen_begin,
    time_lev[1], time_lev[2]
  ), levels = time_lev)
)

# years
year_plot <- seq(date_start + 1, scen_end, by = 7)

sszplot(filter(plot_a_past_pred, year %in% year_plot),
  aes_x = "age", aes_y = "rate_dyas", aes_col = "year",
  labs_y = "naturalization rate (in % per year)", labs_col = "year",
  grid = c(".", "sex"),
  name = plot_name("0721_rate_dyao_past-future_focus-age"),
  width = 14, height = 7,
  multi = uni_d
)
```

### focus years

```{r}
#| label: fig-plot_a_past_pred_year
#| fig-cap: rate by district, year, age and origin; past and future, focus on years
#| fig-keep: all
#| fig-subcap: !expr paste("Origin", levels(uni_d))

# age (subjectively selected)
age_plot_pred <- seq(0, 60, by = 20)

sszplot(filter(plot_a_past_pred, age %in% age_plot_pred),
  aes_x = "year", aes_y = "rate_dyas", aes_col = "age",
  labs_y = "naturalization rate (in % per year)", labs_col = "age",
  grid = c(".", "sex"),
  name = plot_name("0722_rate_dyao_past-future_focus-years"),
  width = 14, height = 6,
  multi = uni_d
)
```
