# Birth

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the birth functions
source(file.path(root, "R/0101_fertility_rate_fun.R"))
source(file.path(root, "R/0102_birth_origin_change_fun.R"))
source(file.path(root, "R/0103_birth_sex_ratio_fun.R"))

# execute the birth functions
fer <- fertility_rate_fun(params)
boc <- birth_origin_change_fun(params)
bsr <- birth_sex_ratio_fun(params)

```

## Fertility

### TFR (total fertility rate)

#### plots to define the base period for fertility prediction

```{r}
#| label: fig-0100
#| fig-cap: "total fertility rate by year"
#| fig-width: 8
#| fig-height: 6

(fer$fer_ya |> 
  group_by(year) |>
  summarize(
    tfr_y = sum_NA(fer_ya / 100),
    .groups = "drop"
  ) |> 
  ggplot(aes(x = year, y = tfr_y)) +
  geom_line(col = params$col_1) +
  labs(x = "", y = "TFR") +
  params$neutral) |> 
  ggplotly()

```

```{r}
#| label: fig-0101
#| fig-cap: "total fertility rate by year and origin"
#| fig-width: 8
#| fig-height: 6

(fer$fer_yao |> 
  group_by(year, origin) |>
  summarize(
    tfr_yo = sum_NA(fer_yao / 100),
    .groups = "drop"
  ) |> 
ggplot(aes(x = year, y = tfr_yo, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +  
  labs(x = "", y = "TFR", color = "origin") +
  params$neutral) |> 
ggplotly()

```




```{r}
#| label: fig-0102
#| fig-cap: "total fertility rate by year and age class (3)"
#| fig-width: 8
#| fig-height: 6

(fer$fer_ya |> 
  left_join(params$look_a1, by = "age") |>
  group_by(year, age_1) |>
  summarize(
    tfr_ya1 = sum_NA(fer_ya / 100),
    .groups = "drop"
  ) |>
ggplot(aes(x = year, y = tfr_ya1, color = age_1)) +
  geom_line() +
  scale_color_manual(values = params$col_a1) +
  labs(x = "", y = "TFR", color = "age") +
  params$neutral) |> 
ggplotly()


```

```{r}
#| label: fig-0103
#| fig-cap: "total fertility rate by year and age class (5)"
#| fig-width: 8
#| fig-height: 6

(fer$fer_ya |> 
  left_join(params$look_a2, by = "age") |>
  group_by(year, age_2) |>
  summarize(
    tfr_ya2 = sum_NA(fer_ya / 100),
    .groups = "drop"
  ) |>
ggplot(aes(x = year, y = tfr_ya2, color = age_2)) +
  geom_line() +
  scale_color_manual(values = params$col_a2) +
  labs(x = "", y = "TFR", color = "age") +
  params$neutral) |> 
ggplotly()

```

```{r}
#| label: fig-0104
#| fig-cap: "total fertility rate by year, origin and age class (3)"
#| fig-width: 16
#| fig-height: 10

(fer$fer_yao |> 
  left_join(params$look_a1, by = "age") |>
  group_by(year, age_1, origin) |>
  summarize(
    tfr_ya1o = sum_NA(fer_yao / 100),
    .groups = "drop"
  ) |> 
ggplot(aes(x = year, y = tfr_ya1o, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +
  facet_wrap(~age_1, nrow = 1) +
  labs(x = "", y = "TFR", color = "origin") +
  params$neutral) |> 
ggplotly()

```

```{r}
#| label: fig-0105
#| fig-cap: "total fertility rate by year, origin and age class (5)"
#| fig-width: 16
#| fig-height: 10

(fer$fer_yao |> 
  left_join(params$look_a2, by = "age") |>
  group_by(year, age_2, origin) |>
  summarize(
    tfr_ya2o = sum_NA(fer_yao / 100),
    .groups = "drop"
  ) |>
ggplot(aes(x = year, y = tfr_ya2o, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +
  facet_wrap(~age_2, nrow = 1) +
  labs(x = "", y = "TFR", color = "origin") +
  params$neutral) |> 
ggplotly()

```


```{r}
#| label: fig-0106
#| fig-cap: "total fertility rate by district, year and origin"
#| fig-width: 11
#| fig-height: 7

fer$fer_dyao |>
  group_by(district, year, origin) |>
  summarize(
    tfr = sum_NA(fer_dyao / 100),
    .groups = "drop"
  ) |> 
plot_ly(
  x = ~year,
  y = ~tfr,
  color = ~origin,
  colors = params$col_o, 
  frame = ~district
) %>%
  add_lines() %>%
  layout(xaxis = list(title = ""), 
         yaxis = list(title = "TFR"))

```



#### fertility plots (before the corrections)

fertility by district, age, origin (last year only)

```{r}
#| label: fig-0110
#| fig-cap: "fertility by district, age, and origin (last year)"
#| fig-width: 20
#| fig-height: 20

(fer$fer_dyao |>
  filter(year == max(year)) |>
  arrange(district, origin, age) |> 
ggplot(aes(x = age, y = fer_dyao, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +
  facet_wrap(~district, ncol = 4) +
  labs(x = "age", y = "fertility rate (in % per year)", color = "origin") +
  params$neutral) |> 
ggplotly(height = 2000)

```



```{r}
#| label: fig-0111
#| fig-cap: "fertility by year, age, origin"
#| fig-width: 11
#| fig-height: 7

fer_yao_plot <- fer$fer_yao |> 
  filter(year >= params$bir_base_begin) 

n_years <- length(unique(fer_yao_plot$year))
pal_year <- colorRampPalette(params$col_6[3:1])(n_years)

(ggplot(fer_yao_plot, aes(x = age, y = fer_yao, color = as.factor(year))) +
  geom_line() +
  scale_color_manual(values = pal_year) +
  facet_wrap(~origin, nrow = 1) +
  labs(x = "age", y = "fertility rate (in % per year)", color = "") +
  params$neutral) |> 
ggplotly()

```



```{r}
#| label: fig-0112
#| fig-cap: "fertility by year, age"
#| fig-width: 11
#| fig-height: 7

(fer$fer_ya |> 
  filter(year >= params$bir_base_begin) |> 
ggplot(aes(x = age, y = fer_ya, color = as.factor(year))) +
  geom_line() +
  scale_color_manual(values = pal_year) +
  labs(x = "age", y = "fertility rate (in % per year)", color = "") +
  params$neutral) |> 
ggplotly()

```

### corrected fertility (tail corrections)

```{r}
#| label: fig-0120
#| fig-cap: "initial vs. corrected fertility (last year only)"
#| fig-width: 15
#| fig-height: 12

cor_level <- c("initial", "corrected")

fer$fer_tail |> 
  filter(year == max(year)) |> 
  select(district, origin, age, fer_dyao, fer) |>
  pivot_longer(c(fer_dyao, fer), names_to = "category", values_to = "fer") |>
  mutate(cat = fact_if(category, cor_level, "fer_dyao")) |>
  select(district, origin, age, cat, fer) |> 
plot_ly(
  x = ~age,
  y = ~fer,
  color = ~origin,
  colors = params$col_o,
  linetype = ~cat,      
  frame = ~district,    
  type = 'scatter',
  mode = 'lines'
) %>%
  layout(
    xaxis = list(title = "age"),
    yaxis = list(title = "fertility rate (in % per year)"),
    legend = list(title = list(text = ""))
  )

```

cumulative population from the tails (last year only)

```{r}
#| label: fig-0121
#| fig-cap: "cumulative population from the tails (last year only)"
#| fig-width: 20
#| fig-height: 20

(fer$fer_tail |> 
  filter(year == max(year)) |> 
  ggplot() +
  geom_line(aes(x = age, y = low_cum, color = origin)) + 
  geom_line(aes(x = age, y = up_cum, color = origin)) +   
  facet_wrap(~district, ncol = 4) + 
  scale_color_manual(values = params$col_o) +
  labs(x = "age", y = "cumulative population from tails") + 
  params$neutral) |> 
ggplotly(height = 2000)

```



```{r}
#| label: fig-0122
#| fig-cap: smooth the corrected fertility rate (last year only)
#| fig-width: 12
#| fig-height: 10

fit_lev <- c("initial", "smoothed")

fer$fer_fit |> 
  filter(year == max(year)) |> 
  select(district, origin, age, fer, fer_fit) |>
  pivot_longer(c(fer, fer_fit), names_to = "category", values_to = "fer") |>
  mutate(cat = fact_if(category, fit_lev, "fer")) |>
  select(district, age, origin, cat, fer) |> 
plot_ly(
  x = ~age,
  y = ~fer,
  color = ~origin,
  colors = params$col_o,
  linetype = ~cat,      
  frame = ~district,    
  type = 'scatter',
  mode = 'lines'
) %>%
  layout(
    xaxis = list(title = "age"),
    yaxis = list(title = "fertility rate (in % per year)"),
    legend = list(title = list(text = ""))
)

```

### Prediction

#### fertility rate: past and future

fertility rate: age distribution by district and year (Swiss only)

```{r}
#| label: fig-0140
#| fig-cap: "fertility rate: age distribution by district and year (Swiss only)"
#| fig-width: 20
#| fig-height: 20

n_years <- length(unique(fer$fer_pred$year))
pal_year <- colorRampPalette(params$col_6[1:5])(n_years)

(fer$fer_pred |> 
  filter(origin == params$uni_o[1]) |> 
  ggplot(aes(x = age, y = fer_all, color = as.factor(year))) +
  geom_line() +           
  facet_wrap(~district, ncol = 4) + 
  scale_color_manual(values = pal_year) +    
  labs(x = "age", y = "fertility rate (in % per year)", color = "") +
  params$neutral) |> 
ggplotly(height = 2000)

```

fertility rate: age distribution by district and year (foreign only)

```{r}
#| label: fig-0141
#| fig-cap: "fertility rate: age distribution by district and year (foreign only)"
#| fig-width: 20
#| fig-height: 20

n_years <- length(unique(fer$fer_pred$year))
pal_year <- colorRampPalette(params$col_6[1:5])(n_years)

(fer$fer_pred |> 
  filter(origin == params$uni_o[2]) |> 
  ggplot(aes(x = age, y = fer_all, color = as.factor(year))) +
  geom_line() +           
  facet_wrap(~district, ncol = 4) + 
  scale_color_manual(values = pal_year) +    
  labs(x = "age", y = "fertility rate (in % per year)", color = "") +
  params$neutral) |> 
ggplotly(height = 2000)

```


#### TFR: past and future

```{r}
#| label: fig-0160
#| fig-cap: TFR by district, year, and origin
#| fig-width: 20
#| fig-height: 20

(fer$fer_ex_past |> 
  bind_rows(fer$fer_ex) |>
  group_by(district, year, origin) |>
  summarize(
    TFR = sum_NA(fer / 100),
    .groups = "drop"
  ) |> 
  ggplot(aes(x = year, y = TFR, color = origin)) +
		  geom_line() +
		  scale_color_manual(values = params$col_o) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "TFR", color = "") +
		  params$neutral) |> 
ggplotly(height = 2000)

```








## Origin

### Past origin change

```{r}
#| label: fig-0170
#| fig-cap: origin change by year, origin

(boc$cha |> 
  group_by(year, origin) |>
  summarize(
    change = sum(change),
    total = sum(total),
    .groups = "drop"
  ) |>
  mutate(cha_yo = if_else(total == 0, NA_real_, round(change / total * 100, params$round_rate))) |> 
  ggplot(aes(x = year, y = cha_yo, color = origin)) +
	  geom_line() +
	  scale_color_manual(values = params$col_o) +
	  labs(x = "", y = "origin change in %", color = "") +
  params$neutral) |> 
ggplotly() 
    
```

```{r}
#| label: fig-0171
#| fig-cap: origin change by year, age class (3), origin
#| fig-width: 12

(boc$cha |> 
  left_join(params$look_a1, by = "age") |>
  group_by(year, age_1, origin) |>
  summarize(
    change = sum(change),
    total = sum(total),
    .groups = "drop"
  ) |>
  mutate(cha_ya1o = if_else(total == 0, NA_real_, round(change / total * 100, 
                                                        params$round_rate))) |>
  ggplot(aes(x = year, y = cha_ya1o, color = origin)) +
	  geom_line() +
    facet_wrap(~age_1, nrow = 1) +
	  scale_color_manual(values = params$col_o) +
	  labs(x = "", y = "origin change in %", color = "") +
  params$neutral) |> 
ggplotly() 

```

```{r}
#| label: fig-0172
#| fig-cap: origin change by year, age class (5), origin
#| fig-width: 12

(boc$cha |> 
  left_join(params$look_a2, by = "age") |>
  group_by(year, age_2, origin) |>
  summarize(
    change = sum(change),
    total = sum(total),
    .groups = "drop"
  ) |>
  mutate(cha_ya2o = if_else(total == 0, NA_real_, round(change / total * 100, 
                                                        params$round_rate))) |>
  ggplot(aes(x = year, y = cha_ya2o, color = origin)) +
	  geom_line() +
    facet_wrap(~age_2, nrow = 1) +
	  scale_color_manual(values = params$col_o) +
	  labs(x = "", y = "origin change in %", color = "") +
  params$neutral) |> 
ggplotly() 

```

```{r}
#| label: fig-0173
#| fig-cap: origin change by district, year, origin
#| fig-width: 20
#| fig-height: 20

(boc$cha |> 
  group_by(district, year, origin) |>
  summarize(
    change = sum(change),
    total = sum(total),
    .groups = "drop"
  ) |>
  mutate(cha_dyo = if_else(total == 0, NA_real_, round(change / total * 100, 
                                                        params$round_rate))) |> 
  ggplot(aes(x = year, y = cha_dyo, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +
  facet_wrap(~district, ncol = 4) +
  labs(x = "", y = "origin change in %", color = "") +
  params$neutral) |> 
ggplotly(height = 2000)

```


### Prediction of origin change

```{r}
#| label: fig-0175
#| fig-cap: predicted origin change by year and origin
#| fig-width: 20
#| fig-height: 20

(ggplot(boc$cha_pred, aes(x = year, y = cha_all, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +
  facet_wrap(~district, ncol = 4) +
  labs(x = "", y = "origin change in %", color = "") +
  params$neutral) |> 
ggplotly(height = 2000)

```

## Sex ratio

### Past proportion

```{r}
#| label: fig-0190
#| fig-cap: past proportion of male babies by year

(ggplot() +
  geom_line(data = bsr$pro_male, aes(x = year, y = pro_male), color = params$col_1) +  
  geom_hline(yintercept = 50, color = params$col_grey, linetype = "dashed") +  
  scale_y_continuous(limits = c(0, 70)) +
  labs(x = "", y = "proportion male in %") +
  params$neutral) |> 
ggplotly()

```

### Prediction of proportion

```{r, eval}
#| label: fig-0191
#| fig-cap: predicted proportion of male babies by year


p <- ggplot(bsr$pro_male_pred) +
  geom_hline(yintercept = 50, color = params$col_grey, linetype = "dashed") +
  geom_line(data = bsr$pro_male_pred, aes(x = year, y = pro_male), color = params$col_1) +    
  geom_point(data = bsr$pro_male, aes(x = year, y = pro_male), color = params$col_1) +
  scale_y_continuous(limits = c(0, 70)) +
  labs(x = "", y = "proportion male in %") +
  params$neutral

ggplotly(p)

```
