
# Death

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the mortality function
source(file.path(root, "R/0201_mortality_rate_fun.R"))

# execute the mortality functions
mor <- mortality_rate_fun(params)

```

## Mortality

### age plots

#### focus: difference in mortality by sex 

For the year `r params$dea_fso_date_end`.

```{r}
#| label: fig-0200
#| fig-cap: "mortality by age, year, sex, region (focus sex)"
#| fig-width: 10
#| fig-height: 8

(mor$mor_yasr |> 
  filter(year == params$dea_fso_date_end) |> 
  mutate(rate = if_else(mor_yasr == 0, 0.001, mor_yasr)) |> 
  ggplot(aes(x = age, y = rate, color = sex)) +
    geom_line() +
    scale_color_manual(values = params$col_s) +
    scale_y_log10() +
    facet_wrap(~region, nrow = 1) +
    labs(x = "age", y = "mortality rate (in % per year)", color = "") +
    params$neutral) |> 
ggplotly()

```

#### focus: difference in mortality by region

For the year `r params$dea_fso_date_end`.

```{r}
#| label: fig-0201
#| fig-cap: "mortality by age, year, region, sex (focus: region)"
#| fig-width: 10
#| fig-height: 8

(mor$mor_yasr |> 
  filter(year == params$dea_fso_date_end) |> 
  mutate(rate = if_else(mor_yasr == 0, 0.001, mor_yasr)) |> 
  ggplot(aes(x = age, y = rate, color = region)) +
    geom_line() +
    scale_color_manual(values = params$col_r) +
    scale_y_log10() +
    facet_wrap(~sex, nrow = 1) +
    labs(x = "age", y = "mortality rate (in % per year)", color = "") +
    params$neutral) |> 
ggplotly()

```

### year plots

#### focus: difference in mortality by sex

```{r}
#| label: fig-0202
#| fig-cap: "mortality by year, age, sex, region"
#| fig-width: 14
#| fig-height: 12

age_select <- c(0, 20, 40, 50, 60, 70, 75, 80, 85)

(mor$mor_yasr |> 
  filter(age %in% age_select) |> 
  mutate(rate = if_else(mor_yasr == 0, 0.001, mor_yasr)) |> 
  ggplot(aes(x = year, y = rate, color = sex)) +
    geom_line() +
    scale_color_manual(values = params$col_s) +
    scale_y_log10() +
    facet_grid(region ~  age, 
               labeller = labeller(age = function(x) paste("age", x))) +
    labs(x = "", y = "mortality rate (in % per year)", color = "") +
    params$neutral + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |> 
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))


```

#### focus: difference in mortality by region

```{r}
#| label: fig-0203
#| fig-cap: "mortality by year, age, region, sex"
#| fig-width: 15
#| fig-height: 12

(mor$mor_yasr |> 
  filter(age %in% age_select) |> 
  mutate(rate = if_else(mor_yasr == 0, 0.001, mor_yasr)) |> 
  ggplot(aes(x = year, y = rate, color = region)) +
    geom_line() +
    scale_color_manual(values = params$col_r) +
    scale_y_log10() +
    facet_grid(sex ~  age, 
               labeller = labeller(age = function(x) paste("age", x))) +
    labs(x = "", y = "mortality rate (in % per year)", color = "") +
    params$neutral + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |> 
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```

### life expectancy calculation

For the year `r params$dea_fso_date_end`.

```{r}
#| label: fig-0204a
#| fig-cap: "cumulative survival probability (last year)"
#| fig-width: 8
#| fig-height: 6

(mor$le_mult |> 
  filter(year == params$dea_fso_date_end) |> 
  ggplot(aes(x = age, y = mult, color = sex)) +
  geom_line() +
  scale_color_manual(values = params$col_s) +  
  labs(x = "age", y = "cumulative survival probability", color = "") +
  params$neutral) |> 
 ggplotly()

```



### method comparison (based on rate vs. death/birth/population)

method differences (rate vs. deaths/births/population) details: see life expectancy function definition

```{r}
#| label: fig-0204b
#| fig-cap: "life expectancy at birth by year: method comparison (Zurich only)"
#| fig-width: 10
#| fig-height: 8

(mor$le_ysr |> 
  filter(region == params$uni_r[1]) |> 
  right_join(mor$le, by = c("year", "sex")) |> 
  rename(basic = life, rate = le_ysr) |>
  pivot_longer(c(rate, basic), names_to = "method", values_to = "life") |> 
  ggplot(aes(x = year, y = life, color = method)) +
    geom_line() +
    facet_wrap(~sex, nrow = 1) +
    scale_color_manual(values = params$col_6[1:2]) +
    labs(x = "", y = "life expectancy at birth", color = "") +
    params$neutral) |> 
ggplotly()  
    
    
```

### life expectancy

#### focus: differences by sex

```{r}
#| label: fig-0205
#| fig-cap: "life expecancy at birth by year, sex and region"
#| fig-width: 10
#| fig-height: 8

(mor$le_ysr |> 
  filter(!is.na(le_ysr)) |> 
  ggplot(aes(x = year, y = le_ysr, color = sex)) +
    geom_line() +
    facet_wrap(~region, nrow = 1) +
    scale_color_manual(values = params$col_s) +
    labs(x = "", y = "life expectancy at birth", color = "") +
    params$neutral) |> 
ggplotly()  

```

#### focus: differences by region

```{r}
#| label: fig-0206
#| fig-cap: "life expecancy at birth by year, region and sex"
#| fig-width: 10
#| fig-height: 8

(mor$le_ysr |> 
  filter(!is.na(le_ysr)) |> 
  ggplot(aes(x = year, y = le_ysr, color = region)) +
    geom_line() +
    facet_wrap(~sex, nrow = 1) +
    scale_color_manual(values = params$col_r) +
    labs(x = "", y = "life expectancy at birth", color = "") +
    params$neutral) |> 
ggplotly()  

```

### mortality over base years (including tail correction)

```{r}
#| label: fig-0207
#| fig-cap: "mortality by age, sex, region (tails corrected)"
#| fig-width: 10
#| fig-height: 10

(mor$mor_asr |> 
  ggplot(aes(x = age, y = mor_asr, color = region)) +
    geom_line() +
    scale_y_log10() +
    facet_wrap(~sex, ncol = 1) +
    scale_color_manual(values = params$col_r) +
    labs(x = "", y = "mortality rate (in % per year)", color = "") +
    params$neutral) |> 
ggplotly()

```

### smoothing with LOESS

```{r}
#| label: fig-0208
#| fig-cap: "mortality by age, sex and region; fitted"
#| fig-width: 10
#| fig-height: 10

fit_lev <- c("initial", "smoothed")

(mor$mor_fit |> 
  pivot_longer(c(mor_asr, mor_fit), names_to = "category", values_to = "mor") |>
  mutate(cat = fact_if(category, fit_lev, "mor_asr")) |> 
  ggplot(aes(x = age, y = mor, color = region, linetype = cat)) +
    geom_line() +
    scale_y_log10() +
    facet_wrap(~sex, ncol = 1) +
    scale_color_manual(values = params$col_r) +
      scale_linetype_manual(values = c("solid", "dotted")) +  
    labs(x = "", y = "mortality rate (in % per year)", 
         color = "", linetype = "") +
    params$neutral) |> 
ggplotly()

```

### ratio Zurich / Switzerland

```{r}
#| label: fig-0209
#| fig-cap: "ratio Zurich/Switzerland by age and sex"
#| fig-width: 10
#| fig-height: 8

(mor$ratio_as |> 
  ggplot(aes(x = age, y = ratio, color = sex)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = params$col_grey) + 
    geom_line() +
    scale_color_manual(values = params$col_s) +
    scale_y_continuous(limits = c(0, NA)) + 
    labs(x = "age", y = "mortality ratio (Zurich / Switzerland)", color = "") +
    params$neutral) |> 
ggplotly()

```

### Zurich: future mortality rate

```{r}
#| label: fig-0210
#| fig-cap: mortality prediction by year, age and sex
#| fig-width: 8
#| fig-height: 10

n_years <- length(params$scen_begin:params$scen_end)
pal_year <- colorRampPalette(params$col_6[5:1])(n_years)

(mor$mor_zh_yas_past_future |> 
  filter(year >= params$scen_begin) |> 
  mutate(rate = if_else(mor_yas == 0, 0.001, mor_yas)) |>   
    ggplot(aes(x = age, y = rate, color = as.factor(year))) +
    geom_line() +
    scale_y_log10() +
    facet_wrap(~sex, ncol = 1) +
    scale_color_manual(values = pal_year) +
    labs(x = "age", y = "mortality rate (in % per year)", color = "") +
    params$neutral) |> 
 ggplotly() 

```

### Zurich and Switzerland: future mortality rate


```{r}
#| label: fig-0210a
#| fig-cap: mortality by year, age, sex, region
#| fig-width: 15
#| fig-height: 12

age_select <- c(0, 20, 40, 50, 60, 70, 75, 80, 85, 90)

(mor$mor_zh_ch_yas_past_future |> 
  filter(age %in% age_select) |> 
  mutate(rate = if_else(mor_yasr == 0, 0.001, mor_yasr)) |> 
  ggplot(aes(x = year, y = rate, color = sex)) +
    geom_line() +
    scale_color_manual(values = params$col_s) +
    scale_y_log10() +
    facet_grid(region ~  age, 
               labeller = labeller(age = function(x) paste("age", x))) +
    labs(x = "", y = "mortality rate (in % per year)", color = "") +
    params$neutral + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |> 
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))


```




### Zurich: life expectancy (including the model data)

#### focus: differences by sex

```{r}
#| label: fig-0211
#| fig-cap: "life expectancy at birth by year, sex, region; with model"
#| fig-width: 10
#| fig-height: 8
#| warning: false
#| message: false

# why? visual check, if life expectancy based on predicted rates are ok

# ZH: life expectancy
le_ys_ZH <- life_exp(
  data = mor$mor_zh_yas_past_future, mor = "mor_yas",
  age = "age", group_cols = c("year", "sex"),
  age_max = params$dea_age_max_le, qx_NA = params$dea_qx_NA_le,
  age_at = params$dea_age_at, radix = params$dea_radix
) |> 
  mutate(region = factor(params$text_r[1], params$uni_r)) |>
  rename(le_ysr = life_exp) |>
  select(year, sex, region, le_ysr)

# plot data (also used in the next chunk)
le_ysr_model_dat <- mor$le_ysr |> 
  filter(region != params$uni_r[1]) |>
  bind_rows(le_ys_ZH) |>
  arrange(year, sex, region) |> 
  filter(!is.na(le_ysr))
    
# plot
(le_ysr_model_dat |>
ggplot(aes(x = year, y = le_ysr, color = sex)) +
    geom_line() +
    facet_wrap(~region, nrow = 1) +
    scale_color_manual(values = params$col_s) +
    labs(x = "", y = "life expectancy at birth", color = "") +
    params$neutral) |>
ggplotly()


```

#### focus: differences by region

```{r}
#| label: fig-0212
#| fig-cap: "life expectancy at birth by year, region, sex; with model"
#| fig-width: 10
#| fig-height: 8

(le_ysr_model_dat |> 
  ggplot(aes(x = year, y = le_ysr, color = region)) +
    geom_line() +
    facet_wrap(~sex, nrow = 1) +
    scale_color_manual(values = params$col_r) +
    labs(x = "", y = "life expectancy at birth", color = "") +
    params$neutral) |> 
ggplotly()  

```


### Zurich: life expectancy at different age

```{r}
#| label: fig-0213
#| fig-cap: "life expectancy at different age (0 and 65), by year, region, sex"
#| fig-width: 10
#| fig-height: 12
#| warning: false
#| message: false

# mortality preparation (CH)
mor_prep_ch <- mor$mor_yasr |> 
  filter(region == params$uni_r[2])

# mortality preparation (ZH/CH, past/future)
mor_prep <- mor$mor_zh_yas_past_future |>
  mutate(region = factor(params$text_r[1], params$uni_r)) |>
  rename(mor_yasr = mor_yas) |>
  select(year, age, sex, region, mor_yasr) |>
  bind_rows(mor_prep_ch)

# life expectancy at different age (0 and 65)
le_ysr_a0 <- life_exp(
  data = mor_prep, mor = "mor_yasr",
  age = "age", group_cols = c("year", "sex", "region"),
  age_max = params$dea_age_max_le, qx_NA = params$dea_qx_NA_le,
  age_at = 0, radix = params$dea_radix
) |>
  mutate(age = 0)

le_ysr_a65 <- life_exp(
  data = mor_prep, mor = "mor_yasr",
  age = "age", group_cols = c("year", "sex", "region"),
  age_max = params$dea_age_max_le, qx_NA = params$dea_qx_NA_le,
  age_at = 65, radix = params$dea_radix
) |>
  mutate(age = 65)

# life expectancy for all combinations (ZH/CH, past/future, sex, region)

# correction for Switzerland: latest values not available yet
# the same correction as before
le_yasr <- le_ysr_a0 |>
  bind_rows(le_ysr_a65) |>
  mutate(le_yasr = if_else((region == params$uni_r[2]) &
           !(year %in% c(params$dea_fso_date_start:params$dea_fso_date_end,
                         params$scen_begin:params$scen_end)),
      NA_real_, life_exp))


(le_yasr |> 
  ggplot(aes(x = year, y = le_yasr, color = as.factor(age))) +
    geom_line() +
    facet_grid(region ~ sex) +
    scale_color_manual(values = params$col_6[1:2]) +
    labs(x = "", y = "life expectancy", color = "age") +
    params$neutral) |> 
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```


