
# Immigration

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the migration functions
source(file.path(root, "R/0301_mig_rate_dy_fun.R"))
source(file.path(root, "R/0302_mig_prop_so_dy_fun.R"))
source(file.path(root, "R/0303_mig_prop_a_dyso_fun.R"))

# execute the migration functions (mig_prop_a_dyso_fun: approx 30 seconds)
ims_rate_dy <- mig_rate_dy_fun(process = "immigration", params = params)
ims_prop_so <- mig_prop_so_dy_fun(process = "immigration", params = params)
ims_prop_a <- mig_prop_a_dyso_fun(process = "immigration", params = params)

```

## immigration\* rate (per district and year)

immigration* rate by district and year

```{r}
#| label: fig-0300
#| fig-cap: immigration* rate by district and year
#| fig-width: 20
#| fig-height: 20

(ggplot(ims_rate_dy$mis_rate_dy) +
		  geom_line(aes(x = year, y = mis_rate_dy), color = params$col_1) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "immigration* rate (in % per year)") +
		  params$neutral) |> 
ggplotly(height = 2000)

	
```


## prediction: future immigration\* rate (per year and district)

### overview

immigration* rate by district and year (past and prediction)

```{r}
#| label: fig-0301
#| fig-cap: immigration* rate by district and year (past and prediction)
#| fig-height: 10

time_lev <- c("past", "future")

(ims_rate_dy$mis_past_pred |> 
    select(district, year, rate_all) |> 
    mutate(time = factor(if_else(year <= params$ims_base_end,
      time_lev[1], time_lev[2]
    ), levels = time_lev)) |> 
 ggplot() +
		  geom_line(aes(x = year, y = rate_all, linetype = time), 
		            color = params$col_1) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "immigration* rate (in % per year)", linetype = "") +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral ) |> 
ggplotly(height = 2000) 
  
```


## migration: distribution of sex and origin

immigration proportion of sex by origin, district and year

```{r}
#| label: fig-0303
#| fig-cap: immigration proportion of sex by origin, district and year
#| fig-width: 20
#| fig-height: 20

 (ggplot(ims_prop_so$mis_dyso) +
		  geom_line(aes(x = year, y = mis_prop_dyso, 
		                color = sex, linetype = origin)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "proportion in % (per district and year, in immigration*", 
		       color = "", linetype = "") +
      scale_color_manual(values = params$col_s) +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly(height = 2000)
  
```

## distribution of sex and origin: prediction

immigration proportion of sex by origin, district and year (past and prediction)

```{r}
#| label: fig-0304
#| fig-cap: immigration proportion of sex by origin, district and year (past and prediction) 
#| fig-width: 20
#| fig-height: 20

(ggplot(ims_prop_so$mis_so_past_pred) +
		  geom_line(aes(x = year, y = prop_all, 
		                color = sex, linetype = origin)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "proportion in % (per district and year, in immigration*", 
		       color = "", linetype = "") +
      scale_color_manual(values = params$col_s) +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly(height = 2000)

```

## age proportion

### immigration\*

immigration* per district, year, sex, origin

```{r}
#| label: fig-0310
#| fig-cap: immigration* per district, year, sex, origin
#| fig-width: 20
#| fig-height: 20

(ggplot(ims_prop_a$mis_dyso) +
		  geom_line(aes(x = year, y = mis_dyso, 
		                color = sex, linetype = origin)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "immigration* per year", 
		       color = "", linetype = "") +
      scale_color_manual(values = params$col_s) +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly(height = 2000)
  
```
immigration* per district, year, sex, origin (free scales)

```{r}
#| label: fig-0311
#| fig-cap: immigration* per district, year, sex, origin; free scales
#| fig-width: 20
#| fig-height: 20

(ggplot(ims_prop_a$mis_dyso) +
		  geom_line(aes(x = year, y = mis_dyso, 
		                color = sex, linetype = origin)) +
		  facet_wrap(~district, ncol = 4, scales = "free_y") +
		  labs(x = "", y = "immigration* per year", 
		       color = "", linetype = "") +
      scale_color_manual(values = params$col_s) +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly(height = 2000)

```

### age proportion: per district, year, sex, origin

#### focus age

immigration* age proportion per district, year, sex (Swiss only)

```{r}
#| label: fig-0312a
#| fig-cap: immigration* age proportion per district, year, sex (Swiss only)
#| fig-width: 20
#| fig-height: 20

(ims_prop_a$mis_dyaso |> 
  filter(year == max(year), origin == params$uni_o[1]) |> 
  ggplot() +
	  geom_line(aes(x = age, y = mis_prop_a, 
	                color = sex, linetype = origin)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "age", y = "proportion in %", color = "") +
    scale_color_manual(values = params$col_s) +
	  params$neutral) |> 
ggplotly(height = 2000)


```


immigration* age proportion per district, year, sex (foreign only)

```{r}
#| label: fig-0312b
#| fig-cap: immigration* age proportion per district, year, sex (foreign only)
#| fig-width: 20
#| fig-height: 20

(ims_prop_a$mis_dyaso |> 
  filter(year == max(year), origin == params$uni_o[2]) |> 
  ggplot() +
	  geom_line(aes(x = age, y = mis_prop_a, 
	                color = sex, linetype = origin)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "age", y = "proportion in %", color = "") +
    scale_color_manual(values = params$col_s) +
	  params$neutral) |> 
ggplotly(height = 2000)

```


#### focus years

```{r}
#| label: fig-0313a
#| fig-cap: immigration* age proportion per selected district, year, sex (Swiss only, focus on years)
#| fig-width: 15
#| fig-height: 15

(ims_prop_a$mis_dyaso |> 
  filter(origin == params$uni_o[1], 
         age %in% seq(0, 60, by = 20),
         district %in% params$d_select) |> 
  ggplot() +
	  geom_line(aes(x = year, y = mis_prop_a, color = as.factor(age))) +
	  facet_grid(district ~ sex) +
	  labs(x = "year", y = "proportion in %", color = "age") +
    scale_color_manual(values = params$col_6[1:4]) +
	  params$neutral) |> 
ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```



```{r}
#| label: fig-0313b
#| fig-cap: immigration* age proportion per selected district, year, sex (foreign only, focus on years)
#| fig-width: 15
#| fig-height: 15

(ims_prop_a$mis_dyaso |> 
  filter(origin == params$uni_o[2], 
         age %in% seq(0, 60, by = 20),
         district %in% params$d_select) |> 
  ggplot() +
	  geom_line(aes(x = year, y = mis_prop_a, color = as.factor(age))) +
	  facet_grid(district ~ sex) +
	  labs(x = "year", y = "proportion in %", color = "age") +
    scale_color_manual(values = params$col_6[1:4]) +
	  params$neutral) |> 
ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```

## smoothing immigration\* with LOESS over years (by district, age, sex)

#### focus age

```{r}
#| label: fig-0314a
#| fig-cap: immigration* smoothed over year (Swiss only, last year, focus on age)
#| fig-width: 15
#| fig-height: 15

fit_lev <- c("initial", "smoothed")

(ims_prop_a$mis_smooth |> 
  filter(origin == params$uni_o[1], 
         year == max(year),
         district %in% params$d_select) |> 
  pivot_longer(c(mis_dyaso, mis_smooth), 
               names_to = "category", values_to = "mis") |>  
  mutate(cat = fact_if(category, fit_lev, "mis_dyaso")) |> 
  ggplot() +
	  geom_line(aes(x = age, y = mis, color = cat)) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "immigration* per year", color = "") +
    scale_color_manual(values = params$col_6[1:2]) +
    params$neutral) |> 
ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```


```{r}
#| label: fig-0314b
#| fig-cap: immigration* smoothed over year (foreign only, last year, focus on age)
#| fig-width: 15
#| fig-height: 15

(ims_prop_a$mis_smooth |> 
  filter(origin == params$uni_o[2], 
         year == max(year),
         district %in% params$d_select) |> 
  pivot_longer(c(mis_dyaso, mis_smooth), 
               names_to = "category", values_to = "mis") |>  
  mutate(cat = fact_if(category, fit_lev, "mis_dyaso")) |> 
  ggplot() +
	  geom_line(aes(x = age, y = mis, color = cat)) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "immigration* per year", color = "") +
    scale_color_manual(values = params$col_6[1:2]) +
    params$neutral) |> 
ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```

#### focus years

```{r}
#| label: fig-0315a
#| fig-cap: immigration* smoothed over year (Swiss only, focus on years)
#| fig-width: 15
#| fig-height: 20

(ims_prop_a$mis_smooth |> 
  filter(origin == params$uni_o[1],
         age %in% seq(0, 60, by = 20),
         district %in% params$d_select) |> 
  pivot_longer(c(mis_dyaso, mis_smooth), 
               names_to = "category", values_to = "mis") |>  
  mutate(cat = fact_if(category, fit_lev, "mis_dyaso")) |> 
  ggplot() +
	  geom_line(aes(x = year, y = mis, color = as.factor(age), linetype = cat)) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "immigration* per year", color = "age", linetype = "") +
    scale_color_manual(values = params$col_6[1:4]) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    # guides(color = "none") +
    params$neutral) |> 
  ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))


```


```{r}
#| label: fig-0315b
#| fig-cap: immigration* smoothed over year (foreign only, focus on years)
#| fig-width: 15
#| fig-height: 20

(ims_prop_a$mis_smooth |> 
  filter(origin == params$uni_o[2],
         age %in% seq(0, 60, by = 20),
         district %in% params$d_select) |> 
  pivot_longer(c(mis_dyaso, mis_smooth), 
               names_to = "category", values_to = "mis") |>  
  mutate(cat = fact_if(category, fit_lev, "mis_dyaso")) |> 
  ggplot() +
	  geom_line(aes(x = year, y = mis, color = as.factor(age), linetype = cat)) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "immigration* per year", color = "age", linetype = "") +
    scale_color_manual(values = params$col_6[1:4]) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    # guides(color = "none") +
    params$neutral) |> 
  ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))


```

## smoothing with LOESS, proportion by age 

### focus age

```{r}
#| label: fig-0318a
#| fig-cap: immigration* age proportion smoothed (Swiss only, focus on age)
#| fig-width: 15
#| fig-height: 20
  
(ims_prop_a$prop_fit |> 
  filter(origin == params$uni_o[1], 
         year == max(year),
         district %in% params$d_select) |> 
    pivot_longer(c(prop_a_smooth, prop_fit), 
                 names_to = "category", values_to = "prop") |>
    mutate(cat = fact_if(category, fit_lev, "prop_a_smooth")) |> 
  ggplot() +
	  geom_line(aes(x = age, y = prop, color = cat)) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "proportion in %", color = "") +
    scale_color_manual(values = params$col_6[1:2]) +
    params$neutral) |>
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```

```{r}
#| label: fig-0318b
#| fig-cap: immigration* age proportion smoothed (foreign only, focus on age)
#| fig-width: 15
#| fig-height: 20
  
(ims_prop_a$prop_fit |> 
  filter(origin == params$uni_o[2], 
         year == max(year),
         district %in% params$d_select) |> 
    pivot_longer(c(prop_a_smooth, prop_fit), 
                 names_to = "category", values_to = "prop") |>
    mutate(cat = fact_if(category, fit_lev, "prop_a_smooth")) |> 
  ggplot() +
	  geom_line(aes(x = age, y = prop, color = cat)) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "proportion in %", color = "") +
    scale_color_manual(values = params$col_6[1:2]) +
    params$neutral) |> 
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1)) 

```


## regression

```{r}
#| label: fig-0319a
#| fig-cap: immigration* age proportion by district, year, sex (Swiss only, future)
#| fig-width: 15
#| fig-height: 20

n_years <- length(params$scen_begin:params$scen_end)
pal_year <- colorRampPalette(params$col_6[5:1])(n_years)

(ims_prop_a$mis_a_past_pred |> 
  filter(origin == params$uni_o[1],
         year >= params$scen_begin,
         district %in% params$d_select) |> 
  ggplot() +
	  geom_line(aes(x = age, y = prop_a, color = as.factor(year))) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "proportion in %", color = "") +
    scale_color_manual(values = rev(pal_year)) +
    params$neutral) |> 
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```


```{r}
#| label: fig-0319b
#| fig-cap: immigration* age proportion by district, year, sex (foreign only, future)
#| fig-width: 15
#| fig-height: 20

n_years <- length(params$scen_begin:params$scen_end)
pal_year <- colorRampPalette(params$col_6[5:1])(n_years)

(ims_prop_a$mis_a_past_pred |> 
  filter(origin == params$uni_o[2],
         year >= params$scen_begin,
         district %in% params$d_select) |> 
  ggplot() +
	  geom_line(aes(x = age, y = prop_a, color = as.factor(year))) +
	  facet_grid(district ~ sex) +
	  labs(x = "age", y = "proportion in %", color = "") +
    scale_color_manual(values = rev(pal_year)) +
    params$neutral) |> 
 ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```

