# Relocation-immigration

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the mortality function
source(file.path(root, "R/0501_relocation_fun.R"))

# execute the relocation function
rel_imm <- relocation_fun(process = "relocation-immigration", params = params)

```


## Past proportion of relocation on immigration*

### Differences by sex

Why this check? Can the variable sex be omitted in the prediction?

Proportion of relocation on immigration* by district, age, sex (Swiss only)

```{r}
#| label: fig-0500a
#| fig-cap: proportion of relocation by district, age, sex (Swiss only)
#| fig-width: 20
#| fig-height: 20

rel_prop_daso <- rel_imm$mis_rel |> 
  group_by(district, age, sex, origin) |> 
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) |>
    mutate(rel_prop_daso = if_else(mis == 0, NA_real_, 
                                   round(rel / mis * 100, params$round_prop)))

(rel_prop_daso |> 
  filter(origin == params$uni_o[1], !is.na(rel_prop_daso)) |> 
  ggplot() +
		  geom_line(aes(x = age, y = rel_prop_daso,
		                color = sex)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "proportion in % (relocation on immigration*)",
		       color = "") +
      scale_color_manual(values = params$col_s) +
		  params$neutral) |> 
ggplotly(height = 1500)

```

Proportion of relocation on immigration* by district, age, sex (foreign only)

```{r}
#| label: fig-0500b
#| fig-cap: proportion of relocation by district, age, sex (foreign only)
#| fig-width: 20
#| fig-height: 20

(rel_prop_daso |> 
  filter(origin == params$uni_o[2], !is.na(rel_prop_daso)) |> 
  ggplot() +
		  geom_line(aes(x = age, y = rel_prop_daso,
		                color = sex)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "proportion in % (relocation on immigration*)",
		       color = "") +
      scale_color_manual(values = params$col_s) +
		  params$neutral) |> 
ggplotly(height = 1500)

```


### Differences by origin

Why this check?
First, to emphasize the need of origin in the model.
Second, the plots can be used to get an idea of the value of the age threshold parameter.

```{r}
#| label: fig-0501
#| fig-cap: proportion of relocation by district, age and origin
#| fig-width: 20
#| fig-height: 20

(rel_imm$mis_rel |> 
    group_by(district, age, origin) |>
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) |>
    mutate(rel_prop_dao = if_else(mis == 0, NA_real_, 
                                  round(rel / mis * 100, 
                                        params$round_prop))) |> 
  filter(!is.na(rel_prop_dao)) |> 
  ggplot() +
		  geom_line(aes(x = age, y = rel_prop_dao, color = origin)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "proportion in % (relocation on immigration*)",
		       color = "") +
      scale_color_manual(values = params$col_o) +
		  params$neutral) |> 
ggplotly(height = 1500)

```

### Which base years?

#### Year, origin

Why? Several plots to determine the base years.

```{r}
#| label: fig-0502
#| fig-cap: proportion by year and origin
#| fig-width: 15
#| fig-height: 12

(rel_imm$mis_rel |> 
  group_by(year, origin) |>
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) |>
    mutate(rel_prop_yo = if_else(mis == 0, NA_real_, 
                                 round(rel / mis * 100, 
                                       params$round_prop))) |> 
ggplot(aes(x = year, y = rel_prop_yo, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +
  labs(x = "", y = "proportion in % (relocation on immigration*)", 
       color = "") +
  scale_y_continuous(limits = c(0, NA)) +
  params$neutral) |> 
ggplotly()
  
```

#### Year, age, origin

```{r}
#| label: fig-0503
#| fig-cap: propportion by year, age and origin
#| fig-width: 20
#| fig-height: 20

age_plot <- seq(0, 80, by = 10)

(rel_imm$mis_rel |> 
  group_by(, year, age, origin) |>
    summarize(
      mis = sum(mis),
      rel = sum(rel),
      .groups = "drop"
    ) |>
    rename(age_num = age) |>
    mutate(
      rel_prop_yao = if_else(mis == 0, NA_real_, 
                             round(rel / mis * 100, 
                                   params$round_prop)),
      age = as.factor(age_num)
    ) |> 
  filter(age_num %in% age_plot) |> 
ggplot(aes(x = year, y = rel_prop_yao, color = origin)) +
  geom_line() +
  scale_color_manual(values = params$col_o) +
  facet_wrap(~age, ncol = 2, 
               labeller = labeller(age = function(x) paste("age", x))) +  
  labs(x = "", y = "proportion in % (relocation on immigration*)", 
       color = "") +
  scale_y_continuous(limits = c(0, NA)) +
  params$neutral) |> 
ggplotly(height = 1200)
  

```

### Aggregate over sex

This plot is the starting point (before smoothing, age corrections).

#### Focus on age

Processes by district, age, year `r max(rel_imm$rem_dyao$year)`, Swiss only.

```{r}
#| label: fig-0504a
#| fig-cap: processes by district, age, last year, Swiss only
#| fig-width: 20
#| fig-height: 20

process_lev <- c("relocation", "immigration*")

process_plot <- rel_imm$rem_dyao |> 
    gather(`rel`, `mis`, key = category, value = count) |>
    mutate(cat = fact_if(category, process_lev, "rel")) |>
    select(district, year, age, origin, cat, count)

(process_plot |>
  filter(origin == params$uni_o[1], year == max(year)) |>
  ggplot() +
		  geom_line(aes(x = age, y = count, color = cat)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "quantity per year", color = "") +
      scale_color_manual(values = params$col_6[1:2]) +
		  params$neutral) |> 
ggplotly(height = 1500)

```


Processes by district, age, year `r max(rel_imm$rem_dyao$year)`, foreign only.

```{r}
#| label: fig-0504b
#| fig-cap: processes by district, age, last year, foreign only
#| fig-width: 20
#| fig-height: 20

(process_plot |>
  filter(origin == params$uni_o[2], year == max(year)) |>
  ggplot() +
		  geom_line(aes(x = age, y = count, color = cat)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "", y = "quantity per year", color = "") +
      scale_color_manual(values = params$col_6[1:2]) +
		  params$neutral) |> 
ggplotly(height = 1500)

```

#### Focus on years

Processes by district (selected), year, age (selected), Swiss only

```{r}
#| label: fig-0505a
#| fig-cap: processes by district (selected), year, age (selected), Swiss only
#| fig-width: 20
#| fig-height: 20

age_plot <- seq(0, 80, by = 20)

process_plot_age <- process_plot |> 
    rename(age_num = age) |>
    mutate(age = as.factor(age_num)) |> 
    filter(district %in% params$d_select, 
           age_num %in% age_plot)

  
(process_plot_age |> 
  filter(origin == params$uni_o[1]) |>
  ggplot() +
		  geom_line(aes(x = year, y = count, color = cat)) +
		  facet_grid(age ~ district, 
               labeller = labeller(age = function(x) paste("age", x))) +   
		  labs(x = "", y = "quantity per year", color = "") +
      scale_color_manual(values = params$col_6[1:2]) +
		  params$neutral) |> 
ggplotly(height = 1500) |> 
  layout(legend = list(x = 1.05, y = 1))
  
```

Processes by district (selected), year, age (selected), foreign only

```{r}
#| label: fig-0505b
#| fig-cap: processes by district (selected), year, age (selected), foreign only
#| fig-width: 20
#| fig-height: 20

(process_plot_age |> 
  filter(origin == params$uni_o[2]) |>
  ggplot() +
		  geom_line(aes(x = year, y = count, color = cat)) +
		  facet_grid(age ~ district, 
               labeller = labeller(age = function(x) paste("age", x))) +   
		  labs(x = "", y = "quantity per year", color = "") +
      scale_color_manual(values = params$col_6[1:2]) +
		  params$neutral) |> 
ggplotly(height = 1500) |> 
  layout(legend = list(x = 1.05, y = 1))

```



### Relocation and immigration* after smoothing: dyao

Processes by district (selected), age, year `r params$date_end`, origin.

```{r}
#| label: fig-0506
#| fig-cap: processes by district, age, last year, origin
#| fig-width: 12
#| fig-height: 12

smooth_lev <- c("initial", "smoothed")

temp_initial <- rel_imm$dyao_smooth |> 
  gather(`rel`, `mis`, key = category, value = count) |>
  mutate(
    cat = fact_if(category, process_lev, "rel"),
    smooth = factor(smooth_lev[1], levels = smooth_lev)
  ) |>
  select(district, year, age, origin, cat, smooth, count)

temp_smooth <- rel_imm$dyao_smooth |> 
  gather(`rel_a`, `mis_a`, key = category, value = count) |>
  mutate(
    cat = fact_if(category, process_lev, "rel_a"),
    smooth = factor(smooth_lev[2], levels = smooth_lev)
  ) |>
  select(district, year, age, origin, cat, smooth, count)

(bind_rows(temp_initial, temp_smooth) |> 
  filter(district %in% params$d_select, 
         year == max(year)) |> 
  ggplot() +
		  geom_line(aes(x = age, y = count, color = cat, linetype = smooth)) +
		  facet_grid(district ~ origin) +
		  labs(x = "", y = "quantity per year", color = "", linetype = "") +
      scale_color_manual(values = params$col_6[1:2]) +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly() |> 
    layout(legend = list(x = 1.05, y = 1))
  

```

### Proportion of relocation on immigration* after smoothing: dyao 

Proportion by district, age, year `r params$date_end` and origin.

```{r}
#| label: fig-0508
#| fig-cap: proportion smooothed by district, age, last year and origin
#| fig-width: 20
#| fig-height: 20

(rel_imm$prop_dyao_smooth |> 
  filter(year == max(year)) |> 
  ggplot() +
		  geom_line(aes(x = age, y = rel_prop_dyao, color = origin)) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "age", y = "proportion in % (relocation on immigration*)",
		       color = "") +
      scale_color_manual(values = params$col_o) +
		  params$neutral) |> 
ggplotly(height = 1500)


```

### Smoothing the proportion

```{r}
#| label: fig-0510
#| fig-cap: processes by district (selected), age and origin (smoothed)
#| fig-width: 12
#| fig-height: 15

  temp_initial <- rel_imm$dao_smooth |> 
    gather(`rel`, `mis`, key = category, value = count) |>
    mutate(
      cat = fact_if(category, process_lev, "rel"),
      smooth = factor(smooth_lev[1], levels = smooth_lev)
    ) |>
    select(district, age, origin, cat, smooth, count)

  temp_smooth <- rel_imm$dao_smooth |> 
    gather(`rel_a`, `mis_a`, key = category, value = count) |>
    mutate(
      cat = fact_if(category, process_lev, "rel_a"),
      smooth = factor(smooth_lev[2], levels = smooth_lev)
    ) |>
    select(district, age, origin, cat, smooth, count)

(bind_rows(temp_initial, temp_smooth) |> 
  filter(district %in% params$d_select) |> 
  ggplot() +
		  geom_line(aes(x = age, y = count, color = cat, linetype = smooth)) +
		  facet_grid(district ~ origin) +
		  labs(x = "", y = "quantity per year", color = "", linetype = "") +
      scale_color_manual(values = params$col_6[1:2]) +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))

```

### Proportions: different aggregation levels

```{r}
#| label: fig-0512
#| fig-cap: !expr paste0("proportions from different aggregation levels (", params$date_end, ")")
#| fig-width: 15
#| fig-height: 15

(rel_imm$prop_agg |> 
  select(district, year, age, origin, rel_prop_dyao, rel_prop_dao) |>
  rename(dyao = rel_prop_dyao, dao = rel_prop_dao) |>
  gather(`dyao`, `dao`, key = category, value = count) |>
  mutate(cat = factor(category, levels = rel_imm$sources)) |> 
  filter(district %in% params$d_select, year == max(year)) |>  
  ggplot() +
		  geom_line(aes(x = age, y = count, color = cat)) +
		  facet_grid(district ~ origin) +
		  labs(x = "", y = "proportion in % (relocation on immigration*)", 
		       color = "") +
      scale_color_manual(values = params$col_6[1:2]) +
		  params$neutral) |> 
ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))

```


### Smoothing (after different levels of aggregation were brought together)

```{r}
#| label: fig-0515
#| fig-cap: !expr paste0("one proportion after smoothing (", params$date_end, ")")
#| fig-width: 15
#| fig-height: 15

(rel_imm$prop_agg_smooth |> 
  gather(`rel_prop`, `prop_smooth`, key = category, value = count) |>
  mutate(cat = fact_if(category, smooth_lev, "rel_prop")) |>
  select(district, year, age, origin, cat, count) |> 
  filter(district %in% params$d_select, year == max(year)) |>  
  ggplot() +
		  geom_line(aes(x = age, y = count, linetype = cat), 
		            col = params$col_1) +
		  facet_grid(district ~ origin) +
		  labs(x = "age", y = "proportion in % (relocation on immigration*)",
		       linetype = "") +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))


```

## Prediction of the relocation proportion on immigration*

### Smoothing

```{r}
#| label: fig-0517
#| fig-cap: prediction by district (selected), last prediction year, age, origin
#| fig-width: 15
#| fig-height: 15

(rel_imm$pred_smooth |> 
  select(district, year, age, origin, pred_all_age, pred_smooth) |>
  gather(pred_smooth, `pred_all_age`, `pred_smooth`, key = category, value = count) |>
  mutate(cat = fact_if(category, smooth_lev, "pred_all_age")) |>
  select(district, year, age, origin, cat, count) |> 
  filter(district %in% params$d_select, year == max(year)) |>  
  ggplot() +
		  geom_line(aes(x = age, y = count, linetype = cat), 
		            col = params$col_1) +
		  facet_grid(district ~ origin) +
		  labs(x = "age", y = "proportion in % (relocation on immigration*)",
		       linetype = "") +
      scale_linetype_manual(values = c("solid", "dotted")) +
		  params$neutral) |> 
ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))


```

### Final prediction

predicted proportions by district, year, age, Swiss only

```{r}
#| label: fig-0518a
#| fig-cap: predicted proportions by district, year, age, Swiss only
#| fig-width: 20
#| fig-height: 20

n_years <- length(params$scen_begin:params$scen_end)
pal_year <- colorRampPalette(params$col_6[5:1])(n_years)

(rel_imm$pred_smooth |> 
  filter(origin == params$uni_o[1]) |> 
  ggplot() +
		  geom_line(aes(x = age, y = pred_smooth, color = as.factor(year))) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "age", y = "proportion in % (relocation on immigration*)", color = "") +
      scale_color_manual(values = rev(pal_year)) +
		  params$neutral) |> 
ggplotly(height = 1500)

```

predicted proportions by district, year, age, foreign only

```{r}
#| label: fig-0518b
#| fig-cap: predicted proportions by district, year, age, foreign only
#| fig-width: 20
#| fig-height: 20

n_years <- length(params$scen_begin:params$scen_end)
pal_year <- colorRampPalette(params$col_6[5:1])(n_years)

(rel_imm$pred_smooth |> 
  filter(origin == params$uni_o[2]) |> 
  ggplot() +
		  geom_line(aes(x = age, y = pred_smooth, color = as.factor(year))) +
		  facet_wrap(~district, ncol = 4) +
		  labs(x = "age", y = "proportion in % (relocation on immigration*)", color = "") +
      scale_color_manual(values = rev(pal_year)) +
		  params$neutral) |> 
ggplotly(height = 1500)

```

