
# Naturalization

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the naturalization function
source(file.path(root, "R/0701_naturalization_rate_fun.R"))

# execute the naturalization function
nat <- naturalization_rate_fun(params = params)

```


## naturalization vs. denaturalization

proportion of naturalization on all citizenship changes (naturalization and denaturalization).

```{r}
#| label: fig-0700
#| fig-cap: proportion of naturalization on all citizenship changes
#| fig-width: 10
#| fig-height: 8


processes <- c("nat", "denat")

(nat$nat_denat |> 
  filter(origin_prev != origin) |> 
  mutate(proc = fact_if(origin_prev, processes, params$uni_o[2])) |> 
  group_by(year, proc) |>
  summarize(
    count = sum(nat),
    .groups = "drop"
  ) |> 
  pivot_wider(names_from = proc, values_from = count) |> 
  replace_na(list(nat = 0, denat = 0)) |>
  mutate(
    total = nat + denat,
    prop_nat = if_else(total == 0, NA_real_, 
                       round(nat / total * 100, params$round_prop))) |> 
  ggplot(aes(x = year, y = prop_nat)) +
  		  geom_line(color = params$col_1) +
        scale_y_continuous(limits = c(0, 100)) +
  		  labs(x = "", y = "proportion in % (naturalization on all citizenship changes)") +
  		  params$neutral) |> 
ggplotly()

```

## naturalization rate: which variables?

### age and sex

```{r}
#| label: fig-0701
#| fig-cap: rate age and sex
#| fig-width: 10
#| fig-height: 8

(nat$nat_pop |> 
  group_by(age, sex) |>
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) |>
  mutate(nat_as = if_else(pop == 0, NA_real_, 
                          round(nat / pop * 100, params$round_rate))) |> 
  filter(!is.na(nat_as)) |> 
ggplot() +
		  geom_line(aes(x = age, y = nat_as, color = sex)) +
		  labs(x = "age", y = "naturalization rate (in % per year)", color = "") +
      scale_color_manual(values = params$col_s) +
		  params$neutral) |> 
ggplotly()


```

### year and sex

```{r}
#| label: fig-0702
#| fig-cap: rate year and sex
#| fig-width: 10
#| fig-height: 8

(nat$nat_pop |> 
  group_by(year, sex) |>
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) |> 
  mutate(nat_ys = if_else(pop == 0, NA_real_,
                          round(nat / pop * 100, params$round_rate))) |> 
  filter(!is.na(nat_ys)) |> 
ggplot() +
		  geom_line(aes(x = year, y = nat_ys, color = sex)) +
		  labs(x = "year", y = "naturalization rate (in % per year)", color = "") +
      scale_color_manual(values = params$col_s) +
      scale_y_continuous(limits = c(0, NA)) +
		  params$neutral) |>
ggplotly()

```

### year, age and sex

```{r}
#| label: fig-0703
#| fig-cap: rate year (last three years), age, sex
#| fig-width: 18
#| fig-height: 10


(nat$nat_pop |> 
group_by(year, age, sex) |>
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) |>
  mutate(nat_yas = if_else(pop == 0, NA_real_, 
                           round(nat / pop * 100, params$round_rate))) |> 
  filter(!is.na(nat_yas), year >= (max(year) - 2)) |>  
ggplot() +
		  geom_line(aes(x = age, y = nat_yas, color = sex)) +
		  labs(x = "age", y = "naturalization rate (in % per year)", color = "") +
      facet_wrap(~as.factor(year), nrow = 1) +
      scale_color_manual(values = params$col_s) +
      scale_y_continuous(limits = c(0, NA)) +
		  params$neutral) |>
ggplotly()

```

### district, year and sex

```{r}
#| label: fig-0706
#| fig-cap: rate district, year and sex
#| fig-width: 20
#| fig-height: 20

(nat$nat_pop |> 
  group_by(district, year, sex) |>
  summarize(
    pop = sum(pop),
    nat = sum(nat),
    .groups = "drop"
  ) |>
  mutate(nat_dys = if_else(pop == 0, NA_real_, 
                           round(nat / pop * 100, params$round_rate))) |> 
  filter(!is.na(nat_dys)) |>  
  ggplot() +
	  geom_line(aes(x = year, y = nat_dys, color = sex)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "age", y = "naturalization rate (in % per year)", color = "") +
    scale_color_manual(values = params$col_s) +
	  params$neutral) |>
ggplotly(height = 2000)

```

## das: smoothing (nat and pop)

```{r}
#| label: fig-0707
#| fig-cap: processes by district, age, sex; smoothed
#| fig-width: 15
#| fig-height: 12

process_lev <- c("naturalization", "population")
smooth_lev <- c("initial", "smoothed")

temp_initial <- nat$nat_pop_das_smooth |> 
  gather(`nat`, `pop`, key = category, value = count) |>
  mutate(
    cat = fact_if(category, process_lev, "nat"),
    smooth = factor(smooth_lev[1], levels = smooth_lev)
  ) |>
  select(district, age, sex, cat, smooth, count)

temp_smooth <- nat$nat_pop_das_smooth |> 
  gather(`nat_a`, `pop_a`, key = category, value = count) |>
  mutate(
    cat = fact_if(category, process_lev, "nat_a"),
    smooth = factor(smooth_lev[2], levels = smooth_lev)
  ) |>
  select(district, age, sex, cat, smooth, count)

(bind_rows(temp_initial, temp_smooth) |>
  filter(district %in% params$d_select) |>
  mutate(cat = factor(cat, levels = rev(process_lev))) |> 
  ggplot() +
	  geom_line(aes(x = age, y = count, color = sex, linetype = smooth)) +
	  facet_grid(cat ~ district, scales = "free_y") +
	  labs(x = "age",
	       y = "naturalization per year, population",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_s) +
    scale_linetype_manual(values = c("solid", "dotted")) +
	  params$neutral) |>
ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))


```

## das: naturalization rate

```{r}
#| label: fig-0708
#| fig-cap: rate by district, age and sex
#| fig-width: 20
#| fig-height: 20

(nat$nat_rate_das |> 
  pivot_longer(cols = c(rate_thres, rate_smooth), 
               names_to = "category", values_to = "count") |>  
  mutate(cat = fact_if(category, smooth_lev, "rate_thres")) |>
  select(district, age, sex, cat, count) |> 
  # filter(district %in% params$d_select) |>
  ggplot() +
	  geom_line(aes(x = age, y = count, color = sex, linetype = cat)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "age", y = "naturalization rate (in % per year)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_s) +
    scale_linetype_manual(values = c("solid", "dotted")) +
	  params$neutral) |>
ggplotly(height = 2000)


```

## ya: smooth processes (preparation for trend by ya)

### by years (subjectively, but last year in the plot)

```{r}
#| label: fig-0709
#| fig-cap: processes (smoothed) by year and age
#| fig-width: 15
#| fig-height: 12

temp_initial_ya <- nat$nat_ya |> 
  pivot_longer(cols = c(nat, pop), names_to = "category", values_to = "count") |>  
  mutate(
    cat = fact_if(category, process_lev, "nat"),
    smooth = factor(smooth_lev[1], levels = smooth_lev)
  ) |>
  select(year, age, cat, smooth, count)

temp_smooth_ya <- nat$nat_ya |> 
  pivot_longer(cols = c(nat_a, pop_a), names_to = "category", values_to = "count") |>    
  mutate(
    cat = fact_if(category, process_lev, "nat_a"),
    smooth = factor(smooth_lev[2], levels = smooth_lev)
  ) |>
  select(year, age, cat, smooth, count)

smooth_plot_ya <- bind_rows(temp_initial_ya, temp_smooth_ya)

(smooth_plot_ya |> 
  filter(year >= (max(year)-2)) |> 
  mutate(cat = factor(cat, levels = rev(process_lev))) |>
  ggplot() +
	  geom_line(aes(x = age, y = count, color = cat, linetype = smooth)) +
	  facet_grid(cat ~ factor(year), scales = "free_y") +
	  labs(x = "age",
	       y = "naturalization per year, population",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_6[1:2]) +
    scale_linetype_manual(values = c("solid", "dotted")) +
	  params$neutral) |>
ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```

## ya: rate

```{r}
#| label: fig-0711
#| fig-cap: rate by year and age, focus on age
#| fig-width: 12
#| fig-height: 8

(nat$nat_rate_ya |>
  pivot_longer(cols = c(rate_thres, rate_smooth), 
               names_to = "category", values_to = "count"
  ) |>
  mutate(cat = fact_if(category, smooth_lev, "rate_thres")) |>
  filter(year >= max(year) - 2) |> 
  select(year, age, cat, count) |> 
  ggplot() +
	  geom_line(aes(x = age, y = count, linetype = cat), 
	            color = params$col_1) +
	  facet_wrap(~ factor(year), nrow = 1) +
	  labs(x = "age",
	       y = "naturalization rate (in % per year)",
	       linetype = "") +
    scale_linetype_manual(values = c("solid", "dotted")) +
	  params$neutral) |> 
ggplotly()

```


## constrained regression

### focus age

```{r}
#| label: fig-0713
#| fig-cap: rate by year and age; past and future, focus on age
#| fig-width: 10
#| fig-height: 7

n_years <- length(params$scen_begin:params$scen_end)
pal_year <- colorRampPalette(params$col_6[1:5])(n_years)

(nat$rate_ya_past_pred |> 
  filter(year >= params$scen_begin) |> 
  ggplot() +
	  geom_line(aes(x = age, y = rate_ya, color = as.factor(year))) +
	  labs(x = "age", y = "naturalization rate (in % per year)", color = "") +
    scale_color_manual(values = pal_year) +
    params$neutral) |> 
 ggplotly()

```

### focus years

```{r}
#| label: fig-0714
#| fig-cap: rate by year and age; past and future, focus on years
#| fig-width: 9
#| fig-height: 13

(nat$rate_ya_past_pred |> 
  filter(age %in% seq(0, 100, by = 10)) |> 
  ggplot() +
	  geom_line(aes(x = year, y = rate_ya), 
	            color = params$col_1) +
	  facet_wrap(~ age, ncol = 4, 
	             labeller = labeller(age = function(x) paste("age", x))) +
	  labs(x = "",
	       y = "naturalization rate (in % per year)") +
	  params$neutral) |> 
ggplotly()

```

## smooth processes (preparation for the trend factor calculation)

```{r}
#| label: fig-0715
#| fig-cap: processes by age, smoothed
#| fig-width: 12
#| fig-height: 15

temp_initial_a <- nat$nat_a |> 
  pivot_longer(cols = c(nat, pop), names_to = "category", values_to = "count") |>
  mutate(
    cat = fact_if(category, process_lev, "nat"),
    smooth = factor(smooth_lev[1], levels = smooth_lev)
  ) |>
  select(age, cat, smooth, count)  

temp_smooth_a <- nat$nat_a |> 
  pivot_longer(cols = c(nat_a, pop_a), names_to = "category", values_to = "count") |>
  mutate(
    cat = fact_if(category, process_lev, "nat_a"),
    smooth = factor(smooth_lev[2], levels = smooth_lev)
  ) |>
  select(age, cat, smooth, count)  
  
 
(bind_rows(temp_initial_a, temp_smooth_a) |> 
  mutate(cat = factor(cat, levels = rev(process_lev))) |>
  ggplot() +
	  geom_line(aes(x = age, y = count, color = cat, linetype = smooth)) +
	  facet_wrap(~ cat, ncol = 1, scales = "free_y") +
	  labs(x = "age",
	       y = "quantity per year (naturalization), quantity (population)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_6[1:2]) +
    scale_linetype_manual(values = c("solid", "dotted")) +
	  params$neutral) |> 
ggplotly() |>
  layout(legend = list(x = 1.05, y = 1))

```

## rate

```{r}
#| label: fig-0716
#| fig-cap: rate by age
#| fig-width: 9
#| fig-height: 7

(nat$nat_rate_a |> 
  pivot_longer(
    cols = c(rate_thres, rate_smooth),
    names_to = "category",
    values_to = "count"
  ) |>
  mutate(cat = fact_if(category, smooth_lev, "rate_thres")) |>
  select(age, cat, count) |> 
  ggplot() +
	  geom_line(aes(x = age, y = count, linetype = cat), 
	            color = params$col_1) +
	  labs(x = "age",
	       y = "naturalization rate (in % per year)",
	       linetype = "") +
    scale_linetype_manual(values = c("solid", "dotted")) +
	  params$neutral) |> 
ggplotly()

```

## trend factor ('rate ya' to 'rate a')

```{r}
#| label: fig-0717
#| fig-cap: trend factor by year and age, focus on age
#| fig-width: 10
#| fig-height: 8

(nat$tf_ya |> 
  ggplot() +
    geom_hline(yintercept = 1, col = params$col_grey) +
	  geom_line(aes(x = age, y = tf_ya, color = as.factor(year))) +
	  labs(x = "age", y = "trend factor", color = "") +
    scale_color_manual(values = pal_year) +
    coord_cartesian(ylim = c(0, 2)) +
    params$neutral) |> 
 ggplotly()

```

```{r}
#| label: fig-0718
#| fig-cap: trend factor by year and age, focus on years
#| fig-width: 12
#| fig-height: 15


(nat$tf_ya |>
  filter(age %in% seq(0, 100, by = 10)) |>
  ggplot() +
	  geom_line(aes(x = year, y = tf_ya),
	            color = params$col_1) +
	  facet_wrap(~ age, ncol = 4,
	             labeller = labeller(age = function(x) paste("age", x))) +
	  labs(x = "", y = "trend factor") +
	  params$neutral) |>
ggplotly()

```

## future rate (dyas)

future naturalization rate by district, year, age, sex (male only)

```{r}
#| label: fig-0721
#| fig-cap: future rate by district, year, age, sex (male only)
#| fig-width: 20
#| fig-height: 20

(nat$rate_dyas_future |> 
  filter(sex %in% params$uni_s[1]) |> 
  ggplot() +
	  geom_line(aes(x = age, y = rate_smoothed, color = as.factor(year))) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "age", y = "naturalization rate (in % per year)", color = "") +
    scale_color_manual(values = pal_year) +
	  params$neutral) |>
ggplotly(height = 2000)


```

future naturalization rate by district, year, age, sex (female only)

```{r}
#| label: fig-0722
#| fig-cap: future rate by district, year, age, sex (female only)
#| fig-width: 20
#| fig-height: 20

(nat$rate_dyas_future |> 
  filter(sex %in% params$uni_s[2]) |> 
  ggplot() +
	  geom_line(aes(x = age, y = rate_smoothed, color = as.factor(year))) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "age", y = "naturalization rate (in % per year)", color = "") +
    scale_color_manual(values = pal_year) +
	  params$neutral) |>
ggplotly(height = 2000)

```


