
# Capacity and reserves

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)
# params <- init_scen(i_scen = "middle_car_rev", vars_init_once = vars_once)

# load the capacity-reserves function
source(file.path(root, "R/0801_capacity_reserves_fun.R"))

# execute the capacity-reserves function
car <- capacity_reserves_fun(params = params)

```


## careb

### plot entire city, all owner categories

```{r}
#| label: fig-0800
#| fig-cap: entire city
#| fig-width: 10
#| fig-height: 14

(car$car_dat |> 
  group_by(year, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~ cat, ncol = 2, scales = "free_y") +
	  labs(x = "year of data delivery",
	       y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
ggplotly() |>
  layout(
    legend = list(orientation = "h",  x = 0.5, 
                  xanchor = "center",  y = -0.15, yanchor = "top"))

```

### plot entire city, by owner

```{r}
#| label: fig-0801
#| fig-cap: entire city by owner
#| fig-width: 10
#| fig-height: 18

(car$car_dat |> 
  group_by(year, residence, plot, owner, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_grid(cat ~ owner, scales = "free_y") +
	  labs(x = "year of data delivery",
	       y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly() |>
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",  
                       y = -0.1, yanchor = "top"))

```

### capacity by district, cooperative housing

```{r}
#| label: fig-0802a
#| fig-cap: capacity by district, cooperative housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[1], owner == params$uni_w[1]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))


```

### capacity by district, private housing

```{r}
#| label: fig-0802b
#| fig-cap: capacity by district, private housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[1], owner == params$uni_w[2]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))

```


### buildings by district, cooperative housing

```{r}
#| label: fig-0802c
#| fig-cap: buildings by district, cooperative housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[2], owner == params$uni_w[1]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))

```


### buildings by district, private housing

```{r}
#| label: fig-0802d
#| fig-cap: buildings by district, private housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[2], owner == params$uni_w[2]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))

```



### reserves by district, cooperative housing

```{r}
#| label: fig-0802e
#| fig-cap: reserves by district, cooperative housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[3], owner == params$uni_w[1]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))

```

### reserves by district, private housing

```{r}
#| label: fig-0802f
#| fig-cap: reserves by district, private housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[3], owner == params$uni_w[2]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))


```


### usage by district, cooperative housing

```{r}
#| label: fig-0802g
#| fig-cap: usage by district, cooperative housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[4], owner == params$uni_w[1]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))


```


### usage by district, private housing

```{r}
#| label: fig-0802h
#| fig-cap: usage by district, private housing
#| fig-width: 20
#| fig-height: 20

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(cat == params$car_category[4], owner == params$uni_w[2]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "year of data delivery", y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))

```

### selected districts

```{r}
#| label: fig-0803a
#| fig-cap: selected districts, Hard
#| fig-width: 10
#| fig-height: 18

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(district == "Hard") |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_grid(cat ~ owner, scales = "free_y") +
	  labs(x = "year of data delivery",
	       y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly() |>
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.1, yanchor = "top"))


```



```{r}
#| label: fig-0803b
#| fig-cap: selected districts, Escher Wyss
#| fig-width: 10
#| fig-height: 18

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(district == "Escher Wyss") |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_grid(cat ~ owner, scales = "free_y") +
	  labs(x = "year of data delivery",
	       y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly() |>
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.1, yanchor = "top"))


```



```{r}
#| label: fig-0803c
#| fig-cap: selected districts, Oerlikon
#| fig-width: 10
#| fig-height: 18

(car$car_dat |> 
group_by(district, year, owner, residence, plot, cat) |>
  summarize(ha = sum(ha), .groups = "drop") |> 
  filter(district == "Oerlikon") |> 
  ggplot() +
	  geom_line(aes(x = year, y = ha, color = residence, linetype = plot)) +
	  facet_grid(cat ~ owner, scales = "free_y") +
	  labs(x = "year of data delivery",
	       y = "area (in ha)",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_e) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly() |>
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.1, yanchor = "top"))
```




## combine information (with the parameters)

### reserves (area)

no area values below zero

```{r}
#| label: fig-0804
#| fig-cap: reserves by district and ownership
#| fig-width: 12
#| fig-height: 16

(car$usage_prop |> 
  mutate(district = fct_rev(district)) |> 
  ggplot() +
  geom_col(aes(x = district, y = reserve_new, fill = owner), 
           position = "dodge") +
  coord_flip() +
	labs(x = "", y = "reserves (in ha)", fill = "") +
  scale_fill_manual(values = params$col_w) +
  params$neutral) |> 
  ggplotly()

```

### usage proportion

```{r}
#| label: fig-0805
#| fig-cap: usage proportion by district and ownership
#| fig-width: 12
#| fig-height: 16

(car$usage_prop |> 
  mutate(district = fct_rev(district)) |> 
  ggplot() +
  geom_col(aes(x = district, y = usage_prop, fill = owner), 
           position = "dodge") +
  coord_flip() +
	labs(x = "", y = "usage proportion (in % of the reserves)", fill = "") +
  scale_fill_manual(values = params$col_w) +
  params$neutral) |> 
  ggplotly()


```

### usage area

```{r}
#| label: fig-0806
#| fig-cap: usage area by district and ownership
#| fig-width: 12
#| fig-height: 16

(car$usage_prop |> 
  mutate(district = fct_rev(district)) |> 
  ggplot() +
  geom_col(aes(x = district, y = usage, fill = owner), 
           position = "dodge") +
  coord_flip() +
	labs(x = "", y = "usage area (in ha)", fill = "") +
  scale_fill_manual(values = params$col_w) +
  params$neutral) |> 
  ggplotly()

```

## usage per year
### usage proportion

```{r}
#| label: fig-0807
#| fig-cap: usage proportion by district and year
#| fig-width: 20
#| fig-height: 20

(car$total |> 
  ggplot() +
	  geom_line(aes(x = year, y = usage_prop, color = owner)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "", 
	       y = "usage proportion (in % of the reserves per year)",
	       color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))

```

### usage area

```{r}
#| label: fig-0808
#| fig-cap: usage area by district and year
#| fig-width: 20
#| fig-height: 20


(car$total |> 
  ggplot() +
	  geom_line(aes(x = year, y = usage_area, color = owner)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "", 
	       y = "usage area (in ha per year)",
	       color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000) |> 
  layout(legend = list(orientation = "h",  x = 0.5, xanchor = "center",
                       y = -0.02, yanchor = "top"))

```
