
# Allocation

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the allocation function
source(file.path(root, "R/1001_allocation_fun.R"))

# execute the allocation function
aca <- allocation_fun(params = params)

```


## allocation: past data

### allocation by year

WHY? e.g. to support the base year decision

```{r}
#| label: fig-0900
#| fig-cap: allocation by year
#| fig-width: 8
#| fig-height: 7

(aca$aca_dat |> 
group_by(year) |>
  summarize(
    apartments = sum_NA(apartments),
    people = sum_NA(people),
    .groups = "drop"
  ) |>
  mutate(aca_y = round(people / apartments, params$round_aca)) |>  
  ggplot() +
	  geom_line(aes(x = year, y = aca_y), col = params$col_1) +
	  labs(x = "", y = "allocation (people per apartment)") +
    expand_limits(y = 0) +
	  params$neutral) |> 
ggplotly()

```

### allocation by year and owner

```{r}
#| label: fig-0901
#| fig-cap: allocation by year and owner
#| fig-width: 8
#| fig-height: 7

(aca$aca_yw |> 
  ggplot() +
	  geom_line(aes(x = year, y = aca_yw, color = owner)) +
	  labs(x = "", y = "allocation (people per apartment)", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +
	  params$neutral) |>
ggplotly()

```

## allocation: past data by district

### allocation

```{r}
#| label: fig-0902
#| fig-cap: allocation by district, year and owner
#| fig-width: 20
#| fig-height: 20

(aca$aca_dyw |> 
  ggplot() +
	  geom_line(aes(x = year, y = aca_dyw, color = owner)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "allocation (people per apartment)", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)


```

### apartments

```{r}
#| label: fig-0904
#| fig-cap: apartments by district, year and owner
#| fig-width: 20
#| fig-height: 20

(aca$aca_dyw |> 
  ggplot() +
	  geom_line(aes(x = year, y = apartments, color = owner)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "apartments", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)

```

### people

```{r}
#| label: fig-0905
#| fig-cap: people by district, year and owner
#| fig-width: 12
#| fig-height: 14

(aca$aca_dyw |> 
  ggplot() +
	  geom_line(aes(x = year, y = people, color = owner)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "people", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)

```

## predictions

### entire city, by owner

```{r}
#| label: fig-0906
#| fig-cap: allocation prediction by year and ownership
#| fig-width: 8
#| fig-height: 7

(aca$aca_yw_past_pred |> 
  ggplot() +
    geom_vline(xintercept = c(params$aca_base_begin, params$aca_base_end), 
               color = params$col_grey, linetype = "dashed") +
	  geom_line(aes(x = year, y = aca_yw_all, color = owner)) +
	  labs(x = "", y = "allocation (people per apartment)", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```

### entire city, by district and owner

```{r}
#| label: fig-0907
#| fig-cap: allocation prediction by district, year and ownership
#| fig-width: 20
#| fig-height: 20

(aca$aca_dyw_past_pred |> 
  ggplot() +
    geom_vline(xintercept = c(params$aca_base_begin, params$aca_base_end), 
               color = params$col_grey, linetype = "dashed") +  
	  geom_line(aes(x = year, y = aca_dyw_all, color = owner)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "allocation (people per apartment)", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)

```

### Escher Wyss, by owner {#sec-escherwyss_ls}

WWHY a different approach for the Escher Wyss district, private housing? same approach as in the living space model.

```{r}
#| label: fig-0908
#| fig-cap: allocation prediction for Escher Wyss district, by year and ownership
#| fig-width: 8
#| fig-height: 7


(aca$aca_yw_past_pred_52 |> 
  ggplot() +
    geom_vline(xintercept = c(params$aca_base_begin_52p, params$aca_base_end), 
               color = params$col_grey, linetype = "dashed") +
	  geom_line(aes(x = year, y = aca_dyw_all, color = owner)) +
	  labs(x = "", y = "allocation (person per apartment)", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```

## combine datasets

### past and prediction

```{r}
#| label: fig-0909
#| fig-cap: allocation prediction by district, year and owner; different datasets
#| fig-width: 20
#| fig-height: 20

(aca$aca_comb |> 
  ggplot() +
    geom_vline(xintercept = c(params$aca_base_begin, params$aca_base_end), 
               color = params$col_grey, linetype = "dashed") +  
	  geom_line(aes(x = year, y = aca, color = owner)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "allocation (people per apartment)", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)

```

### allocation prediction per district and entire city: cooperative housing

```{r}
#| label: fig-0910a
#| fig-cap: allocation prediction per district and entire city, cooperative housing
#| fig-width: 20
#| fig-height: 20

names_before <- c("aca_dyw", "aca_yw", "aca")
names_plot <- c("by district", "entire city", "prediction")

aca_both <- aca$aca_comb |> select(district, year, owner, aca_dyw, aca_yw, aca) |>
  pivot_longer(cols = names_before, names_to = "category", values_to = "aca") |>
  mutate(cat = factor(case_when(
    category == names_before[1] ~ names_plot[1],
    category == names_before[2] ~ names_plot[2],
    TRUE ~ names_plot[3]
  ), levels = names_plot))

(aca_both |> 
  filter(owner == params$uni_w[1]) |> 
  ggplot() +
    geom_vline(xintercept = c(params$aca_base_begin, params$aca_base_end), 
               color = params$col_grey, linetype = "dashed") +  
	  geom_line(aes(x = year, y = aca, color = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "allocation (people per apartment)", color = "") +
    scale_color_manual(values = params$col_6[c(1,3,4)]) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)

```

### prediction per district and entire city: private housing

```{r}
#| label: fig-0910b
#| fig-cap: allocation prediction per district and entire city, private housing
#| fig-width: 20
#| fig-height: 20

(aca_both |> 
  filter(owner == params$uni_w[2]) |> 
  ggplot() +
    geom_vline(xintercept = c(params$aca_base_begin, params$aca_base_end), 
               color = params$col_grey, linetype = "dashed") +  
	  geom_line(aes(x = year, y = aca, color = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "allocation (people per apartment)", color = "") +
    scale_color_manual(values = params$col_6[c(1,3,4)]) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)

```






