
# Housing model

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the housing model function
source(file.path(root, "R/0000_General/0024_projects-reserves.R"))
source(file.path(root, "R/1301_housing_fun.R"))

# execute the housing model function
hou <- housing_fun(params = params)

```

## population based on reserves

With the usage prediction (from capacity reserve module) and living space prediction  
the additional (!) yearly (!) population is calculated.
Therefore, these cumulative population values are added to the past population.


```{r}
#| label: fig-1300
#| fig-cap: people, capacity, reserves
#| fig-width: 20
#| fig-height: 20

(hou$pop_with_past |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dashed") +    
	  geom_line(aes(x = year, y = pop, color = owner)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "people", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 1500)

```


## population based on projects and reserves

### population by district and ownership

```{r}
#| label: fig-1302
#| fig-cap: population by district and owner
#| fig-width: 20
#| fig-height: 20

(hou$pop_fut_past |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dashed") +    
	  geom_line(aes(x = year, y = pop, color = owner)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "people", color = "") +
    scale_color_manual(values = params$col_w) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 1500)  


```

### population by district

```{r}
#| label: fig-1303
#| fig-cap: population by district
#| fig-width: 20
#| fig-height: 20

(hou$pop_d |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dashed") +    
	  geom_line(aes(x = year, y = pop), col = params$col_1) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "people") +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 1500)  

```

### entire city

```{r}
#| label: fig-1304
#| fig-cap: population
#| fig-width: 10
#| fig-height: 8

(hou$pop_all |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dashed") +     
	  geom_line(aes(x = year, y = pop), col = params$col_1) +
	  labs(x = "", y = "people") +
    expand_limits(y = 0) +
	  params$neutral) |> 
ggplotly()

```

## compare projects, reserves, final prediction

### projects, reserves, final prediction by district, year, ownership

Projects, reserves, final prediction by district, year, cooperative housing only.

```{r}
#| label: fig-1305a
#| fig-cap: projects, reserves, final prediction by district, year, cooperative housing only
#| fig-width: 20
#| fig-height: 20

cat_level <- c("projects", "reserves", "final")

comp_pro_car <- hou$pro_car |> 
  replace_na(list(pop = 0, car = 0)) |> 
  mutate(car_cum = if_else(year == params$date_end, pop, car),
         pro_y = if_else(year == params$date_end, pop, new - removed)) |> 
  group_by(district, owner) |> 
      arrange(year) |> 
      mutate(pro_cum = cumsum(pro_y)) |> 
  ungroup() |> 
  arrange(district, owner, year) |> 
  select(district, owner, year, pro_cum, car_cum)


pre_res_fin <- hou$pro_res_all |>
  left_join(comp_pro_car, by = c("district", "year", "owner")) |>
  pivot_longer(c(pro_cum, car_cum, pop), names_to = "category", values_to = "people") |>
  mutate(
    district = factor(district, levels = params$uni_d),
    cat = factor(case_when(
      category == "pro_cum" ~ cat_level[1],
      category == "car_cum" ~ cat_level[2],
      TRUE ~ cat_level[3]
    ), levels = cat_level)
  ) |>
  select(district, year, owner, cat, people)
    
(pre_res_fin |> 
  filter(owner == params$uni_w[1]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = people, color = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "people", color = "") +
    scale_color_manual(values = params$col_6[c(1,3,4)]) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)

```

projects, reserves, final prediction by district, year, private housing only

```{r}
#| label: fig-1305b
#| fig-cap: projects, reserves, final prediction by district, year, private housing only
#| fig-width: 20
#| fig-height: 20

(pre_res_fin |> 
  filter(owner == params$uni_w[2]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = people, color = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "people", color = "") +
    scale_color_manual(values = params$col_6[c(1,3,4)]) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)

```




### people by district and ownership (projects vs. reserves)

```{r}
#| label: fig-1306
#| fig-cap: people by district and ownership (projects vs. reserves)
#| fig-width: 15
#| fig-height: 15

# Comment on the data source for this plot: Why not with the object car_spa?
# It is easier to calculate totals with the original objects and closer to the raw data.

pro_res_level <- c("projects", "reserves")

pro_dw <- hou$pro_aca |> 
  group_by(district, owner) |> 
    summarize(new = sum_NA(new), 
              removed = sum_NA(removed),
              .groups = "drop") |> 
  mutate(pro = new - removed) |> 
  select(district, owner, pro)

pro_car_dw <- hou$car_spa |> 
  group_by(district, owner) |> 
    summarize(car = sum_NA(car), 
              .groups = "drop") |> 
  left_join(pro_dw, by = c("district", "owner")) |> 
  pivot_longer(c(car, pro), names_to = "category", values_to = "people") |> 
  mutate(district = factor(district, levels = rev(params$uni_d)),
         owner = factor(owner, levels = params$uni_w),
         cat = fact_if(category, pro_res_level, "pro")) |> 
  select(district, owner, cat, people)
    
(pro_car_dw |>   
  ggplot() +
  geom_col(aes(x = district, y = people, fill = cat), 
           position = "dodge") +
  coord_flip() +
  facet_wrap( ~ owner) +
	labs(x = "", y = "people", fill = "") +
  scale_fill_manual(values = params$col_6[c(1,3)]) +
  params$neutral) |> 
  ggplotly()  

```

### proportion of the projects on the reserves by district and ownership

```{r}
#| label: fig-1307
#| fig-cap: proportion of the projects on the reserves by district and ownership
#| fig-width: 15
#| fig-height: 15

(pro_car_dw |> 
  group_by(district, owner) |> 
  mutate(prop = people / sum_NA(people) * 100) |> 
  filter(cat == pro_res_level[1]) |> 
  ggplot() +
  geom_col(aes(x = district, y = prop), fill = params$col_1, 
           position = "dodge") +
  coord_flip() +
  facet_wrap( ~ owner) +
	labs(x = "", y = "proportion in % (projects on reserves)") +
  params$neutral) |> 
  ggplotly()    
  
```

### people by district (projects vs. reserves)

```{r}
#| label: fig-1308
#| fig-cap: people by district (projects vs. reserves)
#| fig-width: 10
#| fig-height: 10

(pro_car_dw |> 
  group_by(district, cat) |> 
  summarize(people = sum_NA(people),
            .groups = "drop") |> 
  ggplot() +
  geom_col(aes(x = district, y = people, fill = cat), 
           position = "dodge") +
  coord_flip() +
	labs(x = "", y = "people", fill = "") +
  scale_fill_manual(values = params$col_6[c(1,3)]) +
  params$neutral) |> 
  ggplotly()    
  

```


### proportion of the projects on the reserves by district 

```{r}
#| label: fig-1309
#| fig-cap: proportion of the projects on the reserves by district 
#| fig-width: 10
#| fig-height: 10


(pro_car_dw |> 
  group_by(district, cat) |> 
  summarize(people = sum_NA(people),
            .groups = "drop") |> 
  group_by(district) |> 
  mutate(prop = people / sum_NA(people) * 100) |> 
  filter(cat == pro_res_level[1]) |> 
  ggplot() +
  geom_col(aes(x = district, y = prop), fill = params$col_1, 
           position = "dodge") +
  coord_flip() +
	labs(x = "", y = "proportion in % (projects on reserves)") +
  params$neutral) |> 
  ggplotly()    
  
```

