# Demography and housing

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the demography and housing function
source(file.path(root, "R/1401_demography_housing_fun.R"))

# execute the demography and housing model function (approx. 35 seconds)
deh <- demography_housing_fun(params = params)

```


```{r}
#| label: data_preparation
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

#immigration (for ims-plot)
imm_past <- read_csv(params$imm_od) |> 
  rename(year = EreignisDatJahr) |>    
  left_join(params$look_dis, by = "QuarCd") |> 
  mutate(district = factor(distr, params$uni_d)) |>  
  group_by(district, year) |> 
  summarize(ims = sum(AnzZuzuWir),
            .groups = "drop")


#emigration (for ems-plot)
emi_past <- read_csv(params$emi_od) |>
  rename(year = EreignisDatJahr) |>
  left_join(params$look_dis, by = "QuarCd") |>
  mutate(district = factor(distr, params$uni_d)) |>
  group_by(district, year) |>
  summarize(ems = sum(AnzWezuWir),
            .groups = "drop")

#immigration* (for plot)
ims_past <- read_csv(params$rel_od) |>
  rename(year = EreignisDatJahr, ims = AnzUmzuWir) |>
  left_join(params$look_dis, by = "QuarCd") |>
  mutate(district = factor(distr, params$uni_d)) |>
  select(district, year, ims) |>
  bind_rows(imm_past) |>
  group_by(district, year) |>
  summarize(ims = sum(ims),
            .groups = "drop")

#emigration* (for plot)
ems_past <- read_csv(params$rel_od) |>
  rename(year = EreignisDatJahr, ems = AnzUmzuWir) |>
  left_join(params$look_dis, by = c("QuarBisherCd" = "QuarCd")) |>
  mutate(district = factor(distr, params$uni_d)) |>
  select(district, year, ems) |>
  bind_rows(emi_past) |>
  group_by(district, year) |>
  summarize(ems = sum(ems),
            .groups = "drop")
```

## effects of the population limit

### absolute difference

difference: population limit minus population (based on migration trends)\
negative value: not enough space (immigration decreased, emigration increased)\
positive value: too much space (immigration increased, emigration decreased)\

```{r}
#| label: fig-1400
#| fig-cap: absolute difference of population limit
#| fig-width: 20
#| fig-height: 15

pos_neg <- c("positive", "negative")

dat_pos_neg <- deh$out_bal |>
  mutate(
    differ_prop = differ / pop_limit * 100,   
    cat = factor(if_else(differ < 0, pos_neg[2], pos_neg[1]),
                 levels = pos_neg)
  )

(dat_pos_neg |> 
  ggplot() +
  geom_col(aes(x = year, y = differ, fill = cat),
           position = "dodge") +
  facet_wrap( ~ owner) +
	labs(x = "", 
	     y = "people ('population limit' minus 'population based on trends'", 
	     fill = "") +
	facet_wrap(~district, ncol = 4) +    
  scale_fill_manual(values = params$col_6[c(1,3)]) +
  params$neutral) |>
  ggplotly(height = 1500)

```

### relative difference

```{r}
#| label: fig-1401
#| fig-cap: relative difference of population limit
#| fig-width: 20
#| fig-height: 15

(dat_pos_neg |> 
  ggplot() +
  geom_col(aes(x = year, y = differ_prop, fill = cat),
           position = "dodge") +
  facet_wrap( ~ owner) +
	labs(x = "", 
	     y = "difference in % ('population limit' minus 'population based on trends')", 
	     fill = "") +
	facet_wrap(~district, ncol = 4) +    
  scale_fill_manual(values = params$col_6[c(1,3)]) +
  params$neutral) |>
  ggplotly(height = 1500)

```

## migration correction

### immigration-star correction

```{r}
#| label: fig-1402a
#| fig-cap: immigration-star correction
#| fig-width: 20
#| fig-height: 15

cat <- c("past", "initial", "corrected")
proc <- c("immigration*", "emigration*")  
  
ims_ems_past <- ims_past |> 
    left_join(ems_past, by = c("district", "year")) |> 
    rename(ims_past = ims, ems_past = ems) |> 
    pivot_longer(c(ims_past, ems_past),
    names_to = "category", values_to = "people")  

ims_ems_future <- deh$out_bal |> 
  select(district, year, ims, ems, new_ims3, new_ems3) |> 
  pivot_longer(c(ims, ems, new_ims3, new_ems3),
    names_to = "category", values_to = "people")

mig_correction <- ims_ems_past |>
  bind_rows(ims_ems_future) |>
  mutate(
    cat = factor(case_when(
      category %in% c("ims_past", "ems_past") ~ cat[1],
      category %in% c("ims", "ems") ~ cat[2],
      TRUE ~ cat[3]
    ), levels = cat),
    process = factor(if_else(category %in% c("ims_past", "ims", "new_ims3"),
      proc[1], proc[2]
    ), levels = proc)
  )
  
(mig_correction |> 
  filter(process == proc[1]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = people, color = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "immigration* per year", color = "") +
    scale_color_manual(values = params$col_6[c(1,3,4)]) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 1500)

```

### emigration-star correction


```{r}
#| label: fig-1402b
#| fig-cap: emigration-star correction
#| fig-width: 20
#| fig-height: 15

(mig_correction |> 
  filter(process == proc[2]) |> 
  ggplot() +
	  geom_line(aes(x = year, y = people, color = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "emigration* per year", color = "") +
    scale_color_manual(values = params$col_6[c(1,3,4)]) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 1500)

```




## smoothing over age

Population smoothed over age, last prediction year, Swiss only.

```{r}
#| label: fig-1403a
#| fig-cap: population smoothed over age, last prediction year, Swiss only
#| fig-width: 20
#| fig-height: 20

sel_lev <- c("initial", "smoothed")

age_smoothed <- deh$out_pop_smooth |>
  pivot_longer(c(pop_end_year, pop), names_to = "category", values_to = "pop") |>
  mutate(cat = fact_if(category, sel_lev, equals = "pop_end_year"),
    pop_pyramid = round(if_else(sex == params$uni_s[1], -pop, pop)), 0) |>
  filter(year == max(year))

(age_smoothed |> 
  filter(origin == params$uni_o[1]) |> 
  ggplot() +
    geom_hline(yintercept = 0, color = params$col_grey) +    
	  geom_line(aes(x = age, y = pop_pyramid, color = sex, linetype = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_x") +   
    coord_flip() +
	  labs(x = "people", y = "age", color = "", linetype = "") +
    scale_color_manual(values = params$col_s) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000)

```

Population smoothed over age, last prediction year, foreign only.

```{r}
#| label: fig-1403b
#| fig-cap: population smoothed over age, last prediction year, foreign only
#| fig-width: 20
#| fig-height: 20

(age_smoothed |> 
  filter(origin == params$uni_o[2]) |> 
  ggplot() +
    geom_hline(yintercept = 0, color = params$col_grey) +    
	  geom_line(aes(x = age, y = pop_pyramid, color = sex, linetype = cat)) +
	  facet_wrap(~district, ncol = 4, scales = "free_x") +   
    coord_flip() +
	  labs(x = "people", y = "age", color = "", linetype = "") +
    scale_color_manual(values = params$col_s) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +
	  params$neutral) |> 
  ggplotly(height = 2000)

```




## Wollishofen, population and migration

### by year, age and origin

why Wollishofen? need to check age patterns of this district why named d21? 21 is the number for district Wollishofen

```{r}
#| label: fig-1404a
#| fig-cap: Wollishofen population and migration, age, origin, first prediction year
#| fig-width: 10
#| fig-height: 10

catego <- c("immigration", "emigration", "net migration", "population")
line_catego <- c("initial", "final")


# why not piped into plot? to get limits, and another plot (without origin)

pop_mig_21_yao <- deh$out_dem |>
  filter(district == "Wollishofen") |>
  select(year, age, origin, ims_initial, ims, ems_initial, ems, pop_bir_dea, pop_end_year) |>
  mutate(
    net_initial = ims_initial - ems_initial,
    net = ims - ems
  ) |>
  pivot_longer(
    cols = c(
      "ims_initial", "ims", "ems_initial", "ems",
      "net_initial", "net", "pop_bir_dea", "pop_end_year"
    ),
    names_to = "category", values_to = "people"
  ) |>
  group_by(year, age, origin, category) |>
  summarize(
    people = sum_NA(people),
    .groups = "drop"
  ) |>
  mutate(
    cat = factor(case_when(
      category %in% c("ims", "ims_initial") ~ catego[1],
      category %in% c("ems", "ems_initial") ~ catego[2],
      category %in% c("net", "net_initial") ~ catego[3],
      TRUE ~ catego[4]
    ), levels = catego),
    line_cat = factor(if_else(
      category %in% c("ims_initial", "ems_initial", "net_initial", "pop_bir_dea"),
      line_catego[1], line_catego[2]
    ), levels = line_catego))
  
upper_lim_yao <- pop_mig_21_yao |>   
  group_by(cat) |> 
    summarize(upper = max_NA(people),
              .groups = "drop")

(pop_mig_21_yao |> 
  filter(year == params$scen_begin) |> 
  ggplot() +
	  geom_line(aes(x = age, y = people, color = origin, linetype = line_cat)) +
	  facet_wrap(~ cat, ncol = 2, scales = "free_y") +
	  labs(x = "age",
	       y = "people",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_o) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    scale_y_continuous(limits = c(0, max(upper_lim_yao$upper))) +
	  params$neutral) |> 
ggplotly()  

```


```{r}
#| label: fig-1404b
#| fig-cap: Wollishofen population and migration, age, origin, last prediction year
#| fig-width: 10
#| fig-height: 10

(pop_mig_21_yao |> 
  filter(year == params$scen_end) |> 
  ggplot() +
	  geom_line(aes(x = age, y = people, color = origin, linetype = line_cat)) +
	  facet_wrap(~ cat, ncol = 2, scales = "free_y") +
	  labs(x = "age",
	       y = "people",
	       color = "", linetype = "") +
    scale_color_manual(values = params$col_o) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    scale_y_continuous(limits = c(0, max(upper_lim_yao$upper))) +
	  params$neutral) |> 
ggplotly()  



```


