# Model outputs

```{r, include=FALSE}
#| label: prep
#| echo: false
#| results: 'hide'
#| message: false
#| warning: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# load the demography and housing function
source(file.path(root, "R/1501_model_outputs_fun.R"))

# execute the model outputs function
mod <- model_outputs_fun(params = params, BZO_compare = TRUE)

```

## Population

### by year and scenario

```{r}
#| label: fig-1500
#| fig-cap: population by year and scenario
#| fig-width: 8
#| fig-height: 8

(mod$pop_yc |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = scenario)) +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly()

```

### by year, age and sex

```{r}
#| label: fig-1501
#| fig-cap: population by year, age and sex
#| fig-width: 10
#| fig-height: 13


limits_ <- mod$pop_yas |>
  pull(pop_pyramid) |>
  abs() |>
  max(na.rm = TRUE)

(mod$pop_yas |> 
  ggplot() +
	  geom_line(aes(x = age, y = pop_pyramid, 
	                color = sex, linetype = year_factor)) +
	  labs(x = "age", y = "population", color = "", linetype = "") +
    coord_flip() +
    scale_color_manual(values = params$col_s) +
    scale_linetype_manual(values = c("solid", "dotted")) +    
    scale_y_continuous(limits = c(-limits_, limits_)) +
	  params$neutral) |> 
  ggplotly()


```

### by district and year

```{r}
#| label: fig-1502
#| fig-cap: population by district and year
#| fig-width: 20
#| fig-height: 20

(mod$pop_dy |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +    
	  geom_line(aes(x = year, y = pop), color = params$col_1) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "", y = "population") +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 1500)

```

### by district, year and origin

```{r}
#| label: fig-1503
#| fig-cap: population by district, year and origin
#| fig-width: 10
#| fig-height: 14

pop_dyo <- mod$pop |> 
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(district, year, origin) |>
  summarize(pop = sum(pop), .groups = "drop")

(pop_dyo |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +    
	  geom_line(aes(x = year, y = pop, color = origin)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_o) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 1500)


```

### by district, year and origin; foreign population

```{r}
#| label: fig-1504
#| fig-cap: foreign proportion of population by district, year and origin
#| fig-width: 20
#| fig-height: 20

(pop_dyo |>
  pivot_wider(names_from = "origin", values_from = "pop") |>
  mutate(foreign_prop = foreign / (Swiss + foreign) * 100) |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +    
	  geom_line(aes(x = year, y = foreign_prop), color = params$col_1) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "", y = "foreign proportion in %") +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 1500)  
  
```

### by year, origin, scenario

```{r}
#| label: fig-1505
#| fig-cap: population by year, origin and scenario
#| fig-width: 12
#| fig-height: 8

pop_yoc <- mod$pop |>
  group_by(year, origin, scenario) |>
  summarize(pop = sum(pop), .groups = "drop")

(pop_yoc |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = scenario)) +
    facet_wrap(~origin, nrow = 1) + 
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly()


```

### by year, origin, scenario; foreign population

```{r}
#| label: fig-1506
#| fig-cap: foreign proportion of population by year, origin and scenario
#| fig-width: 10
#| fig-height: 8

(pop_yoc |>
  pivot_wider(names_from = "origin", values_from = "pop") |>
  mutate(foreign_prop = foreign / (Swiss + foreign) * 100) |>
  ggplot() +
	  geom_line(aes(x = year, y = foreign_prop, color = scenario)) +
	  labs(x = "", y = "foreign proportion in %", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly()  
  
```

### by district, year, age and sex

```{r}
#| label: fig-1507
#| fig-cap: population by district, year, age and sex
#| fig-width: 20
#| fig-height: 20

(mod$pop |>
  filter(scenario %in% params$uni_c[c(1, 3)],
         year %in% unique(mod$pop_yas$year)) |>
  group_by(district, year, age, sex) |>
  summarize(pop = sum(pop), .groups = "drop") |>
    mutate(pop_pyramid = if_else(sex == params$uni_s[1], -pop, pop), 
           year_factor = as.factor(year)) |> 
   ggplot() +
	  geom_line(aes(x = age, y = pop_pyramid,
	                color = sex, linetype = year_factor)) +
	  labs(x = "age", y = "population", color = "", linetype = "") +
    coord_flip() +
	  facet_wrap(~district, ncol = 4, scales = "free_x") +
    scale_color_manual(values = params$col_s) +
    scale_linetype_manual(values = c("solid", "dotted")) +
	  params$neutral) |>
  ggplotly(height = 2000) 
  
  
  
```

### by district and year; selected years

plot for slides

```{r}
#| label: fig-1508
#| fig-cap:  population by district and year
#| fig-width: 12
#| fig-height: 10

(mod$pop_dy |> 
  filter(year %in% c(params$date_end, params$scen_end_public)) |> 
  mutate(year_text = paste0("year ", year)) |> 
  ggplot() +
  geom_col(aes(x = district, y = pop, fill = year_text), 
           position = "dodge") +
	labs(x = "", y = "people", fill = "") +
  scale_fill_manual(values = params$col_6[c(1,3)]) +
  params$neutral +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |> 
  ggplotly()
  
```

### by district, year and scenario

```{r}
#| label: fig-1509
#| fig-cap:  population by district, year and scenario
#| fig-width: 20
#| fig-height: 20

(mod$pop |>
    group_by(district, year, scenario) |>
    summarize(pop = sum(pop), .groups = "drop") |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = scenario)) +
	  facet_wrap(~district, ncol = 4, scales = "free_y") +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly(height = 2000)  
  
```

### year, age, scenario (children)

```{r}
#| label: fig-1509a
#| fig-cap: year, age, scenario (children)
#| fig-width: 10
#| fig-height: 6

(mod$pop_yac_children |> 
  ggplot(aes(x = year, y = pop, color = scenario)) +
    geom_line() +
    facet_wrap(~age_5, nrow = 1) +
    scale_color_manual(values = params$col_c) +
    labs(x = "", y = "people", color = "") +
    expand_limits(y = 0) +
    params$neutral) |> 
ggplotly()


```


### year, age, birth version (children)

```{r}
#| label: fig-1509b
#| fig-cap: year, age, birth version (children)
#| fig-width: 10
#| fig-height: 6

(mod$pop_yav_children |>
  ggplot(aes(x = year, y = pop, color = scenario)) +
    geom_line() +
    facet_wrap(~age_5, nrow = 1) +
    scale_color_manual(values = params$col_c) +
    labs(x = "", y = "people", color = "") +
    expand_limits(y = 0) +
    params$neutral) |>
ggplotly()

```


## population: new and previous scenarios

### by year and scenario

```{r}
#| label: fig-1510
#| fig-cap:  previous and new population by year, scenario
#| fig-width: 8
#| fig-height: 8

(mod$pop_yc_new_prev |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = scenario, linetype = cat)) +
	  labs(x = "", y = "population", color = "", linetype = "") +
    scale_color_manual(values = params$col_c) +
    scale_linetype_manual(values = c("solid", "dotted")) +
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly()

```

### age: 80plus, by yso 

```{r}
#| label: fig-1511
#| fig-cap: previous and new population for 80+ by sex and origin
#| fig-width: 10
#| fig-height: 10

new_prev <- factor(c("new", "previous"))

pop_80p_new <- mod$pop_middle |>
  filter(age >= 80, 
         year %in% params$uniy_scen_public) |>  
  group_by(year, sex, origin) |> 
  summarize(pop = sum_NA(pop), .groups = "drop") |> 
  mutate(sex = factor(sex, levels = params$uni_s), 
         origin = factor(origin, levels = params$uni_o), 
         cat = new_prev[1])  

pop_80p_prev <- mod$sce |>
  filter(age_class %in% params$age_3t[9:10],
         scenario == params$text_c[3],
         year %in% params$uniy_scen_public) |>
  group_by(year, sex, origin) |>
  summarize(pop = sum_NA(pop), .groups = "drop") |>
  mutate(cat = new_prev[2])

# WHY not piped into plot? used for total (age 80plus, without so)
pop_80p_new_prev <- pop_80p_new |>
  bind_rows(pop_80p_prev)

(pop_80p_new_prev |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = cat)) +
	  facet_grid(sex ~ origin) +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_6[c(1,3)]) +    
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))
    

```

### 80plus, by year

```{r}
#| label: fig-1512
#| fig-cap: previous and new population for 80+ by year
#| fig-width: 10
#| fig-height: 8

(pop_80p_new_prev |> 
  group_by(year, cat) |> 
  summarize(pop = sum_NA(pop), .groups = "drop") |> 
   ggplot() +
	  geom_line(aes(x = year, y = pop, color = cat)) +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_6[c(1,3)]) +    
    expand_limits(y = 0) +  
	  params$neutral) |> 
  ggplotly()
  

```

### age categories, by year

```{r}
#| label: fig-1513
#| fig-cap: previous and new population by age categories
#| fig-width: 10
#| fig-height: 10

(mod$pop_age_new_prev |>
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = cat)) +
	  facet_wrap(~ age_class, ncol = 3) +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_6[c(1,3)]) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly() |>
    layout(legend = list(x = 1.05, y = 1))

```

### only younger categories

```{r}
#| label: fig-1514
#| fig-cap: previous and new population for 19- by age categories
#| fig-width: 10
#| fig-height: 6

(mod$pop_age_new_prev |> 
  filter(age_class %in% c("0-9", "10-19")) |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = cat)) +
	  facet_wrap(~ age_class, nrow = 1) +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_6[c(1,3)]) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

    
```

### only elder categories

```{r}
#| label: fig-1515
#| fig-cap: previous and new population for 60+ by age categories
#| fig-width: 10
#| fig-height: 6

(mod$pop_age_new_prev |> 
  filter(age_class %in% c("60-69", "70-79", "80+")) |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = cat)) +
	  facet_wrap(~ age_class, nrow = 1) +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_6[c(1,3)]) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```

### new and previous: by dy (selected years only)

```{r}
#| label: fig-1516
#| fig-cap: previous and new population by district
#| fig-width: 12
#| fig-height: 10

pop_dy_new_sel <- mod$pop_dy |>
  filter(year %in% c(params$date_end, mod$comp_year)) |>
  mutate(cat = if_else(year == params$date_end,
                       paste0(params$date_end, " (past)"),
                       paste0(mod$comp_year, " (new)")))

pop_dy_prev_sel <- mod$sce |>
  filter((scenario == params$text_c[3]) &
           (year == mod$comp_year)) |>
  group_by(district, year) |>
    summarize(pop = sum_NA(pop),
              .groups = "drop") |>
  mutate(cat = paste0(mod$comp_year, " (previous)"))

levels_cate <- c(paste0(params$date_end, " (past)"),
                 paste0(mod$comp_year, " (previous)"),
                 paste0(mod$comp_year, " (new)"))


(pop_dy_new_sel |>
  bind_rows(pop_dy_prev_sel) |>
  mutate(cat = factor(cat, levels = levels_cate)) |>
  ggplot() +
  geom_col(aes(x = district, y = pop, fill = cat),
           position = "dodge") +
	labs(x = "", y = "people", fill = "") +
  scale_fill_manual(values = params$col_6[c(1,3,4)]) +
  params$neutral +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |>
  ggplotly()

```

## birth

### by year and scenario (main scenarios)

```{r}
#| label: fig-1520
#| fig-cap: births by year and scenario
#| fig-width: 8
#| fig-height: 8

bir <- mod$bir_past |>
  bind_rows(mod$bir_lower) |>
  bind_rows(mod$bir_middle) |>
  bind_rows(mod$bir_upper) |>
  mutate(district = factor(district, params$uni_d),
         sex = factor(sex, levels = params$uni_s),
         origin = factor(origin, levels = params$uni_o))

(bir |>
  group_by(year, scenario) |>
  summarize(bir = sum(bir), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = bir, color = scenario)) +
	  labs(x = "", y = "births per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```


### by year, origin and scenario

```{r}
#| label: fig-1521
#| fig-cap: births by year, origin and scenario
#| fig-width: 10
#| fig-height: 8

(bir |>
  group_by(year, origin, scenario) |>
  summarize(bir = sum(bir),
            .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = bir, color = scenario)) +
    facet_wrap(~origin, nrow = 1) +  
	  labs(x = "", y = "births per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()  
  
```

### births by year and birth versions (middle scenario)

```{r}
#| label: fig-1521a
#| fig-cap: births by year and birth version (middle scenario)
#| fig-width: 8
#| fig-height: 7

# birth: past and future (birth versions)
bir_ver <- mod$bir_past |>
  bind_rows(mod$bir_middle) |> 
  bind_rows(mod$bir_middle_birth_lower) |> 
  bind_rows(mod$bir_middle_birth_upper) |> 
  mutate(scenario = factor(scenario, levels = params$uni_cb),
         sex = factor(sex, levels = params$uni_s),
         origin = factor(origin, levels = params$uni_o))

(bir_ver |>
  group_by(year, scenario) |>
  summarize(bir = sum(bir), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = bir, color = scenario)) +
	  labs(x = "", y = "births per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```


### TFR by year and birth versions (middle scenario)

```{r}
#| label: fig-1521b
#| fig-cap: TFR by year and birth version (middle scenario)
#| fig-width: 8
#| fig-height: 7

# population: past and future (for birth versions)
pop_ver_yac <- mod$pop_birth_versions |> 
  filter(sex == params$uni_s[2], 
         age >= params$bir_age_begin, 
         age <= params$bir_age_end) |>
  mutate(scenario = factor(scenario, levels = params$uni_cb)) |>
  group_by(year, age, scenario) |>
  summarize(pop = sum_NA(pop), .groups = "drop")

# births (for birth versions)
bir_ver_yac <- bir_ver |>
  rename(age = age_mother) |>
  group_by(year, age, scenario) |>
  summarize(bir = sum_NA(bir), .groups = "drop")

# TFR
pop_bir_ver_yac <- pop_ver_yac |>
  left_join(bir_ver_yac, by = c("year", "age", "scenario")) |>
  replace_na(list(bir = 0))

(pop_bir_ver_yac |>
  mutate(fer_yac = bir / pop) |>
  group_by(year, scenario) |>
  summarize(tfr = sum_NA(fer_yac), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = tfr, color = scenario)) +
	  labs(x = "", y = "TFR", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()  
  
```

### TFR by year, origin of mother, and birth versions (middle scenario)

```{r}
#| label: fig-1521b2
#| fig-cap: TFR by year, origin of mother,  and birth version (middle scenario)
#| fig-width: 8
#| fig-height: 7

# births: past
bir_aoc_past <- read_csv(params$bir_od) |>
  rename(year = EreignisDatJahr, age_mother = AlterVMutterCd, bir = AnzGebuWir) |> 
  mutate(origin_mother = fact_if(HerkunftMutterCd, params$uni_o)) |> 
  group_by(year, age_mother, origin_mother) |> 
  summarize(bir = sum(bir), .groups = "drop") |> 
  mutate(scenario = params$uni_cb[1]) |> 
  select(scenario, year, age_mother, origin_mother, bir)

# births: past and future
bir_aoc_past_future <- mod$bir_dao_middle |> 
  bind_rows(mod$bir_dao_middle_birth_lower) |> 
  bind_rows(mod$bir_dao_middle_birth_upper) |>   
  group_by(scenario, year, age_mother, origin_mother) |> 
  summarize(bir = sum(bir), .groups = "drop") |> 
  bind_rows(bir_aoc_past)

# population: past and future (for birth versions)
pop_ver_yaoc <- mod$pop_birth_versions |> 
  filter(sex == params$uni_s[2], 
         age >= params$bir_age_begin, 
         age <= params$bir_age_end) |>
  mutate(scenario = factor(scenario, levels = params$uni_cb)) |>
  group_by(year, age, origin, scenario) |>
  summarize(pop = sum_NA(pop), .groups = "drop") |> 
  rename(age_mother = age, origin_mother = origin)

# TFR
pop_bir_yaoc <- pop_ver_yaoc |>
  left_join(bir_aoc_past_future, by = c("year", "age_mother", "origin_mother", "scenario")) |>
  replace_na(list(bir = 0)) |> 
  mutate(origin_mother = factor(origin_mother, levels = params$uni_o))

(pop_bir_yaoc |>
  mutate(fer_yaoc = bir / pop) |>
  group_by(year, origin_mother, scenario) |>
  summarize(tfr = sum_NA(fer_yaoc), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = tfr, color = scenario)) +
	  labs(x = "", y = "TFR", color = "") +
    scale_color_manual(values = params$col_c) +
    facet_wrap(~origin_mother, nrow = 1) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()  

```



### fertility rate by year and birth versions (middle scenario)

```{r}
#| label: fig-1521c
#| fig-cap: fertility rate by year and birth version (middle scenario)
#| fig-width: 8
#| fig-height: 7

# fertility rate
(pop_bir_ver_yac |> 
  group_by(year, scenario) |> 
  summarize(bir = sum_NA(bir), pop = sum_NA(pop), .groups = "drop") |> 
  mutate(fer_yc = bir / pop * 100) |> 
  ggplot() +
	  geom_line(aes(x = year, y = fer_yc, color = scenario)) +
	  labs(x = "", y = "fertility rate (in % per year)", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()    
  
```



### births by district, year and origin

```{r}
#| label: fig-1522
#| fig-cap: births by district, year and origin
#| fig-width: 20
#| fig-height: 20

bir_dyo <- bir |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(district, year, origin) |>
  summarize(bir = sum(bir), .groups = "drop")

(bir_dyo |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin,
               color = params$col_grey, linetype = "dotted") +
	  geom_line(aes(x = year, y = bir, color = origin)) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "", y = "births", color = "") +
    scale_color_manual(values = params$col_o) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)

```

### foreign proportion of births by district, year

```{r}
#| label: fig-1523
#| fig-cap: foreign proportion of births by district, year
#| fig-width: 20
#| fig-height: 20

(bir_dyo |>
  pivot_wider(names_from = "origin", values_from = "bir") |>
  mutate(foreign_prop = foreign / (Swiss + foreign) * 100) |> 
   ggplot() +
    geom_vline(xintercept = params$scen_begin,
               color = params$col_grey, linetype = "dotted") +
	  geom_line(aes(x = year, y = foreign_prop), color = params$col_1) +
	  facet_wrap(~district, ncol = 4) +
	  labs(x = "", y = "births: foreign proportion in %") +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)

```

## death

### by year and scenario

```{r}
#| label: fig-1530
#| fig-cap: deaths by year and scenario
#| fig-width: 8
#| fig-height: 8

dea <- mod$dem_future |>
  select(year, age, sex, scenario, dea) |>
  group_by(year, age, sex, scenario) |>
  summarize(dea = sum(dea), .groups = "drop") |>
  bind_rows(mod$dea_past) |>
  left_join(params$look_a3, by = "age") |>
  mutate(sex = factor(sex, levels = params$uni_s))

(dea |>
  group_by(year, scenario) |>
  summarize(dea = sum(dea), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = dea, color = scenario)) +
	  labs(x = "", y = "deaths per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```

### by year, age and scenario

```{r}
#| label: fig-1531
#| fig-cap: deaths by year, age and scenario
#| fig-width: 8
#| fig-height: 10

(dea |>
  group_by(year, age_3, scenario) |>
  summarize(dea = sum(dea), .groups = "drop") |>  
  ggplot() +
	  geom_line(aes(x = year, y = dea, color = scenario)) +
	  labs(x = "", y = "deaths per year", color = "") +
    facet_wrap(~age_3, ncol = 3) +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 800)


```

### by year and sex

for the middle scenario

```{r}
#| label: fig-15232
#| fig-cap: deaths by year and sex
#| fig-width: 8
#| fig-height: 8

(dea |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(year, sex) |>
  summarize(dea = sum(dea), .groups = "drop") |>
   ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +       
	  geom_line(aes(x = year, y = dea, color = sex)) +
	  labs(x = "", y = "deaths per year", color = "") +
    scale_color_manual(values = params$col_s) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly() 
  
```

### by year, age and sex

```{r}
#| label: fig-1533
#| fig-cap: deaths by year, age and sex
#| fig-width: 8
#| fig-height: 10

(dea |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(year, age_3, sex) |>
  summarize(dea = sum(dea), .groups = "drop") |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +   
	  geom_line(aes(x = year, y = dea, color = sex)) +
	  labs(x = "", y = "deaths per year", color = "") +
    facet_wrap(~age_3, ncol = 3) +
    scale_color_manual(values = params$col_s) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 800)
  
```

## immigration

### by year and scenario

```{r}
#| label: fig-1540
#| fig-cap: immigration by year and scenario
#| fig-width: 8
#| fig-height: 8

imm <- mod$dem_future |>
  select(district, year, age, sex, origin, scenario, imm) |>
  bind_rows(mod$imm_past) |>
  left_join(params$look_a3, by = "age") |>
  mutate(district = factor(district, params$uni_d),
         sex = factor(sex, levels = params$uni_s),
         origin = factor(origin, levels = params$uni_o))

(imm |>
  group_by(year, scenario) |>
  summarize(imm = sum(imm), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = imm, color = scenario)) +
	  labs(x = "", y = "immigration per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()


```

### by year and age

```{r}
#| label: fig-1541
#| fig-cap: immigration by year and age
#| fig-width: 10
#| fig-height: 8

(imm |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(year, age_3) |>
  summarize(imm = sum(imm), .groups = "drop") |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = imm), color = params$col_1) +
	  labs(x = "", y = "immigration per year") +
    facet_wrap(~age_3, ncol = 4) +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 800)  
  

```

### by year, age (children) and scenario

```{r}
#| label: fig-1541a
#| fig-cap: immigration by year, age (children) and scenario
#| fig-width: 10
#| fig-height: 6

imm_a5 <- imm |> 
  filter(age < params$age_5[4]) |> 
  left_join(params$look_a5, by = "age") |>   
  group_by(year, age_5, scenario) |>
  summarize(imm = sum(imm), .groups = "drop")

(imm_a5 |> 
  ggplot() +
	  geom_line(aes(x = year, y = imm, color = scenario)) +
	  labs(x = "", y = "immigration per year", color = "") +
    facet_wrap(~age_5, nrow = 1) +  
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```





### by year, age and origin

```{r}
#| label: fig-1542
#| fig-cap: immigration by year, age and origin
#| fig-width: 11
#| fig-height: 7

(imm |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(year, age_3, origin) |>
  summarize(imm = sum(imm), .groups = "drop") |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = imm, color = origin)) +
	  labs(x = "", y = "immigration per year", color = "") +
    facet_wrap(~age_3, ncol = 3) +  
    scale_color_manual(values = params$col_o) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 700)
  
```

### by district, year and origin

```{r}
#| label: fig-1543
#| fig-cap: immigration by district, year and origin
#| fig-width: 20
#| fig-height: 20

imm_dyo <- imm |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(district, year, origin) |>
  summarize(imm = sum(imm), .groups = "drop")

(imm_dyo |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = imm, color = origin)) +
	  labs(x = "", y = "immigration per year", color = "") +
    facet_wrap(~district, ncol = 4, scales = "free_y") +
    scale_color_manual(values = params$col_o) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)


```

### by district, year and origin; foreign population

```{r}
#| label: fig-1544
#| fig-cap: foreign proportion of immigration by district, year and origin
#| fig-width: 20
#| fig-height: 20

(imm_dyo |>
  pivot_wider(names_from = "origin", values_from = "imm") |>
  mutate(foreign_prop = foreign / (Swiss + foreign) * 100) |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = foreign_prop), color = params$col_1) +
	  labs(x = "", y = "foreign proportion (in immigration) in %") +
    facet_wrap(~district, ncol = 4, scales = "free_y") +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)
  
```

## emigration

### by year and scenario

```{r}
#| label: fig-1550
#| fig-cap: emigration by year and scenario
#| fig-width: 8
#| fig-height: 8

# past and future
emi <- mod$dem_future |>
  select(district, year, age, sex, origin, scenario, emi) |>
  bind_rows(mod$emi_past) |>
  left_join(params$look_a3, by = "age") |>
  mutate(district = factor(district, params$uni_d),
         sex = factor(sex, levels = params$uni_s),
         origin = factor(origin, levels = params$uni_o))

(emi |>
  group_by(year, scenario) |>
  summarize(emi = sum(emi), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = emi, color = scenario)) +
	  labs(x = "", y = "emigration per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```

### by year and age

```{r}
#| label: fig-1551
#| fig-cap: emigration by year and age
#| fig-width: 10
#| fig-height: 8

(emi |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(year, age_3) |>
  summarize(emi = sum(emi), .groups = "drop") |> 
   ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = emi), color = params$col_1) +
	  labs(x = "", y = "emigration per year") +
    facet_wrap(~age_3, ncol = 4) +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 800)     
    

```

### by year, age (children) and scenario

```{r}
#| label: fig-1551a
#| fig-cap: emigration by year, age (children) and scenario
#| fig-width: 10
#| fig-height: 6

emi_a5 <- emi |> 
  filter(age < params$age_5[4]) |> 
  left_join(params$look_a5, by = "age") |>   
  group_by(year, age_5, scenario) |>
  summarize(emi = sum(emi), .groups = "drop") 
  
(emi_a5 |> 
  ggplot() +
	  geom_line(aes(x = year, y = emi, color = scenario)) +
	  labs(x = "", y = "emigration per year", color = "") +
    facet_wrap(~age_5, nrow = 1) +  
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()  

```

### by year, age and origin

```{r}
#| label: fig-1552
#| fig-cap: emigration by year, age and origin
#| fig-width: 11
#| fig-height: 7

(emi |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(year, age_3, origin) |>
  summarize(emi = sum(emi), .groups = "drop") |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = emi, color = origin)) +
	  labs(x = "", y = "emigration per year", color = "") +
    facet_wrap(~age_3, ncol = 3) +  
    scale_color_manual(values = params$col_o) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 700)
  
```

### by district, year and origin

```{r}
#| label: fig-1553
#| fig-cap: emigration by district, year and origin
#| fig-width: 20
#| fig-height: 20

emi_dyo <- emi |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(district, year, origin) |>
  summarize(emi = sum(emi), .groups = "drop")

(emi_dyo |> 
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = emi, color = origin)) +
	  labs(x = "", y = "emigration per year", color = "") +
    facet_wrap(~district, ncol = 4, scales = "free_y") +
    scale_color_manual(values = params$col_o) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)

```

### by district, year and origin; foreign population

```{r}
#| label: fig-1554
#| fig-cap: foreign proportion of emigration by district, year and origin
#| fig-width: 20
#| fig-height: 20

(emi_dyo |>
  pivot_wider(names_from = "origin", values_from = "emi") |>
  mutate(foreign_prop = foreign / (Swiss + foreign) * 100) |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = foreign_prop), color = params$col_1) +
	  labs(x = "", y = "foreign proportion (in emigration) in %") +
    facet_wrap(~district, ncol = 4, scales = "free_y") +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)
  
```

## net migration

### by year and scenario

```{r}
#| label: fig-1560
#| fig-cap: net migration by year and scenario
#| fig-width: 8
#| fig-height: 7

# calculate net migration
net <- as_tibble(expand_grid(district = params$uni_d,
                             year = params$date_start:params$scen_end,
                             age = params$age_min:params$age_max,
                             sex = params$uni_s,
                             origin = params$uni_o,
                             scenario = params$uni_c)) |>
  left_join(params$look_a3, by = "age") |>
  left_join(imm, by = c("district", "year", "age", "sex", "origin", "scenario", "age_3")) |>
  left_join(emi, by = c("district", "year", "age", "sex", "origin", "scenario", "age_3")) |>
  replace_na(list(imm = 0, emi = 0)) |>
  mutate(net = if_else(((year < params$scen_begin) & (scenario != params$uni_c[1])) |
                         ((year >= params$scen_begin) & (scenario == params$uni_c[1])),
                       NA_real_, imm - emi))


(net |>
  group_by(year, scenario) |>
  summarize(net = sum(net), .groups = "drop") |>
  ggplot() +
    geom_hline(yintercept = 0, 
               color = params$col_grey) +      
	  geom_line(aes(x = year, y = net, color = scenario)) +
	  labs(x = "", y = "net migration per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()

```

### by district, year and origin

```{r}
#| label: fig-1561
#| fig-cap: net migration by district, year and origin
#| fig-width: 20
#| fig-height: 20

(net |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(district, year, origin) |>
  summarize(net = sum_NA(net), .groups = "drop") |>
  ggplot() +
    geom_hline(yintercept = 0, 
               color = params$col_grey) +      
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = net, color = origin)) +
	  labs(x = "", y = "net migration per year", color = "") +
    facet_wrap(~district, ncol = 4) +
    scale_color_manual(values = params$col_o) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly(height = 1500)  
  
```



### by year, age (children) and scenario

```{r}
#| label: fig-1562
#| fig-cap: net migration by year and age (children)
#| fig-width: 10
#| fig-height: 6

(imm_a5 |> 
  left_join(emi_a5, by = c("year", "age_5", "scenario")) |> 
  mutate(net = imm - emi) |> 
  ggplot() +
    geom_hline(yintercept = 0, 
               color = params$col_grey) +      
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +      
	  geom_line(aes(x = year, y = net, color = scenario)) +
	  labs(x = "", y = "net migration per year", color = "") +
    facet_wrap(~age_5, nrow = 1) +  
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()    
  

```


## naturalization

### by year and scenario

```{r}
#| label: fig-1570
#| fig-cap: naturalization by year and scenario
#| fig-width: 8
#| fig-height: 7

# past and future
nat <- mod$nat_future |>
  bind_rows(mod$nat_past) |>
  left_join(params$look_a3, by = "age") |>
  mutate(
    district = factor(district, levels = params$uni_d),
    sex = factor(sex, levels = params$uni_s)
  )


(nat |>
  group_by(year, scenario) |>
  summarize(nat = sum(nat),
            .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = nat, color = scenario)) +
	  labs(x = "", y = "naturalizations per year", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
	  params$neutral) |>
  ggplotly()


```

### by district and year

```{r}
#| label: fig-1571
#| fig-cap: naturalization by district and year
#| fig-width: 20
#| fig-height: 20

(nat |>
  filter(scenario %in% params$uni_c[c(1, 3)]) |>
  group_by(district, year) |>
  summarize(nat = sum(nat),
            .groups = "drop") |>
  ggplot() +
    geom_vline(xintercept = params$scen_begin, 
               color = params$col_grey, linetype = "dotted") +  
	  geom_line(aes(x = year, y = nat), color = params$col_1) +
	  labs(x = "", y = "naturalizations per year") +
    facet_wrap(~district, ncol = 4) +
    expand_limits(y = 0) +
	  params$neutral + 
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |>
  ggplotly(height = 1500)
    

```

## relocation

### by year, scenario and category

past: no district
WHY? since for future data only one district available
one district does not make sense for relocation
(the district comes from immigration* and emigration*)


```{r}
#| label: fig-1580
#| fig-cap: relocation by year, scenario and category
#| fig-width: 10
#| fig-height: 8

# categories
uni_cat <- c("data", "from immigration*", "from emigration*")

# past
rel_past_yaso <- mod$rel_past |>
  select(year, age, sex, origin, scenario, rel) |>
  group_by(year, age, sex, origin, scenario) |>
  summarize(rel = sum(rel), .groups = "drop") |>
  mutate(cat = factor(uni_cat[1], levels = uni_cat))

# future
rel_future_yaso <- mod$dem_future |>
  select(year, age, sex, origin, scenario, rei, ree) |>
  group_by(year, age, sex, origin, scenario) |>
  summarize(rei = sum(rei), ree = sum(ree), .groups = "drop") |>
  pivot_longer(cols = c("rei", "ree"), values_to = "rel", names_to = "category") |>
  mutate(cat = fact_if(category, uni_cat, equals = "rei", 2, 3)) |>
  select(year, age, sex, origin, scenario, rel, cat)

# past and future combined
(rel_past_yaso |>
  bind_rows(rel_future_yaso) |>
  group_by(year, scenario, cat) |>
  summarize(rel = sum(rel), .groups = "drop") |>
  ggplot() +
	  geom_line(aes(x = year, y = rel, color = scenario, linetype = cat)) +
	  labs(x = "", y = "relocations per year", color = "", linetype = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +
    scale_linetype_manual(values = c("solid", "dashed", "dotted")) +
	  params$neutral) |>
  ggplotly()


```


## BZO comparison

### by year and scenario

```{r}
#| label: fig-1590
#| fig-cap:  BZO versions by year, scenario
#| fig-width: 8
#| fig-height: 8

BZO_names <- c("BZO 2016", "BZO revision")

pop_BZO_old <- mod$pop |>
    group_by(district, year, scenario) |>
    summarize(pop = sum(pop), .groups = "drop") |>
    mutate(BZO = BZO_names[1])

pop_BZO_new <- mod$pop_car_rev |>
  group_by(district, year, scenario) |>
  summarize(pop = sum(pop),
            .groups = "drop") |>
  filter(year <= params$scen_end_public) |>
  mutate(BZO = BZO_names[2], 
         district = factor(district, levels = params$uni_d))

pop_BZO <- pop_BZO_old |>
  bind_rows(pop_BZO_new)

# (pop_BZO |>
#   group_by(year, scenario, BZO) |>
#   summarize(pop = sum(pop),
#             .groups = "drop") |>
#   ggplot() +
# 	  geom_line(aes(x = year, y = pop, color = scenario, linetype = BZO)) +
# 	  labs(x = "", y = "population", color = "", linetype = "") +
#     scale_color_manual(values = params$col_c) +
#     scale_linetype_manual(values = c("solid", "dotted")) +
#     expand_limits(y = 0) +
# 	  params$neutral) |>
#   ggplotly()

```


### by district, year, scenario

```{r}
#| label: fig-1591
#| fig-cap:  BZO versions by district, year, scenario
#| fig-width: 20
#| fig-height: 20

# (pop_BZO |>
#   ggplot() +
# 	  geom_line(aes(x = year, y = pop, color = scenario, linetype = BZO)) +
# 	  facet_wrap(~district, ncol = 3, scales = "free_y") +
# 	  labs(x = "", y = "population", color = "", linetype = "") +
#     scale_color_manual(values = params$col_c) +
#     scale_linetype_manual(values = c("solid", "dotted")) +
#     expand_limits(y = 0) +
# 	  params$neutral) |>
#   ggplotly(height = 2000)


```

comparison for the last pusblished scenario year


```{r}
#| label: fig-1592
#| fig-cap:  BZO versions by district, last scenario year, scenario
#| fig-width: 14
#| fig-height: 12

# (pop_BZO |>
#   filter(year %in% c(params$date_end, params$scen_end_public)) |> 
#   mutate(year_text = paste0("year ", year)) |>
#   ggplot() +
#   geom_col(aes(x = district, y = pop, fill = scenario, alpha = BZO),
#            position = "dodge") +
# 	labs(x = "", y = "people", fill = "", alpha = "") +
#   scale_fill_manual(values = params$col_c) +
#   scale_alpha_manual(values = c(1, 0.5)) +
#   params$neutral +
#   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |>
#   ggplotly()


```



