# Evaluation

```{r, include=FALSE}
#| label: prep
#| output: false
#| cache: true

# project root
library(rprojroot)
root <- find_root(is_rstudio_project)

# get the parameters
source(file.path(root, "R/0000_General/0002_init.R"))
vars_once <- init_once(i_data_path = "~/v/Szen/BevSzen/2026")
params <- init_scen(i_scen = "middle", vars_init_once = vars_once)

# import the data of the past (population, processes)
real_pop <- read_csv(params$pop_od)
real_bir <- read_csv(params$bir_od)
real_dea <- read_csv(params$dea_od)
real_imm <- read_csv(params$imm_od)
real_emi <- read_csv(params$emi_od)
real_nat <- read_csv(params$nat_od)

# as of 2016: similar model structure
similar_model <- 2016

# import the previous scenarios (for population, processes)
scen_pop <- read_csv(params$eval_pop_path) |>
  filter(PublJahr >= similar_model)

scen_prc <- read_csv(params$eval_prc_path) |>
  filter(PublJahr >= similar_model)


```


## Population

### Population by year

```{r}
#| label: fig-1601
#| fig-cap: evaluation by year (middle scenario only)
#| fig-width: 10
#| fig-height: 10

# real population by year
real_pop_y <- real_pop |> 
  rename(year = StichtagDatJahr) |> 
  group_by(year) |> 
  summarize(pop = sum(AnzBestWir), .groups = "drop") |> 
  mutate(cat = "reality") |> 
  select(year, pop, cat)
  
# past scenarios by year (middle scenario only)
scen_pop_y <- scen_pop |> 
  filter(VersionArtCd == 2) |> 
  rename(pub = PublJahr, year = Jahr) |> 
  group_by(pub, year) |> 
  summarize(pop = sum(AnzBestWir), .groups = "drop") |> 
  mutate(cat = paste("scenarios", pub)) |> 
  select(year, pop, cat)


# color palette (reality: black)
col_pal <- colorRampPalette(params$col_6[1:5])(length(unique(scen_pop_y$cat)) + 1)
col_pal[1] <- "black"


(real_pop_y |> 
  bind_rows(scen_pop_y) |> 
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |> 
  ggplot() +
  geom_line(aes(x = year, y = pop, color = cat, linewidth = line_size)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "population", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>  
 ggplotly() 


```


```{r}
#| label: fig-1602
#| fig-cap: evaluation by year (lower and upper scenario)
#| fig-width: 10
#| fig-height: 10


# real population by year
real_pop_y_low_upp <- real_pop_y |> 
  mutate(scenario = factor(params$text_c[1], levels = params$text_c)) |> 
  select(year, cat, scenario, pop)

# past scenarios by year (lower and upper scenario)
scen_pop_y_low_upp <- scen_pop |> 
  filter(VersionArtCd %in% c(1,3)) |> 
  mutate(scenario = factor(if_else(VersionArtCd == 1, params$text_c[2], 
                                   params$text_c[4]), 
                           levels = params$text_c)) |> 
  rename(pub = PublJahr, year = Jahr) |>
  group_by(pub, year, scenario) |>
  summarize(pop = sum(AnzBestWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(year, cat, scenario, pop)

# plot
(real_pop_y_low_upp |> 
  bind_rows(scen_pop_y_low_upp) |> 
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |> 
  ggplot() +
  geom_line(aes(x = year, y = pop, color = cat, 
                linewidth = line_size, linetype = scenario)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  scale_linetype_manual(values = c("solid", "dotdash", "dotted")) +
  labs(x = "", y = "population", color = "", linewidth = "", linetype = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>  
 ggplotly() 


  
```


```{r}
#| label: fig-1603
#| fig-cap: evaluation by year (lower, middle, and upper scenario)
#| fig-width: 7
#| fig-height: 20

# past scenarios by year
scen_pop_yc <- scen_pop |> 
  filter(VersionArtCd %in% c(1,2,3)) |> 
  mutate(scenario = factor(
    case_when(
      VersionArtCd == 1 ~ params$text_c[2],
      VersionArtCd == 2 ~ params$text_c[3],
      TRUE       ~ params$text_c[4]
  ))) |> 
  rename(pub = PublJahr, year = Jahr) |>
  group_by(pub, year, scenario) |>
  summarize(pop = sum(AnzBestWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(year, scenario, cat, pop)


# real population by year
real_pop_yc <- real_pop_y |> 
  mutate(scenario = factor(params$text_c[1], levels = params$text_c)) |> 
  select(year, scenario, pop)

# plot
(expand_grid(year = sort(unique(real_pop_yc$year)),
            scenario = sort(unique(real_pop_yc$scenario)),
            cat = sort(unique(scen_pop_yc$cat))) |> 
  left_join(real_pop_yc, by = c("year", "scenario")) |> 
  bind_rows(scen_pop_yc) |> 
  ggplot() +
  geom_line(aes(x = year, y = pop, color = scenario)) +
  facet_wrap(~ cat, ncol = 2) +
  scale_color_manual(values = params$col_c) +
  labs(x = "", y = "population", color = "") +
  expand_limits(y = 0) +
  params$neutral + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) |> 
 ggplotly(height = 2000)


```

### Population by district and year

Evaluation by district (last publication)

```{r}
#| label: fig-1604
#| fig-cap: evaluation by district (last publication)
#| fig-width: 15
#| fig-height: 25

# real population by district and year
real_pop_dy <- real_pop |> 
  rename(year = StichtagDatJahr) |> 
  left_join(params$look_dis, by = "QuarCd") |> 
  mutate(district = factor(distr, levels = params$uni_d)) |> 
  group_by(year, district) |>   
  summarize(pop = sum(AnzBestWir), .groups = "drop") |> 
  mutate(scenario = params$text_c[1]) |> 
  select(district, year, scenario, pop)         
  

# past scenarios by district and year
scen_pop_dyc <- scen_pop |> 
  mutate(distnum = as.numeric(QuarCd)) |> 
  left_join(params$look_reg, by = "distnum") |> 
  filter(VersionArtCd %in% c(1,2,3)) |> 
  mutate(scenario = factor(
    case_when(
      VersionArtCd == 1 ~ params$text_c[2],
      VersionArtCd == 2 ~ params$text_c[3],
      TRUE       ~ params$text_c[4]
  ))) |> 
  rename(pub = PublJahr, year = Jahr) |>
  group_by(district, pub, year, scenario) |>
  summarize(pop = sum(AnzBestWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(district, year, scenario, cat, pop)

# plot: last publication
(scen_pop_dyc |> 
  filter(cat == max(cat)) |> 
  select(-cat) |> 
  bind_rows(real_pop_dy) |> 
  mutate(scenario = factor(scenario, levels = params$uni_c)) |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = scenario)) +
	  facet_wrap(~district, ncol = 3, scales = "free_y") +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) + 
    params$neutral + 
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      panel.spacing = unit(2.0, "lines"),
      axis.text.y = element_text(margin = margin(r = 10)),
      strip.background = element_blank())) |>  
  ggplotly(height = 2500)


```

Evaluation by district ('first' publication, 2016),
the first publication with the current model structure.

```{r}
#| label: fig-1605
#| fig-cap: evaluation by district ('first' publication, 2016)
#| fig-width: 15
#| fig-height: 25

# plot: first publication
(scen_pop_dyc |> 
  filter(cat == min(cat)) |>     
  select(-cat) |> 
  bind_rows(real_pop_dy) |> 
  mutate(scenario = factor(scenario, levels = params$uni_c)) |> 
  ggplot() +
	  geom_line(aes(x = year, y = pop, color = scenario)) +
	  facet_wrap(~district, ncol = 3, scales = "free_y") +
	  labs(x = "", y = "population", color = "") +
    scale_color_manual(values = params$col_c) +
    expand_limits(y = 0) +  
    params$neutral + 
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      panel.spacing = unit(2.0, "lines"),
      axis.text.y = element_text(margin = margin(r = 10)),
      strip.background = element_blank())) |>     
  ggplotly(height = 2500)


```

### Population by year and age

```{r}
#| label: fig-1606
#| fig-cap: evaluation by year and age category (middle scenario only)
#| fig-width: 20
#| fig-height: 20

# check: age category
# real_pop |> 
#   select(AlterV10Cd, AlterV10Kurz) |> 
#   unique()

# real population by year and age (ten year category)
real_pop_ya <- real_pop |> 
  rename(year = StichtagDatJahr) |> 
  mutate(age_3 = factor(if_else(AlterV10Cd >= 90, params$age_3t[10], AlterV10Kurz), levels = params$age_3t)) |> 
  group_by(year, age_3) |>
  summarize(pop = sum(AnzBestWir), .groups = "drop") |>
  mutate(cat = "reality") |>
  select(year, age_3, cat, pop)

# past scenarios by year (middle scenario only)
scen_pop_ya <- scen_pop |> 
  filter(VersionArtCd == 2) |> 
  rename(pub = PublJahr, year = Jahr) |> 
  left_join(params$look_a3, by = c("AlterVNum" = "age")) |> 
  group_by(pub, year, age_3) |> 
  summarize(pop = sum(AnzBestWir), .groups = "drop") |> 
  mutate(cat = paste("scenarios", pub)) |> 
  select(year, age_3, cat, pop)

# plot
(real_pop_ya |> 
  bind_rows(scen_pop_ya) |> 
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |> 
  ggplot() +
  geom_line(aes(x = year, y = pop, color = cat, linewidth = line_size)) +
  facet_wrap(~age_3, ncol = 2, scales = "free_y") +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "population", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>  
 ggplotly(height = 2000) 

```


## Births

### Births by year

```{r}
#| label: fig-1607
#| fig-cap: birth evaluation by year (middle scenario only)
#| fig-width: 10
#| fig-height: 10

# check data
# unique(real_bir$LebensfaehigkeitCd)

# real births by year
real_bir_y <- real_bir |> 
    rename(year = EreignisDatJahr) |> 
  group_by(year) |> 
  summarize(bir = sum(AnzGebuWir), .groups = "drop") |> 
  mutate(cat = "reality") |> 
  select(year, cat, bir)
  
   
# past scenarios by year (middle scenario only)
scen_bir_y <- scen_prc |> 
  filter(VersionArtCd == 2) |> 
  rename(pub = PublJahr, year = Jahr) |>
  group_by(pub, year) |>
  summarize(bir = sum(AnzGebuWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(year, cat, bir)

# plot
(real_bir_y |> 
  bind_rows(scen_bir_y) |> 
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |> 
  ggplot() +
  geom_line(aes(x = year, y = bir, color = cat, linewidth = line_size)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "births", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>  
 ggplotly() 


```

## Deaths

### Death by year

```{r}
#| label: fig-1608
#| fig-cap: death evaluation by year (middle scenario only)
#| fig-width: 10
#| fig-height: 10

# real deaths by year
real_dea_y <- real_dea |> 
    rename(year = EreignisDatJahr) |> 
  group_by(year) |> 
  summarize(dea = sum(AnzSterWir), .groups = "drop") |> 
  mutate(cat = "reality") |> 
  select(year, cat, dea)

   
# past scenarios by year (middle scenario only)
scen_dea_y <- scen_prc |> 
  filter(VersionArtCd == 2) |> 
  rename(pub = PublJahr, year = Jahr) |>
  group_by(pub, year) |>
  summarize(dea = sum(AnzSterWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(year, cat, dea)

# plot
(real_dea_y |>
  bind_rows(scen_dea_y) |>
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |>
  ggplot() +
  geom_line(aes(x = year, y = dea, color = cat, linewidth = line_size)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "deaths", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>
 ggplotly()

```

## Immigration

### Immigration by year


```{r}
#| label: fig-1609
#| fig-cap: immigration evaluation by year (middle scenario only)
#| fig-width: 10
#| fig-height: 10

# real immigration by year
real_imm_y <- real_imm |> 
    rename(year = EreignisDatJahr) |> 
  group_by(year) |> 
  summarize(imm = sum(AnzZuzuWir), .groups = "drop") |> 
  mutate(cat = "reality") |> 
  select(year, cat, imm)


# past scenarios by year (middle scenario only)
scen_imm_y <- scen_prc |>
  filter(VersionArtCd == 2) |>
  rename(pub = PublJahr, year = Jahr) |>
  group_by(pub, year) |>
  summarize(imm = sum(AnzZuzuWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(year, cat, imm)

# plot
(real_imm_y |>
  bind_rows(scen_imm_y) |>
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |>
  ggplot() +
  geom_line(aes(x = year, y = imm, color = cat, linewidth = line_size)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "immigration", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>
 ggplotly()


```


## Emigration

### Emigration by year

```{r}
#| label: fig-1610
#| fig-cap: emigration evaluation by year (middle scenario only)
#| fig-width: 10
#| fig-height: 10

# real emigration by year
real_emi_y <- real_emi |> 
    rename(year = EreignisDatJahr) |> 
  group_by(year) |> 
  summarize(emi = sum(AnzWezuWir), .groups = "drop") |> 
  mutate(cat = "reality") |> 
  select(year, cat, emi)


# past scenarios by year (middle scenario only)
scen_emi_y <- scen_prc |>
  filter(VersionArtCd == 2) |>
  rename(pub = PublJahr, year = Jahr) |>
  group_by(pub, year) |>
  summarize(emi = sum(AnzWezuWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(year, cat, emi)

# plot
(real_emi_y |>
  bind_rows(scen_emi_y) |>
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |>
  ggplot() +
  geom_line(aes(x = year, y = emi, color = cat, linewidth = line_size)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "emigration", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>
 ggplotly()


```

## Net migration

### net migration by year

```{r}
#| label: fig-1611
#| fig-cap: net migration evaluation by year (middle scenario only)
#| fig-width: 10
#| fig-height: 10


imm_temp <- real_imm_y |>
  bind_rows(scen_imm_y)

(real_emi_y |>
  bind_rows(scen_emi_y) |> 
  left_join(imm_temp, by = c("year", "cat")) |> 
  mutate(net = imm - emi, 
         line_size = ifelse(cat == "reality", "reality", "scenarios")) |>
  ggplot() +
  geom_hline(yintercept = 0, color = params$col_grey) +    
  geom_line(aes(x = year, y = net, color = cat, linewidth = line_size)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "net migration", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>
 ggplotly()


```

## Naturalization

### Naturalization by year


```{r}
#| label: fig-1612
#| fig-cap: naturalization evaluation by year (middle scenario only)
#| fig-width: 10
#| fig-height: 10

# real naturalization by year
real_nat_y <- real_nat |> 
  filter(HerkunftBisherCd == 2, HerkunftCd == 1) |> 
    rename(year = EreignisDatJahr) |> 
  group_by(year) |> 
  summarize(nat = sum(AnzEinbWir), .groups = "drop") |> 
  mutate(cat = "reality") |> 
  select(year, cat, nat)

# past scenarios by year (middle scenario only)
scen_nat_y <- scen_prc |>
  filter(VersionArtCd == 2) |>
  rename(pub = PublJahr, year = Jahr) |>
  group_by(pub, year) |>
  summarize(nat = sum(AnzEinbWir), .groups = "drop") |>
  mutate(cat = paste("scenarios", pub)) |>
  select(year, cat, nat)

# plot
(real_nat_y |>
  bind_rows(scen_nat_y) |>
  mutate(line_size = ifelse(cat == "reality", "reality", "scenarios")) |>
  ggplot() +
  geom_line(aes(x = year, y = nat, color = cat, linewidth = line_size)) +
  scale_color_manual(values = col_pal) +
  scale_linewidth_manual(values = c("reality" = 1.0, "scenarios" = 0.7)) +
  labs(x = "", y = "naturalization", color = "", linewidth = "") +
  expand_limits(y = 0) +
  guides(linewidth = "none") +
  params$neutral ) |>
 ggplotly()






```

